---
title: "Examples of the proper center and recalibrate pipeline: logistic regression"
author: "Juan Li"
date: 2026-06-15
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

# ======== For unknown issue of connecting to R server in positron, load all packages and source code here =========
library(rms)
library(dplyr)
library(stringr)
library(survival)
library(ggplot2)
library(ggpubr) 

# Prepare data for the centered algorithm
source("R/center_prepare.R", echo = FALSE) 
# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_pred.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
# Wrapper function for calibration curves and metrics
# Including the following functions: `calibration`, `calibration.con`, `calibration.OvsP`
source("R/calibration.R", echo = FALSE)
```

# Dataset: Heart Disease Dataset

We will use the [Heart Disease Dataset](https://www.kaggle.com/datasets/johnsmith88/heart-disease-dataset) dataset for the examples.

About this dataset, see this \[notebook\] (https://www.kaggle.com/code/tentotheminus9/what-causes-heart-disease-explaining-the-model)

1.  `age`: continuous. The person's age in years

2.  `sex`: categorical. The person's sex (1 = male, 0 = female)

3.  `cp`: categorical. The chest pain experienced (0: typical angina, 1: atypical angina, 2: non-anginal pain, 3: asymptomatic)

4.  `trestbps`: continuous. The person's resting blood pressure (in mm Hg on admission to the hospital)

5.  `chol`: continuous. The person's serum cholestoral in mg/dl

6.  `fbs`: categorical. The person's fasting blood sugar (\> 120 mg/dl, 1 = true; 0 = false)

7.  `restecg`: categorical. Resting electrocardiographic measurement (0 = normal, 1 = having ST-T wave abnormality, 2 = showing probable or definite left ventricular hypertrophy by Estes' criteria)

8.  `thalach`: continuous. The person's maximum heart rate achieved

9.  `exang`: categorical. Exercise induced angina (0 = no; 1 = yes)

10. `oldpeak`: continuous. ST depression induced by exercise relative to rest ('ST' relates to positions on the ECG plot. See more here)

11. `slope`: categorical. the slope of the peak exercise ST segment (0: upsloping, 1: flat, 2: downsloping)

12. `ca`: The number of major vessels (0-3) colored by flourosopy

13. `thal`: categorical. A blood disorder called thalassemia (1 = normal; 2 = fixed defect; 3 = reversable defect)

14. `target`: Heart disease (0 = no, 1 = yes)

```{r}
data <- read.csv("data/heart.csv", stringsAsFactors = TRUE)
dim(data) # there is no missing data in this dataset

data <- data %>% mutate(thal = ifelse(thal == 0, 1, thal)) # 7 cases that thal == 0, collapse levels to 1`
data_num <- data

# categorical variables
vars_cat <- c("sex", "cp", "fbs", "restecg", "exang", "slope", "thal")
data[vars_cat] <- lapply(data[vars_cat], factor) 

summary(data)
```

## Data sampling

To exaggerate the difference between original and target dataset, let's build a model on males (sex == 1) and test it on females (sex == 0).

```{r}
original <- data %>% filter(sex == 1) %>% select(-sex)
target <- data %>% filter(sex == 0) %>% select(-sex)
```

# Centering

## Original and centered models using the `original` dataset

For illustration purposes, I'll incorporate rcs, polynomials, and interaction involving rcs, which is probably not correct for this dataset.

**Note:** In the current version, the pipeline only consider raw and not orthogonal polynomials, thus speficify `raw = TRUE` in the `poly` function.

```{r}
# The original model without centering
options(scipen = 999)

fit.o <- glm(target ~ 
  cp + trestbps + chol + restecg + thalach + exang + ca + thal +
  fbs * rcs(age, 4) + slope * poly(oldpeak, 2, raw = TRUE), data = original, family = "binomial",
  # make sure the model is most light-weighted without the model frame, the model matrix, the response being returned
  model = FALSE, x = FALSE, y = FALSE) 

# A note that the original knot locations are recorded
attr(fit.o$terms, "predvars")
```

```{r}
# The centered model
centerList <- center_fit(fit.o, original)

# object of the centered model
fit.c <- centerList$fit.c

# check 
# 1: the coefficients of the two models are identical except intercept
cbind(coef(fit.o), coef(fit.c))
# 2: the predicted values of the two models are identical
cbind(head(fit.o$fitted.values), head(fit.c$fitted.values))
```

### model outputs

```{r}
print("Mean values of all terms")
centerList$vec_mean

print("Mean value of the continuous outcome")
centerList$outcome_mean

print("Knot locations")
centerList$rcsList

print("Coefficients")
centerList$fit.c$coefficients
```

## Transport the centered model to a new dataset

```{r}
pred.c.target <- center_predict(centerList, newdata = target)

# check: this produces the same predicted values as apply the original uncentered model on target
pred.o.target <- predict(fit.o, newdata = target)
cbind(head(pred.o.target), head(pred.c.target))
```

# Re-calibration

```{r}
recalibrateList <- center_recalibrate(centerList, newdata = target)
fit.c.recalibrated <- recalibrateList$centerList$fit.c

# check the updated coefficients
outcome_mean_original <-  mean(original$target, na.rm = T)
outcome_mean_target <- mean(target$target, na.rm = T)
outcome_mean_original_logit <- log(outcome_mean_original / (1 - outcome_mean_original))
outcome_mean_target_logit <- log(outcome_mean_target / (1 - outcome_mean_target))

print(paste("Outcome mean in original: ", outcome_mean_original))
print(paste("Outcome mean in target: ", outcome_mean_target))
print(paste("Logit of outcome mean in original: ", outcome_mean_original_logit))
print(paste("Logit of outcome mean in target: ", outcome_mean_target_logit))
print(paste("Difference in logit of outcome mean: ", outcome_mean_target_logit - outcome_mean_original_logit))
print(paste("Difference in model intercept: ", coef(fit.c.recalibrated)[1] - coef(fit.c)[1]))
cbind(coef(fit.c), coef(fit.c.recalibrated))

# check the difference between predicted linear predictors is equal to `delta_recal`
delta_recal <- recalibrateList$delta_recal
print(paste("Global offset by recalibration: ", delta_recal))
pred.c.recalibrated <- center_predict(recalibrateList$centerList, newdata = target)
cbind(head(pred.o.target), head(pred.c.target), head(pred.c.recalibrated), head(pred.c.recalibrated - pred.c.target))
```

```{r}
# null model
fit <- glm(target ~ 1, data = original, family = "binomial",
  # make sure the model is most light-weighted without the model frame, the model matrix, the response being returned
  model = FALSE, x = FALSE, y = FALSE) 
coef(fit)
```

# Calibration plots

```{r}
outcome <- target$target
```

```{r}
# the original model without centering
pred.o <- predict(fit.o, newdata = target, type = "response")
calibration(pred.o, outcome, package = "rms")
```

```{r}
# the original model with centering
pred.c <- center_predict(centerList, newdata = target, type = "response")
calibration(pred.c, outcome, package = "rms")
```

```{r}
# the recalibrated centered model, which only requires outcome mean in newdata
pred.c.recalibrated <- center_predict(recalibrateList$centerList, newdata = target, type = "response")
calibration(pred.c.recalibrated, outcome, package = "rms")
```

```{r}
# update intercept using offset, which requires linked individual-level newdata
df <- data.frame(
  outcome = outcome,
  logit = center_predict(centerList, newdata = target)
)

fit.c.offset <- glm(outcome ~ offset(logit), data = df, family = binomial,
  # make sure the model is most light-weighted without the model frame, the model matrix, the response being returned
  model = FALSE, x = FALSE, y = FALSE) 

pred.c.offset <- predict(fit.c.offset, newdata = df, type = "response")

calibration(pred.c.offset, outcome, package = "rms")
```

```{r}
# update intercept using Platt's calibration, which requires linked individual-level newdata
fit.c.platt <- glm(outcome ~ logit, data = df, family = binomial,
  # make sure the model is most light-weighted without the model frame, the model matrix, the response being returned
  model = FALSE, x = FALSE, y = FALSE) 

pred.c.platt <- predict(fit.c.platt, newdata = df, type = "response")

calibration(pred.c.platt, outcome, package = "rms")
```

## Subgroup

When using `CalibrationCurves::val.prob.ci.2`, it outputs the summary stats, but not plot.

```{r}
prediction <- pred.c.recalibrated
outcome <- target$target
```

### Adding a grouping variable: categorical

```{r}
#| warning: false
#| message: false

res <- calibration(prediction, outcome, group = target$fbs)
res$stats

calibration.OvsP(
  res$stats,
  group = target$fbs,
  group_con = FALSE,
  group_name = c("<= 120 mg/dl", "> 120 mg/dl"),
  xlab = "Fasting blood sugar"
)
```

### Adding a grouping variable: continuous

```{r}
#| warning: false
#| message: false

res <- calibration(prediction, outcome, group = target$age) # default to bin the group variable into 4 groups
res$stats

calibration.OvsP(
  res$stats,
  group = target$age,
  group_con = TRUE,
  xlab = "Age group"
)
```

```{r}
#| warning: false
#| message: false

res <- calibration(prediction, outcome, group = target$age, g.group=10) # parameter `g.group` control the number of groups
res$stats

calibration.OvsP(
  res$stats,
  group = target$age,
  group_con = TRUE,
  xlab = "Age group"
)
```

```{r}

```