---
title: "Test the new functions for survival models"
author: "Juan Li"
date: 2026-06-23
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(CalibrationCurves)
library(survival)
library(survminer)
library(ggsurvfit)
library(gtsummary)
library(riskRegression)

# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_predict.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
```

# Data

See `?CalibrationCurves::trainDataSurvival`.

```{r}
# variables for `time` and `status` are `ryear` and `rfs`, respectively
data(trainDataSurvival) # Rotterdam
data(testDataSurvival) # German Breast Cancer Study Group

original <- trainDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
target <- testDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
summary(original)
summary(target)
```

# Kaplan-Meier (KM) survival vs Baseline survival

```{r}
# -------- KM ------------
# Basic survival fit
fit_KM <- survfit(Surv(ryear, rfs) ~ 1, data = original)

# -------- baseline survival --------
fit.o.c <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # `riskRegression::predictRisk` and `CalibrationCurves::valProbSurvival` require `x=TRUE, y=TRUE`
  model = FALSE, x = TRUE, y = TRUE,
  # also center on dummy variables
  nocenter = NULL) 
baseline_mean <- survfit(fit.o.c)

compare <- list(KM = fit_KM, baseline = baseline_mean)
ggsurvplot_combine(
  compare,
  data = original,
  conf.int = TRUE,       # Adds 95% confidence intervals
  risk.table = TRUE,     # Displays the number of patients at risk over time
  xlab = "Time",         # X-axis label
  ylab = "Survival Probability", # Y-axis label
  ggtheme = theme_minimal() # Clean minimal theme
)
```

# Predict event probability (`prob`) at specific times

## Method 1: Original model **with centering** using `survival::coxph()`, with `nocenter = NULL`

```{r}
# Fit the coxph model
fit.o.c <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # `riskRegression::predictRisk` and `CalibrationCurves::valProbSurvival` require `x=TRUE, y=TRUE`
  model = FALSE, x = TRUE, y = TRUE,
  # also center on dummy variables
  nocenter = NULL) 
```

## Method 2: Still use functions in "center_recalibrate.R", so all the models follow the same pipeline

**Note:** Here, what `center_fit` does are

1.  Run `update(fit.o, nocenter = NULL)` to make sure all dummy variables are also centered.

2.  Store all model outputs to the returned `centerList`.

3.  Calculate outcome mean for recalibration, which is the average probability of having the event at a specific `time` based on the null model (based on the Kaplan-Meier or Aalen-Johansen estimator).

```{r}
fit.o <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # The modified functions relax this requirement, 
  # thus makes the model object light-weight and meet secured environment requirements
  model = FALSE, x = FALSE, y = FALSE) 

# Time points for calibration and recalibration
times <- c(1,3)

# The centered model
centerList <- center_fit(fit.o, original, times = times)

# object of the centered model
fit.c <- centerList$fit.c
```

## Transport the centered model to a new dataset

**Check:** all results below should be identical

```{r}
# -------- Methods 1 + `riskRegression::predictRisk` -------- 
# event probability at specific times
pred.prob.1 <- predictRisk(fit.o.c, newdata = target, times = c(1,3))
# surv = 1 - prob
pred.surv.1 <- 1 - pred.prob.1
head(cbind(pred.prob.1, pred.surv.1))

# -------- Methods 1 + `survfit` -------- 
# surv
surv.fit <- survfit(fit.o.c, newdata = target)
pred.surv.2 <- t(summary(surv.fit, times = c(1,3))$surv)
# prob = 1 - surv
pred.prob.2 <- 1 - pred.surv.2
head(cbind(pred.prob.2, pred.surv.2))

# -------- Methods 2 center_predict `center_predict` -------- 
# prob
pred.prob.3 <- center_predict(centerList, newdata = target, type = "probability", times = c(1,3))
# ssurv
pred.surv.3 <- center_predict(centerList, newdata = target, type = "survival", times = c(1,3))
head(cbind(pred.prob.3, pred.surv.3))
```

# Calibration plots

Compare the outputs of `CalibrationCurves::valProbSurvival` and `valProbSurvival.2`, and the brier score generated by `riskRegression::Score` in `CalibrationCurves::valProbSurvival` , `survival::brier`, and the modified `brier.cox`

```{r}
source("R/calibration.surv.R", echo = FALSE) # Modified function for cox models

timeHorizon <- 3
timeHorizon_brier <- timeHorizon - 0.01 # as implemented within `CalibrationCurves::valProbSurvival`
```

```{r}
# `CalibrationCurves::valProbSurvival`
res.1 <- valProbSurvival(fit.o.c, target, timeHorizon = timeHorizon, plotCal = "ggplot", nk = 5)
res.1$stats$Calibration$InTheLarge
res.1$stats$Calibration$Statistics
res.1$stats$Calibration$BrierScore
res.1$stats$Concordance
res.1$stats$TimeDependentAUC
```

```{r}
# `valProbSurvival.2`, through the wrapper function `calibration.surv`
res.2 <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
res.2$stats$Calibration$InTheLarge
res.2$stats$Calibration$Statistics
res.2$stats$Calibration$BrierScore
res.2$stats$Concordance
res.2$stats$TimeDependentAUC
```

Further, if `fit.o.c` were fitted with `x = FALSE, y = FALSE`, we can still get its calibration curve using `valProbSurvival.2`. (Not `calibration.surv` since there is no `centerList`).

```{r}
fit.o.c.2 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # `riskRegression::predictRisk` and `CalibrationCurves::valProbSurvival` require `x=TRUE, y=TRUE`
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL) 

# get the prediction and outcomes in `target`
pred.lp <- predict(fit.o.c.2, newdata = target, type = "lp")
surv.fit <- survfit(fit.o.c.2, newdata = target)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon_brier)$surv)
outcome.time <- target$ryear
outcome.event <- target$rfs

res.3 <- valProbSurvival.2(
    pred.lp,
    pred.prob,
    pred.prob.brier,
    outcome.time,
    outcome.event,
    # IMPORTANT, make sure this `timeHorizon` matches the one used in `pred.prob` and `pred.prob.brier`
    timeHorizon = timeHorizon, 
    plotCal = "ggplot", 
    nk = 5
  )

res.3$stats$Calibration$InTheLarge
res.3$stats$Calibration$Statistics
res.3$stats$Calibration$BrierScore
res.3$stats$Concordance
res.3$stats$TimeDependentAUC
```

```{r}

```