---
title: "Examples of the proper center and recalibrate pipeline: coxph"
author: "Juan Li"
date: 2026-06-16
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(CalibrationCurves)
library(survival)
library(ggsurvfit)
library(gtsummary)

# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_predict.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
```

# Breast Cancer Survival Data from Rotterdam and Germany

See `?CalibrationCurves::trainDataSurvival`.

```{r}
# variables for `time` and `status` are `ryear` and `rfs`, respectively
data(trainDataSurvival) # Rotterdam
data(testDataSurvival) # German Breast Cancer Study Group
```

Kaplan-Meier plots: https://www.emilyzabor.com/survival-analysis-in-r.html

```{r}
data <- bind_rows(
  trainDataSurvival %>% select(ryear, rfs) %>% mutate(type = "train"),
  testDataSurvival %>% select(ryear, rfs) %>% mutate(type = "test")
)
data$type <- factor(data$type, levels = c("train", "test"))

survfit2(Surv(ryear, rfs) ~ type, data = data) %>% 
  ggsurvfit() +
  add_confidence_interval() +
  add_risktable()
```

## Variables

### in both datasets

`pid`: patient identifier

`age`: age at surgery

`meno`: menopausal status (0 = premenopausal, 1 = postmenopausal)

`size`: tumor size, a factor with levels \<= 20, 20-50, \>50

- `csize`: Categorized tumor size, copied from size (factor)

`grade`: differentiation grade

- `grade3`: Recoded grade factor (levels: "1-2", "3")

`nodes`: number of positive lymph nodes

- `nodes2`: Winsorized node count (numeric)

- `nodes3`: Restricted cubic spline basis for nodes2 (numeric)

- `cnode`: Categorized node involvement (factor: "0", "1-3", "\>3")

`pgr`: progesterone receptors (fmol/l)

- `pgr2`: Winsorized progesterone receptor level (numeric)

- `pgr3`: Restricted cubic spline basis for original pgr (numeric)

`er`: estrogen receptors (fmol/l)

`hormon`: hormonal treatment (0 = no, 1 = yes)

`ryear`: Follow-up time for RFS, in years (numeric)

`rfs`: Recurrence-free survival status (0 = no event, 1 = event) (numeric)

`epoch`: Follow-up epoch indicator after splitting at 5 years (numeric)

### only in `trainDataSurvival`

`year`: year of surgery

`chemo`: chemotherapy

`rtime`: days to relapse or last follow-up

`recur`: 0 = no relapse, 1 = relapse

`dtime`: days to death or last follow-up

`death`: 0 = alive, 1 = dead

`tstart`

### only in `testDataSurvival`

`rfstime`, `status`

```{r}
original <- trainDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
target <- testDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
summary(original)
summary(target)
```

# Centering

## Understand the internal centering in `survival::coxph`

(From Claude) In R’s `survival::coxph` function, predictors are internally mean-centered for numerical stability.

1.  Centering on the continuous variable cannot be disabled — it's hardcoded for numerical stability in the optimizer.

2.  The `coxph` function has a parameter `nocenter`, which controls the centering on dummy, categorical variables. Its default value `c(-1, 0, 1)` means columns that take only values from {-1, 0, 1} (i.e., dummy/binary variables from factors) are not centered, which explains the difference observed below.

3.  Therefore, for survival models, instead of manually doing centering, we should simply set `nocenter = NULL` when fitting the original model and let the function do it itself.

4.  Also, for model output, all the mean values can be found from `fit$means`.

5.  For `rcs` terms, the original knot locations are also stored in `fit$terms`, and used when predicting on a `newdata`.

```{r}
# -------- Continuous variable -------- 
df <- original %>% select(ryear, rfs, pgr2)
mean_pgr2 <- mean(df$pgr2)
df$pgr2_C <- df$pgr2 - mean_pgr2
fit.o <- coxph(Surv(ryear, rfs) ~ pgr2, data = df)
pred.o <- predict(fit.o, newdata = df, type = "lp")
fit.c <- coxph(Surv(ryear, rfs) ~ pgr2_C, data = df)
pred.c <- predict(fit.c, newdata = df, type = "lp")
head(cbind(pred.o, pred.c))
fit.o$means
fit.c$means
```

```{r}
# -------- Continuous variable, rcs -------- 
df <- original %>% select(ryear, rfs, pgr2)
k <- 3
fit.o <- coxph(Surv(ryear, rfs) ~ rcs(pgr2, k), data = df)
pred.o <- predict(fit.o, newdata = df, type = "lp")
# get and center rcs terms
rcs.fit <- rcs(df$pgr2, k)
res <- labelled::remove_attributes(
  rcs.fit,
  names(attributes(rcs.fit))[3:length(names(attributes(rcs.fit)))]
) # don't print the attributes
colnames(res) <- c("pgr2_rcs_1", "pgr2_rcs_2")
df <- bind_cols(df, res)
means <- colMeans(df %>% select(contains("rcs")))
df$`pgr2_rcs_1_C` <- df$`pgr2_rcs_1` - means[1]
df$`pgr2_rcs_2_C` <- df$`pgr2_rcs_2` - means[2]
fit.c <- coxph(Surv(ryear, rfs) ~ pgr2_rcs_1_C + pgr2_rcs_2_C, data = df)
pred.c <- predict(fit.c, newdata = df, type = "lp")
head(cbind(pred.o, pred.c))
fit.o$means
fit.c$means

# A note that the original knot locations are recorded
attr(fit.o$terms, "predvars")
```

```{r}
# -------- Categorical variable -------- 
df <- original %>% select(ryear, rfs, csize)
df <- step_dummy(df, c("csize"))
df$`csize20-50_C` <- df$`csize20-50` - mean(df$`csize20-50`)
df$`csize>50_C` <- df$`csize>50` - mean(df$`csize>50`)

fit.o.1 <- coxph(Surv(ryear, rfs) ~ csize, data = df) # do not center on dummy variables
pred.o.1 <- predict(fit.o.1, newdata = df, type = "lp")
fit.o.2 <- coxph(Surv(ryear, rfs) ~ csize, data = df, nocenter = NULL) # also center on dummy variables
pred.o.2 <- predict(fit.o.2, newdata = df, type = "lp")
fit.c <- coxph(Surv(ryear, rfs) ~ `csize20-50_C` + `csize>50_C`, data = df)
pred.c <- predict(fit.c, newdata = df, type = "lp")
head(cbind(pred.o.1, pred.o.2, pred.c))
fit.o.1$means
fit.o.2$means
fit.c$means
```

```{r}
# -------- Interaction -------- 
df <- original %>% select(ryear, rfs, csize, pgr2)
fit.o <- coxph(Surv(ryear, rfs) ~ csize * pgr2, data = df)
fit.o$means
fit.o <- coxph(Surv(ryear, rfs) ~ csize * pgr2, data = df, nocenter = NULL) # also center on dummy variables
fit.o$means
```

## Fitting the centered model

### Method 1: Original model **with centering** using `survival::coxph()`, with `nocenter = NULL`

**Notes:**

1.  For illustration purposes, I'll incorporate rcs, polynomials, and interaction involving rcs, which is probably not correct for this dataset.

2.  It is default in `coxph` that `model=FALSE, x=FALSE, y=TRUE`. For the purpose of making the model object light-weighted, and to work with secured environment, it is preferred or even required to have `model=FALSE, x=FALSE, y=TRUE`. However, `CalibrationCurves::valProbSurvival` and `riskRegression::plotCalibration` requires `x=TRUE, y=TRUE`.

3.  I (and Claude) have modified the `CalibrationCurves::valProbSurvival` function as `valProbSurvival.2` which

- Doesn't require `x=TRUE, y=TRUE`, thus meets the requirement of the secured environment.

- Use the appropriate prediction and outcome variables instead of the fitted model object, which can then work well with the recalibrated model.

- Also works with Fine-Gray model (with some requirement in the set-up).

- See "3.1_Test_new_functions.qmd", which I compared both functions, and make sure the modified version works as expected.

```{r}
# The original model without centering
options(scipen = 999)

fit.o.c <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL) 
fit.o.c %>% tbl_regression(exp = TRUE) 

# Manually compute the shift
print("Term means: ")
(means <- fit.o.c$means) # verify all terms have been centered
print("Model coefficients: ")
(coef <- fit.o.c$coefficients)
shift <- coef %*% means

# baseline hazard with centering, i.e., the reference is a subject with all average covariates.
fit.o.c.h0 <- basehaz(fit.o.c) 
# baseline hazard without centering, i.e., the reference is a subject with all 0 covariates.
fit.o.c.h0.nc <- basehaz(fit.o.c, centered = FALSE) 

# Check: exp(shift) should equal to fit.o.c.h0$hazard/fit.o.c.h0.nc$hazard
print(paste("Exp of the shift: ", exp(shift)))
cbind(head(fit.o.c.h0$hazard), head(fit.o.c.h0.nc$hazard), head(fit.o.c.h0$hazard / fit.o.c.h0.nc$hazard))
```

### Method 2: Still use functions in "center_recalibrate.R", so all the models follow the same pipeline

**Note:** Here, what `center_fit` does are

1.  Run `update(fit.o, nocenter = NULL)` to make sure all dummy variables are also centered.

2.  Store all model outputs to the returned `centerList`.

3.  Unlike in the linear/logistic regression models, "outcome mean" for survival models (cox, fine-gray) is the average probability of having the event at a specific time (`times`) based on the null model (based on the Kaplan-Meier or Aalen-Johansen estimator). **Note:** I made the decision here to pre-specify `times` instead of outputting the whole `survfit` result, that to make the model output light-weighted.

```{r}
# The original model
fit.o <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE) 

# Time points for calibration and recalibration
times <- c(1, 2, 3, 4)

# The centered model
centerList <- center_fit(fit.o, original, times = times)

# object of the centered model
fit.c <- centerList$fit.c
```

## model outputs: Verify that the two methods produce the same model output

```{r}
# check: coefficients of the three models are identical
cbind(coef(fit.o.c), coef(fit.o), coef(fit.c))
```

```{r}
print("Method 1:")
print("Mean values of all terms")
fit.o.c$means

print("knot locations")
attr(fit.o.c$terms, "predvars")
```

```{r}
print("Method 2:")
print("Mean values of all terms")
centerList$vec_mean

print("knot locations")
centerList$rcsList

print("outcome mean at specific time points:")
centerList$outcome_mean
```

```{r}
fit.o.c.h0 <- basehaz(fit.o.c) 
fit.c.ho <- basehaz(fit.c) 
print("baseline hazard")
head(cbind(fit.o.c.h0$hazard, fit.c.ho$hazard))
```

**Note:** From the identical term mean values, it is verified that the `coxph` function also follows *transfer then center*.

```{r}
cbind(fit.o.c$means, centerList$vec_mean, fit.o.c$means - centerList$vec_mean)
```

## Transport the centered model to a new dataset

**Note:** Here, `center_predict` is a wrapper function of the corresponding code in Method 1, the main reason to have `center_predict` is to automatically incorporate `delta_recal`.

```{r}
# -------- Methods 1 -------- 
pred.m1.lp <- predict(fit.o.c, newdata = target, type = "lp")
pred.m1.risk <- predict(fit.o.c, newdata = target, type = "risk")
surv.fit <- survfit(fit.o.c, newdata = target)
pred.m1.survival <- t(summary(surv.fit, times = times)$surv)
pred.m1.prob <- 1 - pred.m1.survival
head(cbind(pred.m1.lp, pred.m1.risk, pred.m1.survival, pred.m1.prob))

# -------- Methods 2 -------- 
pred.m2.lp <- center_predict(centerList, newdata = target, type = "lp")
pred.m2.risk <- center_predict(centerList, newdata = target, type = "risk")
pred.m2.survival <- center_predict(centerList, newdata = target, type = "survival", times = times)
pred.m2.prob <- center_predict(centerList, newdata = target, type = "probability", times = times)
head(cbind(pred.m2.lp, pred.m2.risk, pred.m2.survival, pred.m2.prob))
```

**Note:**

1.  The `coxph` object cannot be directly updated to reflect the recalibration changes. The current work-around is to do it manually in `predict_centre`.

2.  Therefore, here `center_recalibrate` is mainly used to calculate the model shift, *i.e.* `delta_recal`.

3.  For re-calibration, we can only work with one single `timeHorizon` at a time.

4.  The specific `timeHorizon` should be within the pre-specify `times` when running `center_fit`.

5.  The method in doi:10.1016/j.jclinepi.2025.111895 is as following:

- suppose the risk/probability of event at a specific time is $\theta = 1 - S_{KM}(t)$, the rescaling factor they proposed was $\Delta_{rescaling} = ln[-ln(1-\theta_{derived})] - ln[-ln(1-\theta_{target})]$.

- Then, the recalibrated risk is $\theta_{recal} = 1 - exp(-exp(ln[-ln(1-\theta_{ori})]-\Delta_{rescaling}))$.

- Let's substitute $S_{recal} = 1 - \theta_{recal}$ and $S_{ori} = 1 - \theta_{ori}$ into the equation, then $1 - S_{recal} = 1 - exp(-exp(ln[-ln(S_{ori})]-\Delta_{rescaling}))$, *i.e.*, $S_{recal} = exp(-exp(ln[-ln(S_{ori})]-\Delta_{rescaling}))$

$= e^{-e^{ln[-ln(S_{ori})]-\Delta_{rescaling}}}$

$= e^{-e^{ln[-ln(S_{ori})]}e^{-\Delta_{rescaling}}}$

$= [e^{-e^{ln[-ln(S_{ori})]}}]^{e^{-\Delta_{rescaling}}}$

$= [e^{ln(S_{ori})}]^{e^{-\Delta_{rescaling}}}$

$= [S_{ori}]^{e^{-\Delta_{rescaling}}}$

- In the Cox model, the equation for hazard is $h(t|X_i) = h_0(t)exp(\beta^TX_i)$. And the survival function is $S(t|X_i) = [S_0(t)]^{exp(\beta^TX_i)}$. Therefore, the equation above can be written as $S_{recal} = [S_{ori}]^{e^{-\Delta_{rescaling}}}$

$= ([S_{0,ori}(t)]^{e^{\beta^TX_i}})^{e^{-\Delta_{rescaling}}}$

$= S_{0,ori}(t)^{e^{\beta^TX_i}e^{-\Delta_{rescaling}}}$

$= S_{0,ori}(t)^{e^{(\beta^TX_i -\Delta_{rescaling})}}$

- Note, in my implementation, I make `delta_recal` = $-\Delta_{rescaling} = ln[-ln(1-\theta_{target})] - ln[-ln(1-\theta_{derived})]$ (*i.e.* `outcome_mean_target_clog_log - outcome_mean_original_clog_log`).

- In summary, with `delta_recal`, the predictions will be updated as follows (see `center_predict`):

```         
- `type = 'lp'`: `pred_lp_recal = pred_lp + delta_recal`

- `type = 'risk'` (_i.e._ `exp(pred_lp)`): `pred_risk_recal = pred_risk * exp(delta_recal)`

- `type = 'surv'` (using `survfit`+`summary`): `pred_surv_recal = pred_surv ^ exp(delta_recal)`

- `type = 'probability'` (_i.e._ `1 - pred_surv`): `pred_prob_recal = 1 - pred_surv_recal = 1 - pred_surv ^ exp(delta_recal)`
```

```{r}
timeHorizon <- 3 # specify a time horizon for calculating 'outcome mean'
recalibrateList <- center_recalibrate(centerList, newdata = target, timeHorizon = timeHorizon)
fit.c.recalibrated <- recalibrateList$centerList$fit.c # this is still fit.c

# check the results
surv_fit_original <- survfit(Surv(ryear, rfs) ~ 1, data = original)
outcome_mean_original <- 1 - summary(surv_fit_original, times = timeHorizon)$surv
outcome_mean_original_clog_log <- log(-log(1-outcome_mean_original)) 
      
surv_fit_target <- survfit(Surv(ryear, rfs) ~ 1, data = target)
outcome_mean_target <- 1 - summary(surv_fit_target, times = timeHorizon)$surv
outcome_mean_target_logit <- log(outcome_mean_target / (1 - outcome_mean_target))
outcome_mean_target_clog_log <- log(-log(1-outcome_mean_target))

print(paste0("Outcome mean in original: ", outcome_mean_original))
print(paste0("Outcome mean in target: ", outcome_mean_target))
print(paste0("Clog-log of outcome mean in original: ", outcome_mean_original_clog_log))
print(paste0("Clog-log of outcome mean in target: ", outcome_mean_target_clog_log))
print(paste0("Difference in outcome mean: ", outcome_mean_target - outcome_mean_original))
print(paste0("Difference in clog-log of outcome mean: ", outcome_mean_target_clog_log - outcome_mean_original_clog_log))

# check the difference between predicted lp 
print("The global offset by recalibration:")
(delta_recal <- recalibrateList$delta_recal)
pred.c <- center_predict(centerList, newdata = target, type = "lp")
pred.c.recalibrated <- center_predict(recalibrateList$centerList, newdata = target, type = "lp", delta_recal = delta_recal)
cbind(head(pred.c), head(pred.c.recalibrated), head(pred.c.recalibrated - pred.c))

# check the difference between predicted risk
print(paste("Exp of the global offset by recalibration: ", exp(delta_recal)))
pred.c <- center_predict(centerList, newdata = target, type = "risk")
pred.c.recalibrated <- center_predict(recalibrateList$centerList, newdata = target, type = "risk", delta_recal = delta_recal)
cbind(head(pred.c), head(pred.c.recalibrated), head(pred.c.recalibrated/pred.c))
```

# Calibration plots

Function `CalibrationCurves::valProbSurvival` is modified as `valProbSurvival.2` for the following reasons:

1.  `CalibrationCurves::valProbSurvival` requires `x = TRUE, y = TRUE`, thus does not work on the light-weight model object `fit.o.c`.

2.  Since `coxph` object cannot be directly updated to reflect the change made by recalibration, we need to use specific prediction vectors instead of the fitted model object.

3.  Some further modification for subgroups.

**Note:**

1.  See "3.1_Test_new_functions.qmd" that varified the two function produce identical results (the only thing lacking in `valProbSurvival.2` is the se and confidence interval for brier scores).

2.  For centered models with `centerList`, `valProbSurvival.2` is called through a wrapper function `calibration.surv`.

```{r}
source("R/calibration.surv.R", echo = FALSE) # Modified function for cox models
```

```{r}
# The original model using Method 1
pred.lp <- predict(fit.o.c, target, type = "lp")
surv.fit <- survfit(fit.o.c, newdata = target)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- target$ryear
outcome.event <- target$rfs

res.1 <- valProbSurvival.2(
  pred.lp,
  pred.prob,
  pred.prob.brier,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
# The original model using Method 2
res.2 <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
# The recalibrated model, indicated by `delta_recal`: the linear one
res.3 <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  delta_recal = delta_recal,
  plotCal = "ggplot",
  nk = 5
)
```

**Note:** the `offset(lp)` method doesn't work here

```{r}
# update intercept using Platt's calibration, which requires linked individual-level newdata
# Use `valProbSurvival.2` directly here

target$LP <- center_predict(centerList, newdata = target, type = "lp")
fit.offset <- coxph(Surv(ryear, rfs) ~ LP, data = target,
model = FALSE, x = FALSE, y = FALSE)

pred.lp <- predict(fit.offset, target, type = "lp")
surv.fit <- survfit(fit.offset, newdata = target)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- target$ryear
outcome.event <- target$rfs

res.4 <- valProbSurvival.2(
  pred.lp,
  pred.prob,
  pred.prob.brier,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
res.1$stats$Calibration$Statistics
res.2$stats$Calibration$Statistics
res.3$stats$Calibration$Statistics
res.4$stats$Calibration$Statistics
```

```{r}
for (timeHorizon in times) {
  recalibrateList <- center_recalibrate(
    centerList,
    newdata = target,
    timeHorizon = timeHorizon
  )
  delta_recal <- recalibrateList$delta_recal

  # original model
  res.2 <- calibration.surv(
    centerList,
    newdata = target,
    timeHorizon = timeHorizon,
    plotCal = "ggplot",
    nk = 5
  )

  # recalibrated model
  res.3 <- calibration.surv(
    centerList,
    newdata = target,
    timeHorizon = timeHorizon,
    delta_recal = delta_recal,
    plotCal = "ggplot",
    nk = 5
  )
  print(paste0("Recalibrate at year: ", timeHorizon))
  print(paste0("Shift at that year: ", delta_recal))
  print(res.2$stats$Calibration$Statistics)
  print(res.3$stats$Calibration$Statistics)
}
```

## Subgroup

```{r}
source("R/calibration.OvsP.R", echo = FALSE) # Bar plot of observed vs predicted in each subgroup
```

### Adding a grouping variable: categorical

```{r}
# The original model using Method 2
res.OvsP <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  group = target$grade3,
  plotCal = "ggplot",
  nk = 5
)
res.OvsP

calibration.OvsP(
  res.OvsP,
  group = target$grade3,
  group_con = FALSE,
  xlab = "grade3"
)
```

### Adding a grouping variable: continuous

```{r}
# The original model using Method 2
res.OvsP <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  group = target$pgr2,
  g.group = 4,
  plotCal = "ggplot",
  nk = 5
)
res.OvsP

calibration.OvsP(
  res.OvsP,
  group = target$pgr2,
  group_con = TRUE,
  xlab = "pgr2"
)

# If treating it as a categorical variable
calibration.OvsP(
  res.OvsP,
  group = target$pgr2,
  group_con = FALSE,
  xlab = "pgr2"
)
```

```{r}

```