---
title: "Examples of the proper center and recalibrate pipeline: coxph"
author: "Juan Li"
date: 2026-06-16
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

# ======== For unknown issue of connecting to R server in positron, load all packages and source code here =========
library(rms)
library(Hmisc)
library(dplyr)
library(stringr)
library(survival)
library(ggplot2)
library(ggpubr) 
library(CalibrationCurves)
library(riskRegression)
library(ggsurvfit)
library(gtsummary)
library(timeROC)

# Prepare data for the centered algorithm
source("R/center_prepare.R", echo = FALSE) 
# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_pred.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
# Wrapper function for calibration curves and metrics
# Including the following functions: `calibration`, `calibration.con`, `calibration.OvsP`
source("R/calibration.R", echo = FALSE)
# Modified function for cox models
source("R/calibration.surv.R", echo = FALSE) 
source("R/calibration_curve_cox.R", echo = FALSE) # valProbSurvival.2
```

# Breast Cancer Survival Data from Rotterdam and Germany

See `?CalibrationCurves::trainDataSurvival`.

```{r}
# variables for `time` and `status` are `ryear` and `rfs`, respectively
data(trainDataSurvival) # Rotterdam
data(testDataSurvival) # German Breast Cancer Study Group
```

## Variables

### in both datasets

`pid`: patient identifier

`age`: age at surgery

`meno`: menopausal status (0 = premenopausal, 1 = postmenopausal)

`size`: tumor size, a factor with levels \<= 20, 20-50, \>50

- `csize`: Categorized tumor size, copied from size (factor)

`grade`: differentiation grade

- `grade3`: Recoded grade factor (levels: "1-2", "3")

`nodes`: number of positive lymph nodes

- `nodes2`: Winsorized node count (numeric)

- `nodes3`: Restricted cubic spline basis for nodes2 (numeric)

- `cnode`: Categorized node involvement (factor: "0", "1-3", "\>3")

`pgr`: progesterone receptors (fmol/l)

- `pgr2`: Winsorized progesterone receptor level (numeric)

- `pgr3`: Restricted cubic spline basis for original pgr (numeric)

`er`: estrogen receptors (fmol/l)

`hormon`: hormonal treatment (0 = no, 1 = yes)

`ryear`: Follow-up time for RFS, in years (numeric)

`rfs`: Recurrence-free survival status (0 = no event, 1 = event) (numeric)

`epoch`: Follow-up epoch indicator after splitting at 5 years (numeric)

### only in `trainDataSurvival`

`year`: year of surgery

`chemo`: chemotherapy

`rtime`: days to relapse or last follow-up

`recur`: 0 = no relapse, 1 = relapse

`dtime`: days to death or last follow-up

`death`: 0 = alive, 1 = dead

`tstart`

### only in `testDataSurvival`

`rfstime`, `status`

```{r}
original <- trainDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
target <- testDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
summary(original)
summary(target)
```

# Centering

## Understand the internal centering in `survival::coxph`

(From Claude) In R’s `survival::coxph` function, predictors are internally mean-centered for numerical stability.

1.  Centering on the continuous variable cannot be disabled — it's hardcoded for numerical stability in the optimizer.

2.  The `coxph` function has a parameter `nocenter`, which controls the centering on dummy, categorical variables. Its default value `c(-1, 0, 1)` means columns that take only values from {-1, 0, 1} (i.e., dummy/binary variables from factors) are not centered, which explains the difference observed below.

3.  Therefore, for survival models, instead of manually doing centering, we should simply set `nocenter = NULL` when fitting the original model and let the function do it itself.

4.  Also, for model output, all the mean values can be found from `fit$means`.

5.  For `rcs` terms, the original knot locations are also stored in `fit$terms`, and used when predicting on a `newdata`.

```{r}
# -------- Continuous variable -------- 
df <- original %>% select(ryear, rfs, pgr2)
mean_pgr2 <- mean(df$pgr2)
df$pgr2_C <- df$pgr2 - mean_pgr2
fit.o <- coxph(Surv(ryear, rfs) ~ pgr2, data = df)
pred.o <- predict(fit.o, newdata = df, type = "lp")
fit.c <- coxph(Surv(ryear, rfs) ~ pgr2_C, data = df)
pred.c <- predict(fit.c, newdata = df, type = "lp")
head(cbind(pred.o, pred.c))
fit.o$means
fit.c$means
```

```{r}
# -------- Continuous variable, rcs -------- 
df <- original %>% select(ryear, rfs, pgr2)
k <- 3
fit.o <- coxph(Surv(ryear, rfs) ~ rcs(pgr2, k), data = df)
pred.o <- predict(fit.o, newdata = df, type = "lp")
# get and center rcs terms
rcs.fit <- rcs(df$pgr2, k)
res <- labelled::remove_attributes(
  rcs.fit,
  names(attributes(rcs.fit))[3:length(names(attributes(rcs.fit)))]
) # don't print the attributes
colnames(res) <- c("pgr2_rcs_1", "pgr2_rcs_2")
df <- bind_cols(df, res)
means <- colMeans(df %>% select(contains("rcs")))
df$`pgr2_rcs_1_C` <- df$`pgr2_rcs_1` - means[1]
df$`pgr2_rcs_2_C` <- df$`pgr2_rcs_2` - means[2]
fit.c <- coxph(Surv(ryear, rfs) ~ pgr2_rcs_1_C + pgr2_rcs_2_C, data = df)
pred.c <- predict(fit.c, newdata = df, type = "lp")
head(cbind(pred.o, pred.c))
fit.o$means
fit.c$means

# A note that the original knot locations are recorded
attr(fit.o$terms, "predvars")
```

```{r}
# -------- Categorical variable -------- 
df <- original %>% select(ryear, rfs, csize)
df <- step_dummy(df, c("csize"))
df$`csize20-50_C` <- df$`csize20-50` - mean(df$`csize20-50`)
df$`csize>50_C` <- df$`csize>50` - mean(df$`csize>50`)

fit.o.1 <- coxph(Surv(ryear, rfs) ~ csize, data = df) # do not center on dummy variables
pred.o.1 <- predict(fit.o.1, newdata = df, type = "lp")
fit.o.2 <- coxph(Surv(ryear, rfs) ~ csize, data = df, nocenter = NULL) # also center on dummy variables
pred.o.2 <- predict(fit.o.2, newdata = df, type = "lp")
fit.c <- coxph(Surv(ryear, rfs) ~ `csize20-50_C` + `csize>50_C`, data = df)
pred.c <- predict(fit.c, newdata = df, type = "lp")
head(cbind(pred.o.1, pred.o.2, pred.c))
fit.o.1$means
fit.o.2$means
fit.c$means
```

```{r}
# -------- Interaction -------- 
df <- original %>% select(ryear, rfs, csize, pgr2)
fit.o <- coxph(Surv(ryear, rfs) ~ csize * pgr2, data = df)
fit.o$means
fit.o <- coxph(Surv(ryear, rfs) ~ csize * pgr2, data = df, nocenter = NULL) # also center on dummy variables
fit.o$means
```

## Fitting the centered model

### Method 1: Original model **with centering** using `survival::coxph()`, with `nocenter = NULL`

**Notes:**

1.  For illustration purposes, I'll incorporate rcs, polynomials, and interaction involving rcs, which is probably not correct for this dataset.

2.  It is default in `coxph` that `model=FALSE, x=FALSE, y=TRUE`. For the purpose of making the model object light-weighted, and to work with secured environment, it is preferred or even required to have `model=FALSE, x=FALSE, y=TRUE`. However, `CalibrationCurves::valProbSurvival` and `riskRegression::plotCalibration` requires `x=TRUE, y=TRUE`.

3.  I (and Claude) have modified the `CalibrationCurves::valProbSurvival` function as `valProbSurvival.2` which

- Doesn't require `x=TRUE, y=TRUE`, thus meets the requirement of the secured environment.

- Use the appropriate prediction and outcome variables instead of the fitted model object, which can then work well with the recalibrated model.

- Also works with Fine-Gray model (with some requirement in the set-up).

- See "3.1_Test_new_functions.qmd", which I compared both functions, and make sure the modified version works as expected.

```{r}
# The original model without centering
options(scipen = 999)

fit.o.c <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL) 
fit.o.c %>% tbl_regression(exp = TRUE) 

# Manually compute the shift
print("Term means: ")
(means <- fit.o.c$means) # verify all terms have been centered
print("Model coefficients: ")
(coef <- fit.o.c$coefficients)
shift <- coef %*% means

# baseline hazard with centering, i.e., the reference is a subject with all average covariates.
fit.o.c.h0 <- basehaz(fit.o.c) 
# baseline hazard without centering, i.e., the reference is a subject with all 0 covariates.
fit.o.c.h0.nc <- basehaz(fit.o.c, centered = FALSE) 

# Check: exp(shift) should equal to fit.o.c.h0$hazard/fit.o.c.h0.nc$hazard
print(paste("Exp of the shift: ", exp(shift)))
cbind(head(fit.o.c.h0$hazard), head(fit.o.c.h0.nc$hazard), head(fit.o.c.h0$hazard / fit.o.c.h0.nc$hazard))
```

### Method 2: Still use functions in "center_recalibrate.R", so all the models follow the same pipeline

**Note:** Here, what `center_fit` does are

1.  Run `update(fit.o, nocenter = NULL)` to make sure all dummy variables are also centered.

2.  Store all model outputs to the returned `centerList`.

3.  Unlike in the linear/logistic regression models, "outcome mean" for survival models (cox, fine-gray) is the average probability of having the event at a specific time (`times`) based on the null model (based on the Kaplan-Meier or Aalen-Johansen estimator). **Note:** I made the decision here to pre-specify `times` instead of outputting the whole `survfit` result, that to make the model output light-weighted.

```{r}
# The original model
fit.o <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE) 

# Time points for calibration and recalibration
times <- c(1, 2, 3, 4)

# The centered model
centerList <- center_fit(fit.o, original, times = times)

# object of the centered model
fit.c <- centerList$fit.c
```

## model outputs: Verify that the two methods produce the same model output

```{r}
# check: coefficients of the three models are identical
cbind(coef(fit.o.c), coef(fit.o), coef(fit.c))
```

```{r}
print("Method 1:")
print("Mean values of all terms")
fit.o.c$means

print("knot locations")
attr(fit.o.c$terms, "predvars")
```

```{r}
print("Method 2:")
print("Mean values of all terms")
centerList$vec_mean

print("knot locations")
centerList$rcsList

print("outcome mean at specific time points:")
centerList$outcome_mean
```

```{r}
fit.o.c.h0 <- basehaz(fit.o.c) 
fit.c.ho <- basehaz(fit.c) 
print("baseline hazard")
head(cbind(fit.o.c.h0$hazard, fit.c.ho$hazard))
```

**Note:** From the identical term mean values, it is verified that the `coxph` function also follows *transfer then center*.

```{r}
cbind(fit.o.c$means, centerList$vec_mean, fit.o.c$means - centerList$vec_mean)
```

## Transport the centered model to a new dataset

**Note:** Here, `center_predict` is a wrapper function of the corresponding code in Method 1, the main reason to have `center_predict` is to automatically incorporate `delta_recal`.

```{r}
# -------- Methods 1 -------- 
pred.m1.lp <- predict(fit.o.c, newdata = target, type = "lp")
pred.m1.risk <- predict(fit.o.c, newdata = target, type = "risk")
surv.fit <- survfit(fit.o.c, newdata = target)
pred.m1.survival <- t(summary(surv.fit, times = times)$surv)
pred.m1.prob <- 1 - pred.m1.survival
head(cbind(pred.m1.lp, pred.m1.risk, pred.m1.survival, pred.m1.prob))

# -------- Methods 2 -------- 
pred.m2.lp <- center_predict(centerList, newdata = target, type = "lp")
pred.m2.risk <- center_predict(centerList, newdata = target, type = "risk")
pred.m2.survival <- center_predict(centerList, newdata = target, type = "survival", times = times)
pred.m2.prob <- center_predict(centerList, newdata = target, type = "probability", times = times)
head(cbind(pred.m2.lp, pred.m2.risk, pred.m2.survival, pred.m2.prob))
```

Kaplan-Meier plots: https://www.emilyzabor.com/survival-analysis-in-r.html, which shows that at time = 3, the original data has higher survival probability, thus lower event probability, than the target data.

```{r}
data <- bind_rows(
  trainDataSurvival %>% select(ryear, rfs) %>% mutate(type = "original"),
  testDataSurvival %>% select(ryear, rfs) %>% mutate(type = "target")
)
data$type <- factor(data$type, levels = c("original", "target"))

survfit2(Surv(ryear, rfs) ~ type, data = data) %>% 
  ggsurvfit() +
  add_confidence_interval() +
  add_risktable()
```

**Note:**

1.  The `coxph` object cannot be directly updated to reflect the recalibration changes. The current work-around is to do it manually in `predict_centre`.

2.  Therefore, here `center_recalibrate` is mainly used to calculate the model shift, *i.e.* `delta_recal`.

3.  For re-calibration, we can only work with one single `timeHorizon` at a time.

4.  The specific `timeHorizon` should be within the pre-specified `times` when running `center_fit`.

5.  The method in doi:10.1016/j.jclinepi.2025.111895 is as following:

- suppose the risk/probability of event at a specific time, say, $t=10$, is $\theta = 1 - S_{KM}(10)$, then

  - Observed risk for derivation population is $\theta_{derived} = 1 - S^{derived}_{KM}(10)$

  - Observed risk for target population is $\theta_{target} = 1 - S^{target}_{KM}(10)$

the rescaling factor they proposed was $\Delta_{rescaling} = ln[-ln(1-\theta_{derived})] - ln[-ln(1-\theta_{target})]$.

- Then, the recalibrated risk is $risk_{recal} = 1 - exp(-exp(ln[-ln(1-risk_{ori})]-\Delta_{rescaling}))$.

- Let's substitute $S_{recal} = 1 - risk_{recal}$ and $S_{ori} = 1 - risk_{ori}$ into the equation, then

$S_{recal} = exp(-exp(ln[-ln(S_{ori})]-\Delta_{rescaling}))$

$= e^{-e^{ln[-ln(S_{ori})]-\Delta_{rescaling}}}$

$= e^{-e^{ln[-ln(S_{ori})]}e^{-\Delta_{rescaling}}}$

$= [e^{-e^{ln[-ln(S_{ori})]}}]^{e^{-\Delta_{rescaling}}}$

$= [e^{ln(S_{ori})}]^{e^{-\Delta_{rescaling}}}$

$= [S_{ori}]^{e^{-\Delta_{rescaling}}}$

- In the Cox model, the equation for hazard is $h(t|X_i) = h_0(t)exp(\beta^TX_i)$. And the survival function is $S(t|X_i) = [S_0(t)]^{exp(\beta^TX_i)}$. Therefore, the equation above can be written as $S_{recal} = [S_{ori}]^{e^{-\Delta_{rescaling}}}$

$= ([S_{0,ori}(t)]^{e^{\beta^TX_i}})^{e^{-\Delta_{rescaling}}}$

$= S_{0,ori}(t)^{e^{\beta^TX_i}e^{-\Delta_{rescaling}}}$

$= S_{0,ori}(t)^{e^{(\beta^TX_i -\Delta_{rescaling})}}$

- Note, in my implementation, I make `delta_recal` = $-\Delta_{rescaling} = ln[-ln(1-\theta_{target})] - ln[-ln(1-\theta_{derived})]$ (*i.e.* `outcome_mean_target_cloglog - outcome_mean_original_cloglog`).

- In summary, with `delta_recal`, the predictions will be updated as follows (see `center_predict`):

```         
- `type = 'lp'`: `pred_lp_recal = pred_lp + delta_recal`

- `type = 'risk'` (_i.e._ `exp(pred_lp)`): `pred_risk_recal = pred_risk * exp(delta_recal)`

- `type = 'surv'` (using `survfit`+`summary`): `pred_surv_recal = pred_surv ^ exp(delta_recal)`

- `type = 'probability'` (_i.e._ `1 - pred_surv`): `pred_prob_recal = 1 - pred_surv_recal = 1 - pred_surv ^ exp(delta_recal)`
```

```{r}
timeHorizon <- 3 # specify a time horizon for calculating 'outcome mean'
recalibrateList <- center_recalibrate(centerList, newdata = target, timeHorizon = timeHorizon)
fit.c.recalibrated <- recalibrateList$centerList$fit.c # this is still fit.c

# check the results
prob_original_model <- mean(center_predict(centerList, newdata = original, type = "probability", times = timeHorizon))
prob_original_model_cloglog <- log(-log(1-prob_original_model)) 
print(paste0("Original, average event probability at year 3: ", prob_original_model))
print(paste0("Clog-log: ", prob_original_model_cloglog))

survfit_original_KM <- survfit(Surv(ryear, rfs) ~ 1, data = original)
prob_original_KM <- 1 - summary(survfit_original_KM, times = timeHorizon)$surv
prob_original_KM_cloglog <- log(-log(1-prob_original_KM)) 
print(paste0("Original, KM event probability at year 3: ", prob_original_KM))
print(paste0("Clog-log: ", prob_original_KM_cloglog))
      
survfit_target_KM <- survfit(Surv(ryear, rfs) ~ 1, data = target)
prob_target_KM <- 1 - summary(survfit_target_KM, times = timeHorizon)$surv
prob_target_KM_cloglog <- log(-log(1-prob_target_KM)) 
print(paste0("Target, KM event probability at year 3: ", prob_target_KM))
print(paste0("Clog-log: ", prob_target_KM_cloglog))
```

# Calibration plots

Function `CalibrationCurves::valProbSurvival` is modified as `valProbSurvival.2` for the following reasons:

1.  `CalibrationCurves::valProbSurvival` requires `x = TRUE, y = TRUE`, thus does not work on the light-weight model object `fit.o.c`.

2.  Since `coxph` object cannot be directly updated to reflect the change made by recalibration, we need to use specific prediction vectors instead of the fitted model object.

3.  Some further modification for subgroups.

**Note:**

1.  See "3.1_Test_new_functions.qmd" that varified the two function produce identical results (the only thing lacking in `valProbSurvival.2` is the se and confidence interval for brier scores).

2.  For centered models with `centerList`, `valProbSurvival.2` is called through a wrapper function `calibration.surv`.

```{r}
# The original model using Method 1
pred.lp <- predict(fit.o.c, target, type = "lp")
surv.fit <- survfit(fit.o.c, newdata = target)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- target$ryear
outcome.event <- target$rfs

res.1 <- valProbSurvival.2(
  pred.lp,
  pred.prob,
  pred.prob.brier,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
# The original model using Method 2
res.2 <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
# The recalibrated model, indicated by `delta_recal_model`: the linear one
delta_recal_model <- prob_target_KM_cloglog - prob_original_model_cloglog
print(paste0("delta_recal_model: ", delta_recal_model))

res.recal.model <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  delta_recal = delta_recal_model,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
# The recalibrated model, indicated by `delta_recal_KM`: the linear one
delta_recal_KM <- prob_target_KM_cloglog - prob_original_KM_cloglog
print(paste0("delta_recal_KM: ", delta_recal_KM))

res.recal.KM <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  delta_recal = delta_recal_KM,
  plotCal = "ggplot",
  nk = 5
)
```

(2026-08-26) Based on the methods used in [SCORE2 risk prediction algorithms for CVD](doi:10.1093/eurheartj/ehab309), instead of single summary stats, we can create a vector for multiple age categories (also separated by sex if the model was sex-stratified) to enable Platt scaling using unlinked summary stats.

```{r}
summary(testDataSurvival$age)

df <- target
# Prepare data: dummy, interacteration, rcs, centering
df <- center_prepare(centerList, df)$data

test
# add age
df$age <- testDataSurvival$age
# bin age and 
df <- df %>%
  mutate(
    cage = cut(
      age,
      breaks = c(20, seq(40, 65, 5), 80)
    )
  ) 
table(df$cage)

# observed event probability in each age group at timeHorizon
fit.obs <- coxph(Surv(ryear, rfs) ~ cage, data = df)
surv.fit <- survfit(fit.obs, newdata = df)
pred.survival <- t(summary(surv.fit, times = timeHorizon)$surv)
df$pred.prob <- 1 - pred.survival

# Prepare data: dummy, interacteration, rcs, centering


df_unique <- df %>% 
  distinct(cage, pred.prob)

# get mean values for each variables in each age group
df2 <- df %>%
  select(-c(ryear, rfs, pred.prob)) %>% 
  group_by(cage) %>%
  summarise_all(mean)
df2$age_min <- rep(0, nrow(df2))
df2$age_max <- rep(0, nrow(df2))
df2$age_center <- rep(0, nrow(df2))
for (i in 1:nrow(df2)) {
  vec <- as.numeric(str_extract_all(df2$cage[i], "\\d+")[[1]])
  df2$age_min[i] <- vec[1]
  df2$age_max[i] <- vec[2]
  df2$age_center[i] <- mean(vec)
}
df2 <- df2 %>% left_join(df_unique, by = "cage")

df2$pred_lp <- center_predict(centerList, newdata = df, type = "lp")
df2$pred_prob <- center_predict(centerList, newdata = df, type = "probability", times = timeHorizon)
```

**Note:** the `offset(lp)` method doesn't work here

```{r}
# update intercept using Platt's calibration, which requires linked individual-level newdata
# Use `valProbSurvival.2` directly here

target$LP <- center_predict(centerList, newdata = target, type = "lp")
fit.offset <- coxph(Surv(ryear, rfs) ~ LP, data = target,
model = FALSE, x = FALSE, y = FALSE)

pred.lp <- predict(fit.offset, target, type = "lp")
surv.fit <- survfit(fit.offset, newdata = target)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- target$ryear
outcome.event <- target$rfs

res.Platt <- valProbSurvival.2(
  pred.lp,
  pred.prob,
  pred.prob.brier,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
res.1$stats$Calibration$Statistics
res.2$stats$Calibration$Statistics
res.recal.model$stats$Calibration$Statistics
res.recal.KM$stats$Calibration$Statistics
res.Platt$stats$Calibration$Statistics
```

## Subgroup

### Adding a grouping variable: categorical

```{r}
# The original model using Method 2
res.OvsP <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  group = target$grade3,
  plotCal = "ggplot",
  nk = 5
)
res.OvsP

calibration.OvsP(
  res.OvsP,
  group = target$grade3,
  group_con = FALSE,
  xlab = "grade3"
)
```

### Adding a grouping variable: continuous

```{r}
# The original model using Method 2
res.OvsP <- calibration.surv(
  centerList,
  newdata = target,
  timeHorizon = timeHorizon,
  group = target$pgr2,
  g.group = 4,
  plotCal = "ggplot",
  nk = 5
)
res.OvsP

calibration.OvsP(
  res.OvsP,
  group = target$pgr2,
  group_con = TRUE,
  xlab = "pgr2"
)

# If treating it as a categorical variable
calibration.OvsP(
  res.OvsP,
  group = target$pgr2,
  group_con = FALSE,
  xlab = "pgr2"
)
```

```{r}

```