---
title: "Simulation studies - histogram and non-linear terms"
author: "Juan Li"
date: 2026-04-13
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(dplyr) # Data manipulation
library(tidyr) # Tidy messy data
library(ggplot2) # Data visualization
library(ggpubr) # ggarrange
library(gtsummary)
library(rms)
library(foreach)
library(doParallel)

source("R/util.R")
# functions in "util.R"
# get_rcs: implement the formula of rcs components
# step_dummy: dummy variables
# step_rcs: rcs
# step_interaction: interaction terms
# step_center: centering
# get_mean: get mean values from the original dataset
# root.search: Root searching, not in use in this .qmd file
```

# Simulation study 1 - histogram and non-linear terms

In this simulation study, we will explore the extimated mean values of variour non-linear terms, using different bin-width.

Things to control:

-   Data distribution (N = 10000): uniform, normal, skewed, double-peak

-   Number of bins: seq(5,200,10)

-   Non-linear term:

    -   polynomial: x\^2, x\^3

    -   rcs: k = 4; same distribution, different distribution

The steps are:

1.  Genarate data with different distributions. - **real data**

2.  For each data, generate its histogram using different nbins.

3.  Generate data from each histogram (500 samples) - **simulated data**

4.  Get the mean values of each term from the **simulated data** and compare with the ones from the **real data**.

```{r}
# ------------------- Various non-linear terms and their mean values -------------------
non.linear.mean <- function(data, knots.diff, knots.same = NULL, k = 4) {
  data$x2 <- data$x^2
  data$x3 <- data$x^3
  # rcs: same distribution
  if (is.null(knots.same)) {
    rcs.same <- rcs(data$x, k)
    knots.same <- attributes(rcs.same)$parms
  }
  rcs.res.same <- get_rcs(data$x, knots.same)
  data$rcs.2.same <- rcs.res.same[, 'rcs.2']
  data$rcs.3.same <- rcs.res.same[, 'rcs.3']
  # rcs: different distribution
  rcs.res.diff <- get_rcs(data$x, knots.diff)
  data$rcs.2.diff <- rcs.res.diff[, 'rcs.2']
  data$rcs.3.diff <- rcs.res.diff[, 'rcs.3']

  vec_mean <- colMeans(data)
  return(vec_mean = vec_mean)
}

# ------------------- Histogram and simulate data from it -------------------
hist.simu <- function(data, nbin, vec_seed, knots.diff) {
  k <- length(knots.diff)
  # rcs: same distribution
  rcs.same <- rcs(data$x, k)
  knots.same <- attributes(rcs.same)$parms

  # histogram
  built_plot <- ggplot_build(ggplot(data, aes(x)) + geom_histogram(bins = nbin))
  data.hist <- data.frame(
    nbin = rep(nbin, nbin)
  )
  data.hist <- bind_cols(data.hist, built_plot$data[[1]][, 1:4])

  # simulate data from the histogram, get non-linear terms and their mean values
  #df_mean <- data.frame()

  # Parallel
  # Determine the number of cores to use (it's best to leave one free)
  n_cores <- parallel::detectCores() - 1

  # Create and register the cluster
  my_cluster <- makeCluster(n_cores)
  registerDoParallel(my_cluster)

  df_mean <- foreach(
    i = 1:500,
    .packages = c("ggplot2", "dplyr", "rms"),
    .export = c(
      "non.linear.mean",
      "get_rcs"
    ),
    .combine = "rbind"
  ) %dopar%
    {
      set.seed(vec_seed[i])
      vec <- c()
      for (j in 1:nbin) {
        temp <- runif(
          n = data.hist$count[j],
          min = data.hist$xmin[j],
          max = data.hist$xmax[j]
        )
        vec <- c(vec, temp)
      }
      data_i <- data.frame(x = vec)
      vec_mean <- non.linear.mean(data_i, knots.diff, knots.same, k)
      return(vec_mean)
    }

  # Stop the cluster
  stopCluster(my_cluster)

  vec_025 <- apply(df_mean, 2, quantile, probs = 0.025, na.rm = TRUE)
  vec_50 <- apply(df_mean, 2, quantile, probs = 0.5, na.rm = TRUE)
  vec_975 <- apply(df_mean, 2, quantile, probs = 0.975, na.rm = TRUE)
  res_mean <- data.frame(
    term = names(vec_50),
    nbin = rep(nbin, ncol(df_mean)),
    mean = unname(vec_50),
    mean.lower = unname(vec_025),
    mean.upper = unname(vec_975)
  )

  return(list(data.hist = data.hist, res_mean = res_mean))
}
```

```{r}
# -------------------  set up the simulation -------------------
vec_N <- c(500, 1000, 5000, 10000) # vector of N
vec_type <- c("unif", "norm", "skew", "double") # vector of data type
vec_nbin <- c(5, seq(10, 100, 10)) # vector of nbin
# vector of random seeds
set.seed(123)
vec_seed <- sample.int(1e6, 500)

# ------------------- rcs knots from a different distribution -------------------
k <- 4 # number of knots
dist.diff <- rbeta(n = 10000, shape1 = 2, shape2 = 10)
rcs.diff <- rcs(dist.diff, k)
knots.diff <- attributes(rcs.diff)$parms

res_hist <- data.frame()
res_real <- data.frame()
for (N in vec_N) {
  print(N)
  file_name_hist <- paste0("Simulation 4.1 results/mean_hist_", N, ".csv")
  file_name_real <- paste0("Simulation 4.1 results/mean_real_", N, ".csv")

  if (file.exists(file_name_hist)) {
    mean_hist <- read.csv(file_name_hist)
    mean_real_long <- read.csv(file_name_real)
  } else {
    # ------------------- Generate data -------------------
    set.seed(123)
    unif <- runif(n = N, min = 0, max = 1)
    norm <- rnorm(n = N, mean = 5, sd = 2)
    skew <- rbeta(n = N, shape1 = 10, shape2 = 2)
    double <- c(
      rnorm(n = N / 2, mean = 5, sd = 2),
      rnorm(n = N / 2, mean = 12, sd = 2)
    )
    data <- data.frame(
      type = rep(vec_type, each = N),
      x = c(unif, norm, skew, double)
    )
    data$type <- factor(data$type, levels = vec_type)

    # ------------------- Non-linear terms of real data -------------------
    mean_real <- data.frame()
    for (i_type in vec_type) {
      data_i <- data %>%
        filter(type == i_type) %>%
        droplevels() %>%
        select(-type)
      vec_mean <- non.linear.mean(data_i, knots.diff)
      mean_real <- bind_rows(mean_real, c("type" = i_type, vec_mean))
    }
    mean_real$type <- factor(mean_real$type, levels = vec_type)

    mean_real_long <- mean_real %>%
      pivot_longer(-type, names_to = "term", values_to = "mean")
    mean_real_long$type <- factor(mean_real_long$type, levels = vec_type)
    mean_real_long$term <- factor(
      mean_real_long$term,
      levels = names(mean_real)[-1]
    )
    mean_real_long$mean <- as.numeric(mean_real_long$mean)
    mean_real_long$N <- rep(N, nrow(mean_real_long))
    write.csv(mean_real_long, row.names = FALSE, file = file_name_real)

    # ------------------- Generate data from histogram -------------------
    mean_hist <- data.frame()
    for (i_type in vec_type) {
      data_i <- data %>%
        filter(type == i_type) %>%
        droplevels() %>%
        select(-type)
      for (nbin in vec_nbin) {
        temp <- hist.simu(data_i, nbin, vec_seed, knots.diff)$res_mean
        temp$type <- rep(i_type, nrow(temp))
        mean_hist <- bind_rows(mean_hist, temp)
      }
    }
    mean_hist$N <- rep(N, nrow(mean_hist))
    mean_hist$type <- factor(mean_hist$type, levels = vec_type)
    mean_hist$term <- factor(mean_hist$term, levels = names(mean_real)[-1])
    write.csv(mean_hist, row.names = FALSE, file = file_name_hist)

    # ------------------- Plot -------------------
    pList <- list()
    for (i_type in vec_type) {
      data_i <- data %>% filter(type == i_type) %>% droplevels()
      for (nbin in c(5, 20, 40, 60, 80, 100)) {
        p <- ggplot(data_i, aes(x)) +
          geom_histogram(
            aes(y = ..density..),
            bins = nbin,
            alpha = 0.2,
            color = "black"
          ) +
          geom_density(color = "red") +
          theme_bw() +
          labs(title = paste0("N=", N, ", ", nbin, " bins"))
        pList[[length(pList) + 1]] <- p
      }
    }
    print(ggarrange(plotlist = pList, ncol = 6, nrow = 4))
  }

  res_real <- bind_rows(res_real, mean_real_long)
  res_hist <- bind_rows(res_hist, mean_hist)
}
```

```{r}
vec_term <- c()
res_hist$term <- factor(res_hist$term, levels = unique(res_hist$term))
res_real$term <- factor(res_real$term, levels = unique(res_real$term))

# Plot
levels(res_hist$term) <- c(
  "x",
  "x^2",
  "x^3",
  "rcs.2.same",
  "rcs.3.same",
  "rcs.2.diff",
  "rcs.3.diff"
)
levels(res_real$term) <- c(
  "x",
  "x^2",
  "x^3",
  "rcs.2.same",
  "rcs.3.same",
  "rcs.2.diff",
  "rcs.3.diff"
)

for (i_type in vec_type) {
  p <- ggplot(
    res_hist %>% filter(type == i_type, term != "x") %>% droplevels(),
    aes(x = nbin, y = mean, color = factor(N))
  ) +
    geom_point() +
    geom_linerange(aes(ymin = mean.lower, ymax = mean.upper)) +
    geom_line(alpha = 0.5) +
    geom_hline(
      data = res_real %>% filter(type == i_type, term != "x") %>% droplevels(),
      aes(yintercept = mean, color = factor(N)),
      linetype = "dashed"
    ) +
    facet_wrap(~term, scales = "free", nrow = 3) +
    scale_x_continuous(breaks = vec_nbin) +
    theme_bw() +
    labs(title = i_type, color = "N") +
    theme(text = element_text(size = 12))
  print(p)
}
```

```{r}

```