---
title: "Test the new functions for survival models: fine gray"
author: "Juan Li"
date: 2026-06-30
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(CalibrationCurves)
library(survival)
library(ggsurvfit)
library(gtsummary)
library(riskRegression)
library(prodlim)

# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_predict.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
```

# Monoclonal gammopathy data

See `?survival::mgus2`. A data frame with 1384 observations on the following 10 variables.

`id`: subject identifier

`age`: age at diagnosis, in years

`sex`: a factor with levels F M

`dxyr`: year of diagnosis

`hgb`: hemoglobin

`creat`: creatinine

`mspike`: size of the monoclonal serum splike

`ptime`: time until progression to a plasma cell malignancy (PCM) or last contact, in months

`pstat`: occurrence of PCM: 0=no, 1=yes (**event of interest**)

`futime`: time until death or last contact, in months

`death`: occurrence of death: 0=no, 1=yes (**competing risk**)

```{r}
data <- mgus2
# remove rows with missing
data <- data[complete.cases(data), ]
```

# Model fitting: `riskRegression::FGR` (wrapper of `cmprsk::crr`) vs `survival::finegray` + `coxph`

I was interested to also have a version for this function, the main benefit is its `riskRegression::predictRisk` and `riskRegression::plotCalibration` functions. However, `riskRegression::FGR` handle non-linear terms in a base R way:

1.  polynomials: using `I(X1) + I(X1^2)` instead of `poly(X1, 2, , raw = TRUE)`

2.  rcs: the formula wrapper does not natively parse built-in spline functions like rms::rcs. Instead, you must manually generate the spline basis variables and pass them as standard linear terms.

3.  It has limited ability to work with more complicated variables names, e.g. `csize20-50`, `rcs(pgr2, 3)pgr2'`.

All of above can be solved with some functions to prepare the data, but I want to discuss and see if this is necessary.

**Note:**

1.  As discussed above, since the current `riskRegression::FGR` has some syntax issues with `poly`, `rcs`, and interaction, we will just have a simple formula in this file.

2.  `riskRegression::FGR` using `Hist` not `Surv`

3.  `survival::finegray` creates a expanded, weighted dataset, which has more rows then the original dataset. `riskRegression::FGR` doesn't need it, but instead using a multi-level outcome variable `status`: 0 = censor, 1 = event1, 2 = event2, and then specify the event of interest using the parameter `cause`.

4.  Below, I compared results from `riskRegression::FGR` vs `survival::finegray` + `coxph` and found the model coefficients were close but not identical. This is a known phenomenon and it's because the two functions use different approaches to weighting, baseline hazard handling, and ties.

5.  *Internal Centering*:

- `coxph` centers covariates automatically by subtracting the mean of the covariates (set `nocenter = NULL` to also center on dummy variables), improving convergence speed in iterative Newton-Raphson algorithms.

- `FGR` does **not** natively perform internal variable centering. It internally relies on binomial regression and pseudo-values rather than partial likelihoods, the internal centering and constraint handling differ.

6.  *Baseline Hazard*:

- `finegray` + `coxph`: you can extract the baseline cumulative subdistribution hazard (e.g., using `survival::basehaz`). However, *Subdistribution Hazard (Fine-Gray)* models the instantaneous rate of the event among those who are event-free or have already had a competing event. The baseline hazard here is difficult to interpret clinically on its own because it is a mathematical construct rather than a true population rate. (**Discussion:** what population summary statistics can be used for recalibration?)

- `FGR` utilizes a transformation model or binomial regression on time sequences rather than estimating a traditional partial-likelihood baseline hazard. It inherently handles absolute risk regression, mapping time to the cumulative incidence of the specific event of interest.

7.  *Standard Errors:* Because `finegray` manipulates the risk set size by duplicating records, you must use robust standard errors (e.g., `robust = TRUE` in `coxph`, which also requires the `id` column, _i.e.,_ `id = id, robust = TRUE`). `riskRegression::FGR` uses a pseudo-value approach that naturally handles these without needing to inflate the risk set size identically.

```{r}
# --------- The FGR model -----------
# Prepare dataset and event variable:
# 0 = censor, 1 = pcm (event of interest), 2 = death (competing risk)
data$time <- ifelse(data$pstat == 0, data$futime, data$ptime)
data$status <- ifelse(data$pstat == 0, 2 * data$death, 1)

# Fit the weighted Cox model on the Fine-Gray data
fit.FGR <- FGR(
  # Using `Hist` instead of `Surv`, only includes linear terms now
  formula = Hist(time, status) ~ hgb + creat + mspike + age, 
  data = data, 
  # specify the event of interest
  cause = 1
)
```

```{r}
# --------- The finegray + coxph model -----------
# Fine-Gray model requires a multi-state survival, need to make it a factor
data$status_fct <- factor(data$status)

# Create the Fine-Gray weighted dataset
data_fg <- finegray(Surv(time, status_fct) ~ ., data = data)

# Fit the weighted Cox model on the Fine-Gray data
fit.survival <- coxph(
  Surv(fgstart, fgstop, fgstatus) ~ hgb + creat + mspike + age, 
  # using the Fine-Gray weighted dataset
  data = data_fg, weight = fgwt,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL,
  # MUST: for robust standard errors
  id = id, robust = TRUE)

# IMPORTANT: the mean values in `fit.survival` are of `data_fg` not `data`!!!
print("Mean values in `fit.survival`: ")
fit.survival$means
print("Corresponding mean values in `data_fg`: ")
X_fg <- data_fg %>% select(hgb, creat, mspike, age)
colMeans(X_fg)
print("Corresponding mean values in `data`: ")
X <- data %>% select(hgb, creat, mspike, age)
colMeans(X)
```

```{r}
# --------- Compare results of the two models -----------
print("Coefficients: similar but not identical")
cbind(fit.FGR$crrFit$coef, fit.survival$coefficients)

print("Predicted linear predictors: different by a constant due to centering")
# `FGR`
X <- as.matrix(data %>% select(hgb, creat, mspike, age))
coef <- fit.FGR$crrFit$coef
lp.FGR <- as.vector(X %*% coef)
# `survival`
lp.survival <- predict(fit.survival, newdata = data, type = "lp")
print(paste0("Sum of variable means times coefficients: ", colMeans(X_fg) %*% coef))
head(cbind(lp.FGR, lp.survival, lp.survival - lp.FGR))
```

# Performance metrics

```{r}
# Define the time point for evaluation
timeHorizon <- 300
```

## Brier

Even with `x = TRUE, y = TRUE`, the commonly used functions have issues directly working with `fit.survival.2`:

1. `survival::brier` only works with the cox model: "Error in `brier()`: ! one or more flags are >0 in survcheck".

2. `riskRegression::Score`: "Error in `predictCox()`: ! predictCox does not know how to handle Cox models fitted with weights"

3. `pec::pec`: "Error in `UseMethod()`: ! no applicable method for 'predictEventProb' applied to an object of class "coxph""

```{r}
# Update `fit.survival` with `x = TRUE, y = TRUE` 
# to test if it can work with some evaluation functions
# fit.survival.2 <- update(fit.survival, x = TRUE, y = TRUE)
```

```{r}
# brier(fit.survival.2, times = timeHorizon)
```  

```{r}
# Score(
#   object = list("Fine-Gray" = fit.survival.2),
#   formula = Hist(time, status_fct) ~ 1,      # Baseline formula matching your outcome format
#   data = data,                          # Original validation data format
#   metrics = c("brier"),           # Concordance (AUC) and Brier Score
#   plots = "calibration",                 # Prepares calibration data
#   cause = 1,                             # Cause of interest
#   times = timeHorizon                    # Your arbitrary evaluation time
# )
```

```{r}
# pec::pec(list(FineGray = fit.survival), 
#                data = data, 
#                formula = Surv(time, status_fct) ~ 1, 
#                times = timeHorizon)
```

Therefore, I modified `brier.cox` to `brier.fg` to calculate brier score for `fit.survival`, and use results of `riskRegression::Score` on `fit.FGR` as the benchmark.

```{r}
timeHorizon <- 300
eval_FGR <- Score(
  object = list("Fine-Gray" = fit.FGR),
  formula = Hist(time, status) ~ 1,      # Baseline formula matching your outcome format
  data = data,                          # Original validation data format
  metrics = c("auc", "brier"),           # Concordance (AUC) and Brier Score
  plots = "calibration",                 # Prepares calibration data
  cause = 1,                             # Cause of interest
  times = timeHorizon                    # Your arbitrary evaluation time
)

# Brier Score
eval_FGR$Brier$score
```

The main question here is which dataset to use, `data` or `data_fg`? This applies to both the fine-gray model and the NULL model.

```{r}
surv.fit <- survfit(fit.survival, data = data_fg)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
fgstart <- data_fg$fgstart
fgstop <- data_fg$fgstop
fgstatus <- data_fg$fgstatus
fgwt <- data_fg$fgwt
brier.fg(pred.prob, fgstart, fgstop, fgstatus, fgwt, times = timeHorizon)
```

```{r}
mean(data$age)
data$age_C <- data$age - mean(data$age)

data_fg_1 <- finegray(Surv(time, status_fct) ~ age, data = data)
mean(data_fg_1$age)
data_fg_2 <- finegray(Surv(time, status_fct) ~ age_C, data = data)
mean(data_fg_2$age_C)
```


```{r}

```