---
title: "Test the new functions for survival models: fine gray"
author: "Juan Li"
date: 2026-06-30
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(CalibrationCurves)
library(survival)
library(ggsurvfit)
library(gtsummary)
library(riskRegression)
library(prodlim)

source("R/valProbSurvival.fg.R", echo = FALSE)
```

# Monoclonal gammopathy data

See `?survival::mgus2`. A data frame with 1384 observations on the following 10 variables.

`id`: subject identifier

`age`: age at diagnosis, in years

`sex`: a factor with levels F M

`dxyr`: year of diagnosis

`hgb`: hemoglobin

`creat`: creatinine

`mspike`: size of the monoclonal serum splike

`ptime`: time until progression to a plasma cell malignancy (PCM) or last contact, in months

`pstat`: occurrence of PCM: 0=no, 1=yes (**event of interest**)

`futime`: time until death or last contact, in months

`death`: occurrence of death: 0=no, 1=yes (**competing risk**)

```{r}
data <- mgus2
# remove rows with missing
data <- data[complete.cases(data), ]

# Prepare dataset and event variable:
# 0 = censor, 1 = pcm (event of interest), 2 = death (competing risk)
data$time <- ifelse(data$pstat == 0, data$futime, data$ptime)
data$status <- ifelse(data$pstat == 0, 2 * data$death, 1)
data$status <- factor(data$status)
```

```{r}
# Create the Fine-Gray weighted dataset
data_fg <- finegray(Surv(time, status) ~ ., data = data)

# Fit the weighted Cox model on the Fine-Gray data
fit <- coxph(
  Surv(fgstart, fgstop, fgstatus) ~ hgb + creat + mspike + age, 
  # using the Fine-Gray weighted dataset
  data = data_fg, weight = fgwt,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL,
  # MUST: for robust standard errors
  id = id, robust = TRUE)
```

__Note 1:__ The mean values in `fit` are of `data_fg` not `data`!!!

```{r}
print("Mean values in `fit`: ")
fit$means
print("Corresponding mean values in `data_fg`: ")
X_fg <- data_fg %>% select(hgb, creat, mspike, age)
colMeans(X_fg)
print("Corresponding mean values in `data`: ")
X <- data %>% select(hgb, creat, mspike, age)
colMeans(X)
```

__Note 2:__ Predicted linear predictors are identical using the below two methods

```{r}
# Method 1
pred.lp.fg <- predict(fit, newdata = data_fg, type = "lp")

# Method 2
pred.lp <- predict(fit, newdata = data, type = "lp")
df <- data %>% select(time, status)
df$LP <- pred.lp
df_fg <- finegray(Surv(time, status) ~ ., data = df)

all(pred.lp.fg == df_fg$LP)
```

# Performance metrics using `valProbSurvival.fg`

```{r}
# Define the time point for evaluation
timeHorizon <- 300

# Prediction and outcome in the dataset of interest,
# here is the detivation dataset, but could be any newdata
pred.lp <- predict(fit, newdata = data, type = "lp")
surv.fit <- survfit(fit, data)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- data$time
outcome.event <- data$status

res <- valProbSurvival.fg(
    pred.lp,
    pred.prob,
    pred.prob.brier,
    outcome.time,
    outcome.event,
    # IMPORTANT, make sure this `timeHorizon` matches the one used in `pred.prob` and `pred.prob.brier`
    timeHorizon = timeHorizon, 
    plotCal = "ggplot", 
    nk = 5
  )
```

# Compare with results using `riskRegression`

```{r}
# --------- The FGR model -----------
# Fit the weighted Cox model on the Fine-Gray data
fit.FGR <- FGR(
  # Using `Hist` instead of `Surv`, only includes linear terms now
  formula = Hist(time, status) ~ hgb + creat + mspike + age, 
  data = data, 
  # specify the event of interest
  cause = 1
)

eval_FGR <- Score(
  object = list("Fine-Gray" = fit.FGR),
  formula = Hist(time, status) ~ 1,      # Baseline formula matching your outcome format
  data = data,                          # Original validation data format
  metrics = c("auc", "brier"),           # Concordance (AUC) and Brier Score
  plots = "calibration",                 # Prepares calibration data
  cause = 1,                             # Cause of interest
  times = timeHorizon,                    # Your arbitrary evaluation time
  summary = "ipa"
)

plotCalibration(eval_FGR, times=timeHorizon, models=c("Fine-Gray"))
```

```{r}
# --------- Compare metrics -----------
print("AUC: noticable differences. Note `valProbSurvival.fg` also calculates time dependent AUC using `timeROC`")
res$stats$Concordance
res$stats$TimeDependentAUC
eval_FGR$AUC$score

print("Brier: almost identical. Note I used `riskRegression::Score` inside `valProbSurvival.fg`")
res$stats$Calibration$BrierScore
eval_FGR$Brier$score

print("Calibration: `riskRegression::Score` does not output calibration summary stats.")
res$stats$Calibration$InTheLarge
res$stats$Calibration$Statistics
```

# Compare with results using "calculate-performance-metrics.R" 

```{r}
source("R/calculate-performance-metrics.R", echo = FALSE)

# ------------ Compare metrics ---------
print("AUC: almost identical, slight difference in confidence intervals.")
res$stats$Concordance
# C-statistic (ours)
compute_c_statistic_fg(fit, data, time_to_event = "time", event_type = "status")

print("Brier: almost identical.")
res$stats$Calibration$BrierScore
# Brier (ours)
compute_brier_fg(pred.prob.brier, data, timeHorizon - 0.01, time_to_event = "time", event_type = "status")
```

# Trick in testing: loading defalut values of a function into the global environment

```{r}
# # 1. Safely extract all arguments from the function
# vars <- formals(valProbSurvival.fg)

# # 2. Drop required arguments (like fit and valdata) that lack a default value
# defaults_only <- vars[sapply(vars, function(x) !is.symbol(x) && x != "")]

# # 3. Unpack the actual default values into your Global Environment
# list2env(defaults_only, envir = .GlobalEnv)
```

```{r}

```