---
title: "Test the new functions for survival models: fine gray"
author: "Juan Li"
date: 2026-06-30
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

# ======== For unknown issue of connecting to R server in positron, load all packages and source code here =========
library(rms)
library(Hmisc)
library(dplyr)
library(stringr)
library(survival)
library(ggplot2)
library(ggpubr) 
library(CalibrationCurves)
library(riskRegression)
library(ggsurvfit)
library(gtsummary)
library(timeROC)

# Prepare data for the centered algorithm
source("R/center_prepare.R", echo = FALSE) 
# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_pred.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
# Wrapper function for calibration curves and metrics
# Including the following functions: `calibration`, `calibration.con`, `calibration.OvsP`
source("R/calibration.R", echo = FALSE)
# Modified function for cox models
source("R/calibration.surv.R", echo = FALSE) 
source("R/calibration_curve_fg.R", echo = FALSE) # valProbSurvival.fg
```

# Monoclonal gammopathy data

See `?survival::mgus2`. A data frame with 1384 observations on the following 10 variables.

`id`: subject identifier

`age`: age at diagnosis, in years

`sex`: a factor with levels F M

`dxyr`: year of diagnosis

`hgb`: hemoglobin

`creat`: creatinine

`mspike`: size of the monoclonal serum splike

`ptime`: time until progression to a plasma cell malignancy (PCM) or last contact, in months

`pstat`: occurrence of PCM: 0=no, 1=yes (**event of interest**)

`futime`: time until death or last contact, in months

`death`: occurrence of death: 0=no, 1=yes (**competing risk**)

```{r}
data <- mgus2
# remove rows with missing
data <- data[complete.cases(data), ]

# Prepare dataset and event variable:
# 0 = censor, 1 = pcm (event of interest), 2 = death (competing risk)
data$time <- ifelse(data$pstat == 0, data$futime, data$ptime)
data$status <- ifelse(data$pstat == 0, 2 * data$death, 1)
data$status <- factor(data$status)
```

```{r}
# Create the Fine-Gray weighted dataset
data_fg <- finegray(Surv(time, status) ~ ., data = data)

# Fit the weighted Cox model on the Fine-Gray data
fit <- coxph(
  Surv(fgstart, fgstop, fgstatus) ~ hgb + creat + mspike + age, 
  # using the Fine-Gray weighted dataset
  data = data_fg, weight = fgwt,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL,
  # MUST: for robust standard errors
  id = id, robust = TRUE)
```

**Note 1:** The mean values in `fit` are of `data_fg` not `data`!!!

```{r}
print("Mean values in `fit`: ")
fit$means
print("Corresponding mean values in `data_fg`: ")
X_fg <- data_fg %>% select(hgb, creat, mspike, age)
colMeans(X_fg)
print("Corresponding mean values in `data`: ")
X <- data %>% select(hgb, creat, mspike, age)
colMeans(X)
```

**Note 2:** Predicted linear predictors are identical using the below two methods

```{r}
# Method 1
pred.lp.fg <- predict(fit, newdata = data_fg, type = "lp")

# Method 2
pred.lp <- predict(fit, newdata = data, type = "lp")
df <- data %>% select(time, status)
df$LP <- pred.lp
df_fg <- finegray(Surv(time, status) ~ ., data = df)

all(pred.lp.fg == df_fg$LP)
```

# Performance metrics using `valProbSurvival.fg`

```{r}
# Define the time point for evaluation
timeHorizon <- 300

# Prediction and outcome in the dataset of interest,
# here is the detivation dataset, but could be any newdata
pred.lp <- predict(fit, newdata = data, type = "lp")
surv.fit <- survfit(fit, data)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- data$time
outcome.event <- data$status

res <- valProbSurvival.fg(
    pred.lp,
    pred.prob,
    pred.prob.brier,
    outcome.time,
    outcome.event,
    # IMPORTANT, make sure this `timeHorizon` matches the one used in `pred.prob` and `pred.prob.brier`
    timeHorizon = timeHorizon, 
    plotCal = "ggplot", 
    nk = 4
  )
```

# Compare with results using "calculate-performance-metrics.R"

```{r}
source("R/calculate-performance-metrics_JL.R", echo = FALSE)

# ------------ Compare metrics ---------
print("AUC: almost identical, slight difference in confidence intervals.")
res$stats$Concordance
# C-statistic (ours)
compute_c_statistic_fg(fit, data, time_to_event = "time", event_type = "status")

print("Brier: almost identical. Note `valProbSurvival.fg` uses `riskRegression::Score` for Brier.")
res$stats$Calibration$BrierScore
# Brier (ours)
compute_brier_fg(pred.prob.brier, data, timeHorizon - 0.01, time_to_event = "time", event_type = "status")

# Calibration curve
res_cal <- compute_fg_calibration_curve(fit, data, horizon = timeHorizon, n_knots = 4,
time_to_event = "time", event_type = "status")
lim <- range(c(res_cal$curve$predicted, res_cal$curve$observed))
ggplot(res_cal$curve, aes(x = predicted, y = observed)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_rug(data = data.frame(p = res_cal$p_hat), aes(x = p, y = NULL),
           sides = "b", alpha = 0.2) +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1)) +
  theme_minimal()

print("ICI: not identical")
res$stats$Calibration$Statistics
compute_ici_fg(pred.prob, data, horizon = timeHorizon, time_to_event = "time", event_type = "status")
```

```{r}
# Define the time point for evaluation
timeHorizon <- 300
nk <- 4

pred.lp <- predict(fit, newdata = data, type = "lp")
surv.fit <- survfit(fit, data)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
outcome.time <- data$time
outcome.event <- data$status

valdata <- data.frame(
  time = outcome.time,
  event = outcome.event,
  LP = pred.lp,
  pred = pred.prob
)
# make sure the variables names are correct
names(valdata) <- c("time", "event", "LP", "pred")
# Create the Fine-Gray weighted dataset
valdata_fg <- finegray(Surv(time, event) ~ ., data = valdata)

# ------- Compare `survival` and `rms` -------
# Results are (almost) identical
library(rms)
fit.survival <- coxph(
  Surv(fgstart, fgstop, fgstatus) ~ rcs(pred, nk),
  data = valdata_fg,
  weight = fgwt,
  # also center on dummy variables
  nocenter = NULL
)
pred.survival <- 1 -
  t(
    summary(
      survfit(fit.survival, newdata = valdata_fg),
      times = timeHorizon
    )$surv
  )

calRCSFormula <- as.formula(paste0(
  "Surv(fgstart, fgstop, fgstatus) ~ rcs(pred, ",
  nk,
  ")"
))
fit.rms <- cph(
  Surv(fgstart, fgstop, fgstatus) ~ rcs(pred, nk),
  x = TRUE,
  y = TRUE,
  surv = TRUE,
  data = valdata_fg,
  # IMPORTANT to use the weight
  weights = fgwt
)
pred.rms <- 1 - survest(fit.rms, times = timeHorizon, newdata = valdata_fg)$surv
cbind(fit.survival$coefficients, fit.rms$coefficients)
summary(pred.survival - pred.rms)

# ------------- ICI ---------
res$stats$Calibration$Statistics

dat_cal <- cbind.data.frame(
  obs.survival = pred.survival,
  obs = 1 - survest(fit.rms, times = timeHorizon, newdata = valdata_fg)$surv,
  lower = 1 - survest(fit.rms, times = timeHorizon, newdata = valdata_fg)$upper,
  upper = 1 - survest(fit.rms, times = timeHorizon, newdata = valdata_fg)$lower,
  pred = valdata_fg$pred
)
absdiff_cph <- abs(dat_cal$pred - dat_cal$obs)
(res_stats <- c(
  ICI = mean(absdiff_cph),
  setNames(quantile(absdiff_cph, c(0.5, 0.9)), c("E50", "E90")),
  Emax = max(absdiff_cph)
))

absdiff_cph.2 <- abs(dat_cal$pred - dat_cal$obs.survival)
(res_stats <- c(
  ICI = mean(absdiff_cph.2),
  setNames(quantile(absdiff_cph.2, c(0.5, 0.9)), c("E50", "E90")),
  Emax = max(absdiff_cph.2)
))


# ---------- `compute_ici_fg ` -------------
compute_ici_fg_2(valdata$pred, data, horizon = timeHorizon, n_knots = nk, time_to_event = "time", event_type = "status")
```

# Compare with results using `riskRegression`

```{r}
# --------- The FGR model -----------
# Fit the weighted Cox model on the Fine-Gray data
fit.FGR <- FGR(
  # Using `Hist` instead of `Surv`, only includes linear terms now
  formula = Hist(time, status) ~ hgb + creat + mspike + age, 
  data = data, 
  # specify the event of interest
  cause = 1
)

eval_FGR <- Score(
  object = list("Fine-Gray" = fit.FGR),
  formula = Hist(time, status) ~ 1,      # Baseline formula matching your outcome format
  data = data,                          # Original validation data format
  metrics = c("auc", "brier"),           # Concordance (AUC) and Brier Score
  plots = "calibration",                 # Prepares calibration data
  cause = 1,                             # Cause of interest
  times = timeHorizon,                    # Your arbitrary evaluation time
  summary = "ipa"
)

plotCalibration(eval_FGR, times=timeHorizon, models=c("Fine-Gray"))
```

```{r}
# --------- Compare metrics -----------
print("AUC: noticable differences. Note `valProbSurvival.fg` also calculates time dependent AUC using `timeROC`")
res$stats$Concordance
res$stats$TimeDependentAUC
eval_FGR$AUC$score

print("Brier: almost identical. Note I used `riskRegression::Score` inside `valProbSurvival.fg`")
res$stats$Calibration$BrierScore
eval_FGR$Brier$score

print("Calibration: `riskRegression::Score` does not output calibration summary stats.")
res$stats$Calibration$InTheLarge
res$stats$Calibration$Statistics
```

# Trick in testing: loading defalut values of a function into the global environment

```{r}
# 1. Safely extract all arguments from the function
vars <- formals(valProbSurvival.fg)

# 2. Drop required arguments (like fit and valdata) that lack a default value
defaults_only <- vars[sapply(vars, function(x) !is.symbol(x) && x != "")]

# 3. Unpack the actual default values into your Global Environment
list2env(defaults_only, envir = .GlobalEnv)
```

```{r}

```