---
title: "Examples of the proper center and recalibrate pipeline: fine gray, using `survival::finegray` and `survival::coxph`"
author: "Juan Li"
date: 2026-06-18
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

# ======== For unknown issue of connecting to R server in positron, load all packages and source code here =========
library(rms)
library(Hmisc)
library(dplyr)
library(stringr)
library(survival)
library(ggplot2)
library(ggpubr) 
library(CalibrationCurves)
library(riskRegression)
library(ggsurvfit)
library(gtsummary)
library(timeROC)

# Prepare data for the centered algorithm
source("R/center_prepare.R", echo = FALSE) 
# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_pred.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
# Wrapper function for calibration curves and metrics
# Including the following functions: `calibration`, `calibration.con`, `calibration.OvsP`
source("R/calibration.R", echo = FALSE)
# Modified function for cox models
source("R/calibration.surv.R", echo = FALSE) 
source("R/calibration_curve_fg.R", echo = FALSE) # valProbSurvival.fg
```

# Monoclonal gammopathy data

See `?survival::mgus2`. A data frame with 1384 observations on the following 10 variables.

`id`: subject identifier

`age`: age at diagnosis, in years

`sex`: a factor with levels F M

`dxyr`: year of diagnosis

`hgb`: hemoglobin

`creat`: creatinine

`mspike`: size of the monoclonal serum splike

`ptime`: time until progression to a plasma cell malignancy (PCM) or last contact, in months

`pstat`: occurrence of PCM: 0=no, 1=yes (**event of interest**)

`futime`: time until death or last contact, in months

`death`: occurrence of death: 0=no, 1=yes (**competing risk**)

```{r}
data <- mgus2
# remove rows with missing
data <- data[complete.cases(data), ]

# data split: to extragrate the difference between `original` and `target`
original <- data %>% filter(sex == "M") %>% select(-sex, -id)
summary(original)
target <- data %>% filter(sex == "F") %>% select(-sex, -id)
summary(target)
```

Cumulative incidence for competing risks, see [here](https://www.emilyzabor.com/survival-analysis-in-r.html#part-3-competing-risks)

```{r}
# Prepare dataset and event variable:
# 0 = censor, 1 = pcm (event of interest), 2 = death (competing risk)
etime <- with(data, ifelse(pstat == 0, futime, ptime))
event <- with(data, ifelse(pstat == 0, 2 * death, 1))
event <- factor(event, 0:2)
data$time <- etime
data$status <- event

library(tidycmprsk)
cuminc(Surv(time, status) ~ 1, data = data %>% filter(sex == "M")) %>% 
  ggcuminc(outcome = c("1", "2")) + 
  ylim(c(0, 1)) + 
  labs(
    x = "Days",
    title = "original: male"
  ) + 
  add_confidence_interval() 

cuminc(Surv(time, status) ~ 1, data = data %>% filter(sex == "F")) %>% 
  ggcuminc(outcome = c("1", "2")) + 
  ylim(c(0, 1)) + 
  labs(
    x = "Days",
    title = "target: female"
  ) + 
  add_confidence_interval() 
```

# Centering

## Fitting the centered model

With `survival::finegray` and `survival::coxph`, internal centering is implemented for fine-gray models. **HOWEVER,** the mean values here are of the expanded dataset, `original_fg`, that is created by `survival::finegray`; instead of the original `original` dataset. **Therefore, mannually centering with the appropriate mean values is required.**

```{r}
#| echo: false
#| warning: false
#| message: false

# The original model without centering
options(scipen = 999)

# Prepare dataset and event variable:
# 0 = censor, 1 = pcm (event of interest), 2 = death (competing risk)
original <- original %>% 
  mutate(etime = ifelse(pstat == 0, futime, ptime),
  event = ifelse(pstat == 0, 2 * death, 1)) 
# Fine-Gray model requires a multi-state survival, need to make it a factor
original$event <- factor(original$event)

# Create the Fine-Gray weighted dataset
original_fg <- finegray(Surv(etime, event) ~ ., data = original)

# Fit the weighted Cox model on the Fine-Gray data
fit.o <- coxph(Surv(fgstart, fgstop, fgstatus) ~ hgb + poly(creat, 2, raw = TRUE) + mspike * rcs(age,3), 
  # using the Fine-Gray weighted dataset
  data = original_fg, weight = fgwt,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL)
```

**Note:** For fine-gray models, `center_fit` need to use both `original` and `original_fg`. What the function does is:

1.  Manually center using mean values of `original` instead `original_fg`

2.  Store all model outputs to the returned `centerList`.

3.  Unlike in the linear/logistic regression models, "outcome mean" for survival models (cox, fine-gray) is the average probability of having the event at a specific time (`times`) based on the null model (based on the Kaplan-Meier or Aalen-Johansen estimator). **Note:** I made the decision here to pre-specify `times` instead of outputting the whole `survfit` result, that to make the model output light-weighted.

4.  For fine-gray: the current implementation is using

```{r}
# Time points for calibration and recalibration
times <- c(100, 200, 300)

# The centered model: note here using `original_fg` instead of `original`
centerList <- center_fit(fit.o, data = original, times = times, data_fg = original_fg)

# object of the centered model
fit.c <- centerList$fit.c
```

```{r}
# Manually compute the shift
print("Term means: ")
(means <- fit.o.c$means) # verify all terms have been centered
print("Model coefficients: ")
(coef <- fit.o.c$coefficients)
shift <- coef %*% means

# baseline hazard with centering, i.e., the reference is a subject with all average covariates.
fit.o.c.h0 <- basehaz(fit.o.c) 
# baseline hazard without centering, i.e., the reference is a subject with all 0 covariates.
fit.o.c.h0.nc <- basehaz(fit.o.c, centered = FALSE) 

# Check: exp(shift) should equal to fit.o.c.h0$hazard/fit.o.c.h0.nc$hazard
print(paste("Exp of the shift: ", exp(shift)))
cbind(head(fit.o.c.h0$hazard), head(fit.o.c.h0.nc$hazard), head(fit.o.c.h0$hazard / fit.o.c.h0.nc$hazard))
```

## model outputs: Verify that the two methods produce the same model output

```{r}
# check: coefficients of the three models are identical
cbind(coef(fit.o.c), coef(fit.o), coef(fit.c))
```

```{r}
print("Method 1:")
print("Mean values of all terms")
fit.o.c$means

print("knot locations")
attr(fit.o.c$terms, "predvars")
```

```{r}
print("Method 2:")
print("Mean values of all terms")
centerList$vec_mean

print("knot locations")
centerList$rcsList
```

```{r}
fit.o.c.h0 <- basehaz(fit.o.c) 
fit.c.ho <- basehaz(fit.c) 
print("baseline hazard")
head(cbind(fit.o.c.h0$hazard, fit.c.ho$hazard))
```

**Note:** From the identical term mean values, it is verified that the `coxph` function also follows *transfer then center*.

```{r}
cbind(fit.o.c$means, centerList$vec_mean, fit.o.c$means - centerList$vec_mean)
```

## Transport the centered model to a new dataset

**Note:** For prediction, we don't need to prepare the data using `survival::finegray`, which makes sense since only the predictors are required in `newdata`.

```{r}
timeHorizon <- 300
# -------- Methods 1 -------- 
pred.m1.lp <- predict(fit.o.c, newdata = target, type = "lp")
pred.m1.risk <- predict(fit.o.c, newdata = target, type = "risk")
surv.fit <- survfit(fit.o.c, newdata = target)
pred.m1.survival <- t(summary(surv.fit, times = timeHorizon)$surv)
pred.m1.prob <- 1 - pred.m1.survival # CIF of the main event
head(cbind(pred.m1.lp, pred.m1.risk, pred.m1.survival, pred.m1.prob))

# -------- Methods 2 -------- 
pred.m2.lp <- center_predict(centerList, newdata = target, type = "lp")
pred.m2.risk <- center_predict(centerList, newdata = target, type = "risk")
pred.m2.survival <- center_predict(centerList, newdata = target, type = "survival", times = timeHorizon)
pred.m2.prob <- center_predict(centerList, newdata = target, type = "probability", times = timeHorizon)
head(cbind(pred.m2.lp, pred.m2.risk, pred.m2.survival, pred.m2.prob))
```

# Re-calibration

**Work in Progress** Recalibration of the fine-gray model requires some discussion before the implimentation.

# Calibration plots

```{r}
res.2 <- calibration.cox(
  centerList,
  newdata = target_fg,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)

target_fg$pred.lp <- predict(fit.o.c, newdata = target_fg, type = "lp")
argzConc <- alist(data = target_fg, reverse = TRUE)
argzConc$obj <- as.formula("Surv(fgstart, fgstop, fgstatus) ~ pred.lp") #argzConc$obj = update(fit$formula, "~ - . + LP")
HarrellC <- do.call("concordance", argzConc)
```

# Trick in testing: loading defalut values of a function into the global environment

```{r}
# 1. Safely extract all arguments from the function
vars <- formals(valProbSurvival.fg)

# 2. Drop required arguments (like fit and valdata) that lack a default value
defaults_only <- vars[sapply(vars, function(x) !is.symbol(x) && x != "")]

# 3. Unpack the actual default values into your Global Environment
list2env(defaults_only, envir = .GlobalEnv)
```

```{r}

```