---
title: "Checking custom metric functions of `survival::finegray()` + `survival::coxph()`"
subtitle: "survival::finegray+coxph vs riskRegression::FGR"
author: "DemPoRT V2"
date: last-modified
format:
  html:
    toc: true
---

## Overview

This document compares two approaches for fitting Fine-Gray subdistribution hazard models and computing performance metrics:

1. **Our current approach**: `survival::finegray()` + `survival::coxph()` with custom metric functions
2. **Alternative approach**: `riskRegression::FGR()` + `riskRegression::Score()`

Both fit the same Fine-Gray model but differ in implementation:

- `coxph()` internally centers variables on the **expanded finegray dataset** means, not the original data means. This shifts the baseline hazard and therefore absolute predicted risks.
- `riskRegression::Score()` provides published IPCW Brier and AUC implementations for competing risks; our pipeline uses custom implementations.

```{r}
library(survival)
library(riskRegression)
library(prodlim)
library(dplyr)

source("R/calculate-performance-metrics.R")
```

## Comparison

```{r}
data <- mgus2 # from survival
# remove rows with missing
data <- data[complete.cases(data), ]

# Prepare dataset and event variable:
# 0 = censor, 1 = pcm (event of interest), 2 = death (competing risk)
# To work with `compute_c_statistic_fg` and `compute_brier_fg`, 
# we must name the variables as `time_to_event` and `event_type`,
# this limits the general use of these two functions outside DemPoRT
data$time_to_event <- ifelse(data$pstat == 0, data$futime, data$ptime)
data$event_type <- ifelse(data$pstat == 0, 2 * data$death, 1)

# The time point of interest
timeHorizon <- 300
```

### Some initial checking on the finegray + coxph model

1. `coxph()` internally centers variables on the **expanded finegray dataset** means, not the original data means.

2. Two versions of `compute_fg_pred_risk` produce the almost identical predicted risk.

```{r}
# --------- The finegray + coxph model -----------
# Create the Fine-Gray weighted dataset
data_fg <- finegray(Surv(time_to_event, factor(event_type)) ~ ., data = data, etype = 1)

# Fit the weighted Cox model on the Fine-Gray data
fit_ours <- coxph(
  Surv(fgstart, fgstop, fgstatus) ~ hgb + creat + mspike + age, 
  # using the Fine-Gray weighted dataset
  data = data_fg, weight = fgwt,
  # also center on dummy variables
  nocenter = NULL,
  # MUST: for robust standard errors
  id = id, robust = TRUE,
  # CRITICAL FOR MATCHING COEFFICIENTS with riskRegression::FGR, 
  # still not identical (coefficients agree to ~4 decimal places)
  ties = "breslow")

# IMPORTANT: the mean values in `fit_ours` are of `data_fg` not `data`!!!
print("Mean values in `fit_ours`: ")
fit_ours$means
print("Corresponding mean values in `data_fg`: ")
X_fg <- data_fg %>% select(hgb, creat, mspike, age)
colMeans(X_fg)
print("Corresponding mean values in `data`: ")
X <- data %>% select(hgb, creat, mspike, age)
colMeans(X)

# ------------ Versions of `compute_fg_pred_risk` ---------
pred_risk_ours_old <- compute_fg_pred_risk_old(fit_ours, data, timeHorizon)
pred_risk_ours <- compute_fg_pred_risk(fit_ours, data, timeHorizon)
print("Versions of `compute_fg_pred_risk`: essentially the same.")
head(cbind(pred_risk_ours_old, pred_risk_ours))
summary(pred_risk_ours_old - pred_risk_ours)
```

### Coefficients of the finegray + coxph model vs the FGR model

The two models can produce slightly different coefficients because of the different methods used on how to tie tied event times and construct the status variable, even with `ties = "breslow"` in `coxph`.

```{r}
# --------- The FGR model -----------
fit_fgr <- FGR(
  # Using `Hist` instead of `Surv`, only includes linear terms now
  formula = Hist(time_to_event, event_type) ~ hgb + creat + mspike + age, 
  data = data, 
  # specify the event of interest
  cause = 1
)

# ------------ Compare coefficients ---------
print("Coefficients: similar but not identical")
cbind(fit_ours$coefficients, fit_fgr$crrFit$coef)
```

```{r}
# ------------ Compare predicted risk ---------
print("Predicted risk: similar but not identical")
# `survival`
pred_risk_ours <- compute_fg_pred_risk(fit_ours, data, timeHorizon)
# `riskRegression`
pred_risk_fgr <- predictRisk(
  object = fit_fgr,
  newdata = data,
  times = timeHorizon,
  cause = 1
)
print(head(cbind(pred_risk_ours, pred_risk_fgr)))
summary(pred_risk_ours - pred_risk_fgr)
```

```{r}
# ------------ Compare predicted linear predictor ---------
print("Predicted lp: difference due to centering")
# `survival`
pred_lp_ours <- predict(fit_ours, data, type = "lp")
# `riskRegression`
X <- as.matrix(data %>% select(hgb, creat, mspike, age))
coef <- fit_fgr$crrFit$coef
pred_lp_fgr <- as.vector(X %*% coef)
print(head(cbind(pred_lp_ours, pred_lp_fgr, pred_lp_ours - pred_lp_fgr)))
# The difference is due to centering
X_fg <- as.matrix(data_fg %>% select(hgb, creat, mspike, age))
print(paste0(
  "Sum of variable means in `data_fg` times coefficients: ",
  colMeans(X_fg) %*% coef
))
```

```{r}
# ------------ Compare metrics ---------
# C-statistic (ours)
c_ours <- compute_c_statistic_fg(fit_ours, data)
# Brier (ours)
brier_ours <- compute_brier_fg(pred_risk_ours, data, timeHorizon)

# Score (FGR)
eval_fgr <- Score(
  object  = list("FGR" = fit_fgr),
  formula = Hist(time_to_event, event_type) ~ 1,
  data    = data,
  metrics = c("auc", "brier"),
  cause   = "1",
  times   = timeHorizon,
  se.fit  = FALSE
)
fgr_auc   <- as.numeric(eval_fgr$AUC$score[model == "FGR", AUC])
fgr_brier <- as.numeric(eval_fgr$Brier$score[model == "FGR", Brier])
fgr_brier_null <- as.numeric(eval_fgr$Brier$score[model == "Null model", Brier])
fgr_brier_scaled <- 1 - fgr_brier / fgr_brier_null
print("AUC: difference requires investigation")
print(c(c_ours$c_stat, fgr_auc))
print("Brier: slightly different")
print(c(brier_ours$brier, fgr_brier))
print(c(brier_ours$brier_scaled, fgr_brier_scaled))
```

```{r run-comparison}
results <- list()

for (sex in c("male", "female")) {
  model_data <- tar_read_raw(paste0(sex, "_model_data"))
  model_data$event_type <- factor(model_data$event_type)
  fg_model <- tar_read_raw(paste0(sex, "_fine_and_gray_model"))
  predictors <- names(coef(fg_model))

  # Fit FGR
  fit_fgr <- FGR(
    as.formula(paste0("Hist(time_to_event, event_type) ~ ",
                       paste(predictors, collapse = " + "))),
    data = model_data, cause = "1"
  )

  # Coefficients
  coef_ours <- coef(fg_model)
  coef_fgr  <- fit_fgr$crrFit$coef
  shared    <- intersect(names(coef_ours), names(coef_fgr))
  max_coef_diff <- max(abs(as.numeric(coef_ours[shared] - coef_fgr[shared])))

  # Predicted risk
  pred_risk_ours <- compute_fg_pred_risk(fg_model, model_data, horizon)
  pred_risk_fgr  <- as.numeric(
    predictRisk(fit_fgr, newdata = model_data, times = horizon, cause = "1")
  )

  # C-statistic (ours)
  c_ours <- compute_c_statistic_fg(fg_model, model_data)

  # Score (FGR)
  eval_fgr <- Score(
    object  = list("FGR" = fit_fgr),
    formula = Hist(time_to_event, event_type) ~ 1,
    data    = model_data,
    metrics = c("auc", "brier"),
    cause   = "1",
    times   = horizon,
    se.fit  = FALSE
  )
  fgr_auc   <- as.numeric(eval_fgr$AUC$score[model == "FGR", AUC])
  fgr_brier <- as.numeric(eval_fgr$Brier$score[model == "FGR", Brier])
  fgr_brier_null <- as.numeric(eval_fgr$Brier$score[model == "Null model", Brier])
  fgr_brier_scaled <- 1 - fgr_brier / fgr_brier_null

  # Brier (ours)
  brier_ours <- compute_brier_fg(pred_risk_ours, model_data, horizon)

  # Calibration slope (ours)
  cal_ours <- compute_calibration_fg(pred_risk_ours, model_data, horizon,
                                     model_for_slope = fg_model)

  # Calibration slope (FGR) — LP computed manually (crr has no predict(type="lp"))
  lp_fgr <- as.numeric(
    as.matrix(model_data[, names(coef_fgr), drop = FALSE]) %*% coef_fgr
  )
  cal_data <- model_data
  cal_data$lp_fgr <- lp_fgr
  fg_cal_data <- survival::finegray(
    survival::Surv(time_to_event, factor(event_type)) ~ lp_fgr,
    data = cal_data, etype = "1"
  )
  slope_model_fgr <- survival::coxph(
    survival::Surv(fgstart, fgstop, fgstatus) ~ lp_fgr,
    data = fg_cal_data, weights = fg_cal_data$fgwt
  )
  cal_slope_fgr <- as.numeric(coef(slope_model_fgr)[["lp_fgr"]])

  results[[sex]] <- list(
    c_ours = c_ours$c_stat, c_fgr = fgr_auc,
    brier_ours = brier_ours$brier, brier_fgr = fgr_brier,
    brier_scaled_ours = brier_ours$brier_scaled, brier_scaled_fgr = fgr_brier_scaled,
    cal_slope_ours = cal_ours$slope, cal_slope_fgr = cal_slope_fgr,
    pred_mean_ours = mean(pred_risk_ours), pred_mean_fgr = mean(pred_risk_fgr),
    pred_corr = cor(pred_risk_ours, pred_risk_fgr),
    pred_mad = mean(abs(pred_risk_ours - pred_risk_fgr)),
    max_coef_diff = max_coef_diff,
    mean_coef_diff = mean(abs(as.numeric(coef_ours[shared] - coef_fgr[shared])))
  )

  cat(toupper(sex), "done.\n")
}
```

## Final Summary

```{r final-summary}
m <- results$male
f <- results$female

metrics <- data.frame(
  Metric = c("C-statistic", "Brier score", "Brier scaled",
             "Calibration slope", "Predicted risk (mean)"),
  Male_Ours = c(m$c_ours, m$brier_ours, m$brier_scaled_ours,
                m$cal_slope_ours, m$pred_mean_ours),
  Male_FGR = c(m$c_fgr, m$brier_fgr, m$brier_scaled_fgr,
               m$cal_slope_fgr, m$pred_mean_fgr),
  Female_Ours = c(f$c_ours, f$brier_ours, f$brier_scaled_ours,
                  f$cal_slope_ours, f$pred_mean_ours),
  Female_FGR = c(f$c_fgr, f$brier_fgr, f$brier_scaled_fgr,
                 f$cal_slope_fgr, f$pred_mean_fgr)
)

# Round numeric columns
metrics[, 2:5] <- round(metrics[, 2:5], 6)

knitr::kable(metrics, caption = "Performance metrics: finegray+coxph vs FGR")

agreement <- data.frame(
  Metric = c("Pred. risk correlation", "Pred. risk mean abs diff",
             "Mean coef abs diff", "Max coef abs diff"),
  Male = c(round(m$pred_corr, 6), round(m$pred_mad, 6),
           round(m$mean_coef_diff, 6), round(m$max_coef_diff, 6)),
  Female = c(round(f$pred_corr, 6), round(f$pred_mad, 6),
             round(f$mean_coef_diff, 6), round(f$max_coef_diff, 6))
)
knitr::kable(agreement, caption = "Agreement between approaches")
```

## Interpretation

| Metric | Expected | Why |
|--------|----------|-----|
| Coefficients | Identical | Same underlying Fine-Gray model |
| Predicted risk | Shifted | `coxph` centers on expanded finegray data means, not original data |
| C-statistic | Match | Rank-based, unaffected by calibration shift |
| Brier score | Close | Depends on predicted risk shift + IPCW weighting method |
| Brier scaled | Close | Derived from Brier, inherits any differences |
| Calibration slope | Close | Both use LP-based secondary Fine-Gray model |
