---
title: "Examples of the proper center and recalibrate pipeline: coxph"
author: "Juan Li"
date: 2026-06-16
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(CalibrationCurves)
library(survival)
library(ggsurvfit)
library(gtsummary)

# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_predict.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
```

# Breast Cancer Survival Data from Rotterdam and Germany

See `?CalibrationCurves::trainDataSurvival`.

```{r}
# variables for `time` and `status` are `ryear` and `rfs`, respectively
data(trainDataSurvival) # Rotterdam
data(testDataSurvival) # German Breast Cancer Study Group
```

Kaplan-Meier plots: https://www.emilyzabor.com/survival-analysis-in-r.html

```{r}
data <- bind_rows(
  trainDataSurvival %>% select(ryear, rfs) %>% mutate(type = "train"),
  testDataSurvival %>% select(ryear, rfs) %>% mutate(type = "test")
)
data$type <- factor(data$type, levels = c("train", "test"))

survfit2(Surv(ryear, rfs) ~ type, data = data) %>% 
  ggsurvfit() +
  add_confidence_interval() +
  add_risktable()
```

## Variables

### in both datasets

`pid`: patient identifier

`age`: age at surgery

`meno`: menopausal status (0 = premenopausal, 1 = postmenopausal)

`size`: tumor size, a factor with levels \<= 20, 20-50, \>50

- `csize`: Categorized tumor size, copied from size (factor)

`grade`: differentiation grade

- `grade3`: Recoded grade factor (levels: "1-2", "3")

`nodes`: number of positive lymph nodes

- `nodes2`: Winsorized node count (numeric)

- `nodes3`: Restricted cubic spline basis for nodes2 (numeric)

- `cnode`: Categorized node involvement (factor: "0", "1-3", "\>3")

`pgr`: progesterone receptors (fmol/l)

- `pgr2`: Winsorized progesterone receptor level (numeric)

- `pgr3`: Restricted cubic spline basis for original pgr (numeric)

`er`: estrogen receptors (fmol/l)

`hormon`: hormonal treatment (0 = no, 1 = yes)

`ryear`: Follow-up time for RFS, in years (numeric)

`rfs`: Recurrence-free survival status (0 = no event, 1 = event) (numeric)

`epoch`: Follow-up epoch indicator after splitting at 5 years (numeric)

### only in `trainDataSurvival`

`year`: year of surgery

`chemo`: chemotherapy

`rtime`: days to relapse or last follow-up

`recur`: 0 = no relapse, 1 = relapse

`dtime`: days to death or last follow-up

`death`: 0 = alive, 1 = dead

`tstart`

### only in `testDataSurvival`

`rfstime`, `status`

```{r}
original <- trainDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
target <- testDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
summary(original)
summary(target)
```

# Centering

## Fitting the centered model using `survival::coxph()`, with `nocenter = NULL`

This way, `survival::coxph()` does its internal centering, including the dummy variables.

```{r}
# The original model without centering
options(scipen = 999)

fit.o.c <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + csize * rcs(pgr2,3), data = original,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL) 
```

## Transport the centered model to a new dataset

**Note:** Here, `center_predict` is a wrapper function of the corresponding code in Method 1, the main reason to have `center_predict` is to automatically incorporate `delta_recal`.

```{r}
pred.m1.lp <- predict(fit.o.c, newdata = target, type = "lp")
pred.m1.risk <- predict(fit.o.c, newdata = target, type = "risk") # exp(lp)
surv.fit <- survfit(fit.o.c, newdata = target)
pred.m1.survival <- t(summary(surv.fit, times = 3)$surv) # probability of survival at a specific time
pred.m1.prob <- 1 - pred.m1.survival # probability of event at a specific time
head(cbind(pred.m1.lp, pred.m1.risk, pred.m1.survival, pred.m1.prob))
```

## Recalibrate

**Note:**

1.  The `coxph` object cannot be directly updated to reflect the recalibration changes. The current work-around is to do it manually in `predict_centre`.

2.  Therefore, here `center_recalibrate` is mainly used to calculate the model shift, *i.e.* `delta_recal`.

3.  For re-calibration, we can only work with one single `timeHorizon` at a time.

4.  The specific `timeHorizon` should be within the pre-specify `times` when running `center_fit`.

5.  The method in doi:10.1016/j.jclinepi.2025.111895 is as following:

- suppose the risk/probability of event at a specific time is $\theta = 1 - S_{KM}(t)$, the rescaling factor they proposed was $\Delta_{rescaling} = ln[-ln(1-\theta_{derived})] - ln[-ln(1-\theta_{target})]$.

- Then, the recalibrated risk is $\theta_{recal} = 1 - exp(-exp(ln[-ln(1-\theta_{ori})]-\Delta_{rescaling}))$.

- Let's substitute $S_{recal} = 1 - \theta_{recal}$ and $S_{ori} = 1 - \theta_{ori}$ into the equation, then $1 - S_{recal} = 1 - exp(-exp(ln[-ln(S_{ori})]-\Delta_{rescaling}))$, *i.e.*, $S_{recal} = exp(-exp(ln[-ln(S_{ori})]-\Delta_{rescaling}))$

$= e^{-e^{ln[-ln(S_{ori})]-\Delta_{rescaling}}}$

$= e^{-e^{ln[-ln(S_{ori})]}e^{-\Delta_{rescaling}}}$

$= [e^{-e^{ln[-ln(S_{ori})]}}]^{e^{-\Delta_{rescaling}}}$

$= [e^{ln(S_{ori})}]^{e^{-\Delta_{rescaling}}}$

$= [S_{ori}]^{e^{-\Delta_{rescaling}}}$

- In the Cox model, the equation for hazard is $h(t|X_i) = h_0(t)exp(\beta^TX_i)$. And the survival function is $S(t|X_i) = [S_0(t)]^{exp(\beta^TX_i)}$. Therefore, the equation above can be written as $S_{recal} = [S_{ori}]^{e^{-\Delta_{rescaling}}}$

$= ([S_{0,ori}(t)]^{e^{\beta^TX_i}})^{e^{-\Delta_{rescaling}}}$

$= S_{0,ori}(t)^{e^{\beta^TX_i}e^{-\Delta_{rescaling}}}$

$= S_{0,ori}(t)^{e^{(\beta^TX_i -\Delta_{rescaling})}}$

- Note, in my implementation, I make `delta_recal` = $-\Delta_{rescaling} = ln[-ln(1-\theta_{target})] - ln[-ln(1-\theta_{derived})]$ (*i.e.* `outcome_mean_target_clog_log - outcome_mean_original_clog_log`).

```{r}
# Time point for calibration and recalibration
timeHorizon <- 3

# Probability of event at the specific `timeHorizon`
# In original data
surv_fit <- survfit(Surv(ryear, rfs) ~ 1, data = original)
prob.original <- 1 - t(summary(surv_fit, times = timeHorizon)$surv)[1]
# In target data
surv_fit <- survfit(Surv(ryear, rfs) ~ 1, data = target)
prob.target <- 1 - t(summary(surv_fit, times = timeHorizon)$surv)[1]

# The rescaling factor
prob.original.cloglog <- log(-log(1 - prob.original))
prob.target.cloglog <- log(-log(1 - prob.target))
delta_recal <- prob.target.cloglog - prob.original.cloglog
```

# Calibration plots

```{r}
source("R/valProbSurvival.2.R", echo = FALSE) # Modified function for cox models
```

```{r}
# The original model
pred.lp <- predict(fit.o.c, target, type = "lp")
surv.fit <- survfit(fit.o.c, newdata = target)
pred.surv <- t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob <- 1 - pred.surv
pred.surv.brier <- t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
pred.prob.brier <- 1 - pred.surv.brier
outcome.time <- target$ryear
outcome.event <- target$rfs

res.1 <- valProbSurvival.2(
  pred.lp,
  pred.prob,
  pred.prob.brier,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
# The recalibrated model
pred.lp.recal <- pred.lp + delta_recal
pred.prob.recal <- 1 - pred.surv ^ exp(delta_recal)
pred.prob.brier.recal <- 1 - pred.surv.brier ^ exp(delta_recal)

res.2 <- valProbSurvival.2(
  pred.lp.recal,
  pred.prob.recal,
  pred.prob.brier.recal,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

**Note:** the `offset(lp)` method doesn't work for `coxph`.

```{r}
# update intercept using Platt's calibration, which requires linked individual-level newdata
# Use `valProbSurvival.2` directly here

target$LP <- pred.lp
fit.platt <- coxph(Surv(ryear, rfs) ~ LP, data = target,
model = FALSE, x = FALSE, y = FALSE)

pred.lp.platt <- predict(fit.platt, target, type = "lp")
surv.fit <- survfit(fit.platt, newdata = target)
pred.prob.platt <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier.platt <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)

res.3 <- valProbSurvival.2(
  pred.lp.platt,
  pred.prob.platt,
  pred.prob.brier.platt,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5
)
```

```{r}
res.1$stats$Calibration$Statistics
res.2$stats$Calibration$Statistics
res.3$stats$Calibration$Statistics
```

```{r}

```