---
title: "Simulation studies - quantiles and non-linear terms"
author: "Juan Li"
date: 2026-07-29
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(dplyr) # Data manipulation
library(tidyr) # Tidy messy data
library(ggplot2) # Data visualization
library(ggpubr) # ggarrange
library(gtsummary)
library(rms)
library(foreach)
library(doParallel)
```

# Simulation study 1 - quantiles and non-linear terms

In this simulation study, we will explore the extimated mean values of variour non-linear terms, using different numbers of quantiles.

**Things to control:**

- Data distribution (N = 10000): uniform, normal, skewed, double-peak

- Number of quantiles: nbreak = c(3, 5, 10, 20, 50, 100). Quantiles including `min` and `max`, therefore `nbreak = 3` means `min`, `median`, and `max`.

- Non-linear term:

  - polynomial: x\^2, x\^3

  - rcs: k = 4; same distribution, different distribution

**The steps are:**

1.  Genarate data with different distributions. - **real data**

2.  For each data, generate its quantiles using different nbreak.

3.  Generate data from each quantile (500 samples): both assume uniform distribution within each bin - **simulated data**

4.  Get the mean values of each term from the **simulated data** and compare with the ones from the **real data**.

```{r}
# ------------------- Various non-linear terms and their mean values -------------------
non.linear.mean <- function(df, knots.diff, k = 4) {
  df$x2 <- df$x^2
  df$x3 <- df$x^3
  # rcs: same distribution
  rcs.same <- rcs(df$x, k)
  knots.same <- attributes(rcs.same)$parms
  rcs.res.same <- get_rcs(df$x, knots.same)
  df$rcs.2.same <- rcs.res.same[, 2]
  df$rcs.3.same <- rcs.res.same[, 3]
  # rcs: different distribution
  rcs.res.diff <- get_rcs(df$x, knots.diff)
  df$rcs.2.diff <- rcs.res.diff[, 2]
  df$rcs.3.diff <- rcs.res.diff[, 3]

  vec_mean <- colMeans(df)
  return(vec_mean = vec_mean)
}

# ------------------- Extract rcs terms -------------------
get_rcs <- function(vec, knots = NULL, k = 4) {
  if (is.null(knots)) {
    rcs.fit <- rcs(vec, k)
  } else {
    rcs.fit <- rcs(vec, knots)
  }
  # extract rcs terms
  res <- labelled::remove_attributes(
    rcs.fit,
    names(attributes(rcs.fit))[3:length(names(attributes(rcs.fit)))]
  ) # don't print the attributes
  return(res)
}

# ------------------- Quantile and simulate data from it -------------------
quant.simu <- function(df, nbreak, vec_seed, knots.diff, nsample) {
  # number of rcs knots
  k <- length(knots.diff)

  # quantile
  quantiles <- seq(0, 100, length.out = nbreak) # including min (0%) and max (100%)
  quant_vec <- quantile(df$x, probs = quantiles/100)
  quantile_counts <- table(cut(df$x, breaks = quant_vec, include.lowest = TRUE))
  data.quant <- data.frame(
    quantiles = quantiles,
    quant_vec = unname(quant_vec)
  )

  # bins
  nbin <- nbreak - 1
  data.bin <- data.frame(
    count = as.numeric(unname(quantile_counts)),
    lower = data.quant$quant_vec[1:nbin],
    upper = data.quant$quant_vec[2:nbreak]
  )

  # simulate data from the quantile, get non-linear terms and their mean values
  # Parallel
  # Determine the number of cores to use (it's best to leave one free)
  n_cores <- parallel::detectCores() - 1

  # Create and register the cluster
  my_cluster <- makeCluster(n_cores)
  registerDoParallel(my_cluster)

  df_mean <- foreach(
    i = 1:nsample,
    .packages = c("ggplot2", "dplyr", "rms"),
    .export = c(
      "non.linear.mean",
      "get_rcs"
    ),
    .combine = "rbind"
  ) %dopar%
    {
      set.seed(vec_seed[i])
      vec <- c()
      for (j in 1:nbin) {
        temp <- runif(
          n = data.bin$count[j],
          min = data.bin$lower[j],
          max = data.bin$upper[j]
        )
        vec <- c(vec, temp)
      }
      data_i <- data.frame(x = vec)
      vec_mean <- non.linear.mean(data_i, knots.diff, k)
      return(vec_mean)
    }

  # Stop the cluster
  stopCluster(my_cluster)

  vec_025 <- apply(df_mean, 2, quantile, probs = 0.025, na.rm = TRUE)
  vec_50 <- apply(df_mean, 2, quantile, probs = 0.5, na.rm = TRUE)
  vec_975 <- apply(df_mean, 2, quantile, probs = 0.975, na.rm = TRUE)
  res_mean <- data.frame(
    term = names(vec_50),
    nbreak = rep(nbreak, ncol(df_mean)),
    mean = unname(vec_50),
    mean.lower = unname(vec_025),
    mean.upper = unname(vec_975)
  )

  return(list(data.quant = data.quant, data.bin = data.bin, res_mean = res_mean))
}
```

```{r}
# -------------------  set up the simulation -------------------
vec_N <- c(500) #c(500, 1000, 5000, 10000) # vector of N
vec_type <- c("unif") #c("unif", "norm", "skew", "double") # vector of data type
vec_nbreak <- c(3) #c(3, seq(5, 100, 5)) # vector of nbin
vec_nbreak_plot <- c(3) #c(3, 5, 10, 20, 50, 100)
# vector of random seeds
set.seed(123)
nsample <- 5 # 500
vec_seed <- sample.int(1e6, nsample)

# ------------------- rcs knots from a different distribution -------------------
k <- 4 # number of knots
dist.diff <- rbeta(n = 10000, shape1 = 2, shape2 = 10)
rcs.diff <- rcs(dist.diff, k)
knots.diff <- attributes(rcs.diff)$parms

# ------------------- Generate data -------------------
res_quant <- data.frame()
res_real <- data.frame()
for (N in vec_N) {
  print(N)
  file_name_quant <- paste0("Paper_qmd/Simulation_1_results/mean_quant_", N, ".csv")
  file_name_real <- paste0("Paper_qmd/Simulation_1_results/mean_real_", N, ".csv")

  if (file.exists(file_name_quant)) {
    mean_quant <- read.csv(file_name_quant)
    mean_real_long <- read.csv(file_name_real)
  } else {
    set.seed(123)
    unif <- runif(n = N, min = 0, max = 1)
    norm <- rnorm(n = N, mean = 5, sd = 2)
    skew <- rbeta(n = N, shape1 = 10, shape2 = 2)
    double <- c(
      rnorm(n = N / 2, mean = 5, sd = 2),
      rnorm(n = N / 2, mean = 12, sd = 2)
    )
    data <- data.frame(
      type = rep(vec_type, each = N),
      x = c(unif, norm, skew, double)
    )
    data$type <- factor(data$type, levels = vec_type)

    # ------------------- Non-linear terms of real data -------------------
    mean_real <- data.frame()
    for (i_type in vec_type) {
      data_i <- data %>%
        filter(type == i_type) %>%
        droplevels() %>%
        select(-type)
      vec_mean <- non.linear.mean(data_i, knots.diff)
      mean_real <- bind_rows(mean_real, c("type" = i_type, vec_mean))
    }
    mean_real$type <- factor(mean_real$type, levels = vec_type)

    mean_real_long <- mean_real %>%
      pivot_longer(-type, names_to = "term", values_to = "mean")
    mean_real_long$type <- factor(mean_real_long$type, levels = vec_type)
    mean_real_long$term <- factor(
      mean_real_long$term,
      levels = names(mean_real)[-1]
    )
    mean_real_long$mean <- as.numeric(mean_real_long$mean)
    mean_real_long$N <- rep(N, nrow(mean_real_long))
    write.csv(mean_real_long, row.names = FALSE, file = file_name_real)

    # ------------------- Generate data from quantiles -------------------
    mean_quant <- data.frame()
    for (i_type in vec_type) {
      data_i <- data %>%
        filter(type == i_type) %>%
        droplevels() %>%
        select(-type)
      for (nbreak in vec_nbreak) {
        temp <- quant.simu(data_i, nbreak, vec_seed, knots.diff, nsample)$res_mean
        temp$type <- rep(i_type, nrow(temp))
        mean_quant <- bind_rows(mean_quant, temp)
      }
    }
    mean_quant$N <- rep(N, nrow(mean_quant))
    mean_quant$type <- factor(mean_quant$type, levels = vec_type)
    mean_quant$term <- factor(mean_quant$term, levels = names(mean_real)[-1])
    write.csv(mean_quant, row.names = FALSE, file = file_name_quant)

    # ------------------- Plot -------------------
    pList <- list()
    for (i_type in vec_type) {
      data_i <- data %>% filter(type == i_type) %>% droplevels()
      for (nbreak in vec_nbreak_plot) {
        p <- ggplot(data_i, aes(x)) +
          geom_quantogram(
            aes(y = ..density..),
            bins = nbin,
            alpha = 0.2,
            color = "black"
          ) +
          geom_density(color = "red") +
          theme_bw() +
          labs(title = paste0("N=", N, ", ", nbin, " bins"))
        pList[[length(pList) + 1]] <- p
      }
    }
    print(ggarrange(plotlist = pList, ncol = 6, nrow = 4))
  }

  res_real <- bind_rows(res_real, mean_real_long)
  res_quant <- bind_rows(res_quant, mean_quant)
}
```

```{r}
vec_term <- c()
res_quant$term <- factor(res_quant$term, levels = unique(res_quant$term))
res_real$term <- factor(res_real$term, levels = unique(res_real$term))

# Plot
levels(res_quant$term) <- c(
  "x",
  "x^2",
  "x^3",
  "rcs.2.same",
  "rcs.3.same",
  "rcs.2.diff",
  "rcs.3.diff"
)
levels(res_real$term) <- c(
  "x",
  "x^2",
  "x^3",
  "rcs.2.same",
  "rcs.3.same",
  "rcs.2.diff",
  "rcs.3.diff"
)

for (i_type in vec_type) {
  p <- ggplot(
    res_quant %>% filter(type == i_type, term != "x") %>% droplevels(),
    aes(x = nbin, y = mean, color = factor(N))
  ) +
    geom_point() +
    geom_linerange(aes(ymin = mean.lower, ymax = mean.upper)) +
    geom_line(alpha = 0.5) +
    geom_hline(
      data = res_real %>% filter(type == i_type, term != "x") %>% droplevels(),
      aes(yintercept = mean, color = factor(N)),
      linetype = "dashed"
    ) +
    facet_wrap(~term, scales = "free", nrow = 3) +
    scale_x_continuous(breaks = vec_nbreak) +
    theme_bw() +
    labs(title = i_type, color = "N") +
    theme(text = element_text(size = 12))
  print(p)
}
```

```{r}

```