---
title: "Simulation studies - Recalibration"
author: "Juan Li"
date: 2026-04-14
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(dplyr) # Data manipulation
library(tidyr) # Tidy messy data
library(ggplot2) # Data visualization
library(ggpubr) # ggarrange
library(gtsummary)
library(rms)
library(fGarch) # for skewed distribution with target mean and sd
library(foreach)
library(doParallel)

source("R/centering.R")
# functions in "centering.R"
# get_rcs: implement the formula of rcs components
# step_dummy: dummy variables
# step_rcs: rcs
# step_interaction: interaction terms
# step_center: centering
# get_mean: get mean values from the original dataset
# root.search: Root searching, not in use in this .qmd file

source("R/calibration.R") # for calibration plots
```

# Simulation study 2 - Recalibration of logistic regression

## Data generation

In this simulation study, we will generate data as follows:

1.  We simulated derivation and validation samples of size N = 10000

2.  For each subject in the *derivation sample*, we simulated a continuous covariate from a standard normal distribution: x \~ N(target_mu, target_sd). Default: `target_mu` = 0, `target_sd` = 1.

3.  For each subject in the *validation sample*, we simulated x using a skew-normal distribution with the shifted mean (`target_mu` + `mu_shift`) and same `target_sd` as in the derivation sample. Default: `mu_shift` = 0.5, `xi` = -2. (`xi` is the shape/skewness parameter of the skew-normal distribution)

4.  For each subject in the derivation sample, a linear predictor was defined as LP = x + alpha \* x\^2 + error, where error \~ N(0, error_sd). Default: `alpha` = 0.5, `error_sd` = 0.5.

5.  For each subject in the validation sample, one linear predictor was defined: LP = drift + slope \* (x + alpha \* x\^2) + error, where error \~ N(0, error_sd).

- When `drift` = 0, and `slope` = 1, the `LP` is the same as in the \_derivation sample: LP = x + alpha \* x\^2 + error.

- When `drift` != 0, and `slope` = 1, the `LP` is with a different intercept: LP = drift + x + alpha \* x\^2 + error, where error \~ N(0, error_sd).

- When `drift` != 0, and `slope` != 1, the `LP` is with different intercept and different slope.

6.  For each linear predictor in both samples, a binary outcome variable y was simulated from a Bernoulli distribution: Y = Be(p), where p = exp(LP)/(1+exp(LP)).

```{r}
data.lr <- function(
  N.dev = 10000,
  N.ext = 10000,
  target_mu = 0,
  target_sd = 1,
  mu_shift = 0.5,
  xi = -2,
  alpha = 0.5,
  error_sd = 0.5,
  drift = 0,
  slope = 1,
  seed = 123
) {
  # N.dev     ...Sample size of the development dataset
  # N.ext     ...Sample size of the validation/external dataset
  # target_mu ...Target mean of the two distributions
  # target_sd ...Target standard deviation of the two distributions
  # mu_shift  ...Mean shift in the validation dataset
  # xi        ...Shape/skewness parameter of the skew-normal distribution.
  #               If xi>0, right-skewed, positive skewness. If xi<0, left-skewed, negative skewness.
  # alpha     ...Coefficient of the quadratic term in the linear predictor
  # error_sd  ...Standard deviation of the error term
  # drift     ...Intercept drift in the validation dataset
  # slope     ...Slope change in the validation dataset
  # seed      ...Random seed for reproducibility

  set.seed(seed)

  # derivation dataset
  data.dev <- data.frame(
    x = rnorm(n = N.dev, mean = target_mu, sd = target_sd),
    error = rnorm(n = N.dev, mean = 0, sd = error_sd)
  ) %>%
    mutate(x2 = x^2) %>%
    mutate(lp = x + alpha * x2 + error) %>% # the linear predictor
    mutate(p = exp(lp) / (1 + exp(lp))) # inverse-logit function, can also use plogis(lp)
  data.dev$y <- rbinom(n = N.dev, size = 1, prob = data.dev$p)

  # validation/external dataset
  data.ext <- data.frame(
    x = rsnorm(n = N.ext, mean = target_mu + mu_shift, sd = target_sd, xi = xi),
    error = rnorm(n = N.ext, mean = 0, sd = error_sd)
  ) %>%
    mutate(x2 = x^2) %>%
    mutate(lp = drift + slope * (x + alpha * x2) + error) %>% # the linear predictor
    mutate(p = exp(lp) / (1 + exp(lp))) # inverse-logit function, can also use plogis(lp)
  data.ext$y <- rbinom(n = N.ext, size = 1, prob = data.ext$p)

  return(list(data.dev = data.dev, data.ext = data.ext))
}
```

## Fit and center models

1.  Two logistic regression models were then fit to the simulated derivation sample: (i) a correctly specified regression model; and (ii) an incorrectly specified regression model that omit the quadratic term (x\^2).

2.  The two models were transferred into centered models using the specific steps.

```{r}
fit.center.lr <- function(data.dev, model = "correct") {
  # -------- The original models --------
  if (model == "correct") {
    fit.o <- glm(y ~ x + x2, data = data.dev, family = "binomial")
  } else {
    fit.o <- glm(y ~ x, data = data.dev, family = "binomial")
  }

  # -------- centering: data.dev --------
  vars_mean <- c("x", "x2")
  # get mean values
  means.dev <- get_mean(data.dev, vars_mean)
  # centering vars in the data.dev dataset
  vars_center <- names(means.dev)
  data.dev <- step_center(data.dev, vars_center, means.dev)

  # -------- The centered models --------
  if (model == "correct") {
    fit.c <- glm(y ~ x_C + x2_C, data = data.dev, family = "binomial")
  } else {
    fit.c <- glm(y ~ x_C, data = data.dev, family = "binomial")
  }

  # outcome mean
  y_mean_dev <- mean(data.dev$y)
  y_mean_logit_dev <- log(y_mean_dev/(1-y_mean_dev))
  # global offset
  offset <- coef(fit.c)[1] - y_mean_logit_dev

  return(list(
    data.dev = data.dev,
    means.dev = means.dev,
    fit.o = fit.o,
    fit.c = fit.c,
    y_mean_dev = y_mean_dev,
    y_mean_logit_dev = y_mean_logit_dev,
    offset = offset
  ))
}
```

## Transport model to the validation sample and influence of intercept shift on calibration

1.  The two centered models were applied to the validation sample and recalibrated with various shifts of the intercept. Two shifts of interest:

- For linked data, update the intercept using a logistic regression with fixed slope and different intercept (**optimal shift**)

- For unlinked data, update the intercept using `Logit of outcome mean in the validation sample` + `global offset`, *i.e.,* `shift` = `Logit of outcome mean in the validation sample` - `Logit of outcome mean in the derivation sample`. (**recalibration shift**)

2.  Update the model with an arbitrary intercept shift and get the corresponding calibration plot and metrics.

3.  Eavg (ICI), E90, and Emax were used as the main evaluation metrics.

```{r}
recal.lr <- function(
  data.ext,
  means.dev,
  fit,
  offset,
  vec_shift,
  plot = FALSE
) {
  # -------- centering: data.ext --------
  # centering vars in the data.ext dataset using mean values in the `data.dev` dataset
  vars_center <- names(means.dev)
  data.ext <- step_center(data.ext, vars_center, means.dev)

  #  -------- prediction --------
  data.ext$pred.p <- predict(fit, newdata = data.ext, type = "response")
  data.ext$pred.logit <- predict(fit, newdata = data.ext)

  # -------- recalibration --------
  # Update model intercept to the new outcome mean
  y_mean_ext <- mean(data.ext$y)
  y_mean_logit_ext <- log(y_mean_ext / (1 - y_mean_ext))
  fit.rc <- fit
  fit.rc$coefficients[1] <- y_mean_logit_ext + offset
  shift.rc <- fit.rc$coefficients[1] - fit$coefficients[1]
  data.ext$pred.p.rc <- predict(fit.rc, newdata = data.ext, type = "response")

  # optimal shift
  fit.opt <- glm(
    y ~ 1 + offset(pred.logit),
    data = data.ext,
    family = "binomial"
  )
  shift.opt <- fit.opt$coefficients[1]
  data.ext$pred.p.opt <- predict(fit.opt, newdata = data.ext, type = "response")

  # -------- calibration metrics --------
  # the original model
  df <- data.frame(
    y = data.ext$y,
    p = data.ext$pred.p,
    logit = data.ext$pred.logit
  )
  plot.o <- plot.calibration(df)$plot

  # the updated model
  df_res <- data.frame()
  pList_1 <- list()
  pList_2 <- list()

  vec_shift <- unique(c(vec_shift, 0, shift.rc, shift.opt))
  for (i_shift in 1:length(vec_shift)) {
    shift <- vec_shift[i_shift]
    # update model with an intercept shift
    fit.new <- fit
    fit.new$coefficients[1] <- fit.new$coefficients[1] + shift
    # prediction
    pred.p.new <- predict(fit.new, newdata = data.ext, type = "response")
    pred.logit.new <- predict(fit.new, newdata = data.ext)
    # calibration
    res.cali <- calibration(
      pred.p.new,
      data.ext$y,
      package = "rms",
      pl = FALSE
    )
    vec_res <- c(shift, res.cali)
    names(vec_res)[1] <- "shift"
    df_res <- bind_rows(df_res, vec_res)

    # plot
    if (plot) {
      df <- data.frame(
        y = data.ext$y,
        p = pred.p.new,
        logit = pred.logit.new
      )
      p1 <- plot.calibration(
        df,
        color = "red",
        titlestr = paste0("shift = ", shift),
        plot.o = plot.o
      )$plot
      pList_1[[i_shift]] <- p1

      p2 <- ggplot(data, aes(pred.p.new)) +
        geom_histogram(bins = 101, fill = "white", color = "black") +
        theme_bw()
      pList_2[[i_shift]] <- p2
    }
  }
  df_res <- df_res %>% 
    mutate(type_shift = case_when(
      shift == 0 ~ "no",
      shift == shift.rc ~ "recalibration",
      shift == shift.opt ~ "optimal",
      TRUE ~ "other"
    ))

  return(list(
    df_res = df_res,
    pList_1 = pList_1,
    pList_2 = pList_2
  ))
}
```

# Run the simulation studies

```{r}
# Fixed parameters
N.dev <- 10000
N.ext <- 10000
target_mu <- 0
target_sd <- 1
mu_shift <- 0.5
xi <- -2
error_sd <- 0.5
seed <- 123

# Simulated parameters
vec_alpha <- seq(0.1, 1, by = 0.1) # default: 0.5 #
vec_drift <- seq(-2, 2, by = 0.5) # default: 0 #
vec_slope <- c(0.5, 0.75, 1, 1.33, 2) # default: 1 #
grid <- expand.grid(alpha = vec_alpha, drift = vec_drift, slope = vec_slope)

# Arbitrary intercept shifts
vec_shift <- seq(-2, 2, by = 0.5)

# Two types of model
vec_model <- c("correct", "incorrect")

file_name <- "Simulation 4.2 results/df_res.csv"
if (file.exists(file_name)) {
  df_res <- read.csv(file_name)
} else {
  df_res <- data.frame()
  for (i_grid in seq_len(nrow(grid))) {
    alpha <- grid$alpha[i_grid]
    drift <- grid$drift[i_grid]
    slope <- grid$slope[i_grid]

    # ----------- Generate data samples -----------
    data <- data.lr(
      N.dev = N.dev,
      N.ext = N.ext,
      target_mu = target_mu,
      target_sd = target_sd,
      mu_shift = mu_shift,
      xi = xi,
      alpha = alpha,
      error_sd = error_sd,
      drift = drift,
      slope = slope,
      seed = seed
    )
    data.dev <- data$data.dev
    data.ext <- data$data.ext
    # number of cases
    ncase.dev <- sum(data.dev$y)
    ncase.ext <- sum(data.ext$y)

    # ----------- Fit and center models -----------
    for (i_model in vec_model) {
      res.fit <- fit.center.lr(data.dev, i_model)
      means.dev <- res.fit$means.dev
      fit <- res.fit$fit.c
      offset <- res.fit$offset

      # ----------- Transport model to the validation sample & Explore different shifts -----------
      res.recal <- recal.lr(data.ext, means.dev, fit, offset, vec_shift, plot = FALSE)
      df_res_i <- res.recal$df_res
      df_res_i$model <- i_model
      df_res_i$alpha <- alpha
      df_res_i$drift <- drift
      df_res_i$slope <- slope
      df_res_i$ncase.dev <- ncase.dev
      df_res_i$ncase.ext <- ncase.ext

      df_res <- bind_rows(df_res, df_res_i)
    }
  }

  write.csv(df_res, row.names = FALSE, file = file_name)
}
head(df_res)
```

```{r}
# ------------------- Plot -------------------
df_Plot <- df_res %>% select(model, type_shift, alpha, drift, slope, shift, Eavg, E90, Emax) 
df_Plot_long <- df_Plot %>% 
  pivot_longer(-c(model, type_shift, alpha, drift, slope, shift), names_to = "metric", values_to = "value") %>% 
  mutate(type_shift_2 = factor(ifelse(type_shift == "other", 1, 2)))
df_Plot_long$metric <- factor(df_Plot_long$metric, levels = c("Eavg", "E90", "Emax"))
df_Plot_long$type_shift <- factor(df_Plot_long$type_shift, levels = c("other", "no", "recalibration", "optimal"))
```

```{r}
alpha_default <- 0.5
drift_default <- 0.5
slope_default <- 1

# influence of alpha
df_Plot_long_alpha <- df_Plot_long %>% filter(drift == drift_default, slope == slope_default) %>% droplevels()
ggplot(df_Plot_long_alpha, aes(shift, value, shape = type_shift, color = alpha)) + 
  geom_point(aes(size = type_shift_2))+
  geom_line(aes(group = alpha), alpha = 0.5, linewidth = 0.5)+
  geom_vline(xintercept = 0, linetype = "dashed", color = "black")+
  scale_shape_manual(values = c("other" = 16, "no" = 19, "recalibration" = 23, "optimal" = 8))+
  scale_size_manual(values = c(1, 3))+
  facet_wrap(model ~ metric, ncol = 3, scales = "free")+
  theme_bw()

# influence of drift
df_Plot_long_drift <- df_Plot_long %>% filter(alpha == alpha_default, slope == slope_default) %>% droplevels()
ggplot(df_Plot_long_drift, aes(shift, value, shape = type_shift, color = drift)) + 
  geom_point(aes(size = type_shift_2))+
  geom_line(aes(group = drift), alpha = 0.5, linewidth = 0.5)+
  geom_vline(xintercept = 0, linetype = "dashed", color = "black")+
  scale_shape_manual(values = c("other" = 16, "no" = 19, "recalibration" = 23, "optimal" = 8))+
  scale_size_manual(values = c(1, 3))+
  facet_wrap(model ~ metric, ncol = 3, scales = "free")+
  theme_bw()

# influence of slope
df_Plot_long_slope <- df_Plot_long %>% filter(alpha == alpha_default, drift == drift_default) %>% droplevels()
ggplot(df_Plot_long_slope, aes(shift, value, shape = type_shift, color = slope)) + 
  geom_point(aes(size = type_shift_2))+
  geom_line(aes(group = slope), alpha = 0.5, linewidth = 0.5)+
  geom_vline(xintercept = 0, linetype = "dashed", color = "black")+
  scale_shape_manual(values = c("other" = 16, "no" = 19, "recalibration" = 23, "optimal" = 8))+
  scale_size_manual(values = c(1, 3))+
  facet_wrap(model ~ metric, ncol = 3, scales = "free")+
  theme_bw()
```

```{r}
df_Plot_long_avg_correct <- df_Plot_long %>% filter(metric == "Eavg", model == "correct", drift %in% c(-2.0, -1.0, 0.0, 1.0, 2.0)) %>% droplevels()
df_Plot_long_avg_correct$drift <- paste0("drift = ", df_Plot_long_avg_correct$drift)
df_Plot_long_avg_correct$slope <- paste0("slope = ", df_Plot_long_avg_correct$slope)
ggplot(df_Plot_long_avg_correct, aes(shift, value, shape = type_shift, color = alpha)) + 
  geom_point(aes(size = type_shift_2))+
  geom_line(aes(group = alpha), alpha = 0.5, linewidth = 0.5)+
  geom_vline(xintercept = 0, linetype = "dashed", color = "black")+
  scale_shape_manual(values = c("other" = 16, "no" = 19, "recalibration" = 23, "optimal" = 8))+
  scale_size_manual(values = c(1, 3))+
  facet_wrap(drift ~ slope, ncol = 5, scales = "free")+
  labs(title = "Correct model")+
  theme_bw()

df_Plot_long_avg_incorrect <- df_Plot_long %>% filter(metric == "Eavg", model == "incorrect", drift %in% c(-2.0, -1.0, 0.0, 1.0, 2.0)) %>% droplevels()
df_Plot_long_avg_incorrect$drift <- paste0("drift = ", df_Plot_long_avg_incorrect$drift)
df_Plot_long_avg_incorrect$slope <- paste0("slope = ", df_Plot_long_avg_incorrect$slope)
ggplot(df_Plot_long_avg_incorrect, aes(shift, value, shape = type_shift, color = alpha)) + 
  geom_point(aes(size = type_shift_2))+
  geom_line(aes(group = alpha), alpha = 0.5, linewidth = 0.5)+
  geom_vline(xintercept = 0, linetype = "dashed", color = "black")+
  scale_shape_manual(values = c("other" = 16, "no" = 19, "recalibration" = 23, "optimal" = 8))+
  scale_size_manual(values = c(1, 3))+
  facet_wrap(drift ~ slope, ncol = 5, scales = "free")+
  labs(title = "Incorrect model")+
  theme_bw()
```

```{r}

```