---
title: "Simulation studies - Recalibration of the Cox model"
author: "Juan Li"
date: 2026-09-02
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false
#| echo: false

# ======== For unknown issue of connecting to R server in positron, load all packages and source code here =========
library(rms)
library(Hmisc)
library(dplyr)
library(stringr)
library(survival)
library(ggplot2)
library(ggpubr) 
library(CalibrationCurves)
library(riskRegression)
library(ggsurvfit)
library(gtsummary)
library(timeROC)

# Prepare data for the centered algorithm
source("R/center_prepare.R", echo = FALSE) 
# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_pred.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
# Wrapper function for calibration curves and metrics
# Including the following functions: `calibration`, `calibration.con`, `calibration.OvsP`
source("R/calibration.R", echo = FALSE)
# Modified function for cox models
source("R/calibration.surv.R", echo = FALSE) 
source("R/calibration_curve_cox.R", echo = FALSE) # valProbSurvival.2
```

# Create simulation dataset

```{r}
set.seed(123)
n <- 500
# Generate covariates 
sex <- rbinom(n, 1, 0.5)
age <- runif(n, min = 40, max = 80)
x <- rnorm(n)
var_mat <- as.matrix(cbind(sex, age, x))

# True coefficient (log hazard ratio) and baseline hazard parameters
beta_vec <- c(0.75, 0.045, -0.3) 
lambda <- 0.01  # scale parameter for Weibull
rho <- 1.5      # shape parameter for Weibull (1 = exponential)

# Generate survival times using inverse cumulative distribution function
u <- runif(n)
time <- (-log(u) / (lambda * exp(var_mat %*% beta_vec)))^(1 / rho)

# Generate random censoring times (e.g., uniform up to max time)
censoring_time <- runif(n, min = 0, max = 150)

# Observed time and status indicator (1 = event, 0 = censored)
obs_time <- pmin(time, censoring_time)
status <- as.numeric(time <= censoring_time)

# Combine into a data frame
sim_data <- data.frame(time = obs_time, status = status, sex = sex, age = age, x = x)

# Fit the Cox model to check
fit <- coxph(Surv(time, status) ~ sex + age + x, data = sim_data)
summary(fit)

survfit2(Surv(time, status) ~ sex, data = sim_data) %>% 
  ggsurvfit() +
  add_confidence_interval() +
  add_risktable()+
  ylim(0,1)
```

```{r}
#| echo: false

original <- sim_data %>% filter(sex == 0) |> droplevels()
target <- sim_data %>% filter(sex == 1) |> droplevels()
```

## Fitting the centered model

```{r}
#| echo: false

# The original model
fit.o.c <- coxph(Surv(time, status) ~ age + x, data = original,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL) 
fit.o.c
```

## Comparing different methods to re-calibrate the centered model to a new dataset

### Calibration of the original centered model on `original`

```{r}
#| echo: false

timeHorizon <- 3

pred.lp.o <- predict(fit.o.c, original, type = "lp")
surv.fit <- survfit(fit.o.c, newdata = original)
pred.prob.o <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier.o <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- original$time
outcome.event <- original$status

res.o.o <- valProbSurvival.2(
  pred.lp.o,
  pred.prob.o,
  pred.prob.brier.o,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
```

### Calibration of the original centered model on `target`

```{r}
#| echo: false

timeHorizon <- 3

pred.lp.o <- predict(fit.o.c, target, type = "lp")
surv.fit <- survfit(fit.o.c, newdata = target)
pred.prob.o <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier.o <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- target$time
outcome.event <- target$status

res.o <- valProbSurvival.2(
  pred.lp.o,
  pred.prob.o,
  pred.prob.brier.o,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
```

### Recalibration with single shift: `delta_recal_KM`

Based on methods in [Liu *et al*, 2025](https://doi.org/10.1016/j.jclinepi.2025.111895).

1.  Suppose the probability of event at a specific `timeHorizon`, is $\theta = 1 - S_{KM}(timeHorizon)$, then

- Observed risk for derivation population is $\theta_{derived} = 1 - S^{derived}_{KM}(timeHorizon)$

- Observed risk for target population is $\theta_{target} = 1 - S^{target}_{KM}(timeHorizon)$

2.  The rescaling factor proposed in the paper was $\Delta_{rescaling} = ln[-ln(1-\theta_{derived})] - ln[-ln(1-\theta_{target})]$.

3.  Then, the recalibrated risk is $risk_{recal} = 1 - exp(-exp(ln[-ln(1-risk_{ori})]-\Delta_{rescaling}))$.

4.  Let's substitute $S_{recal} = 1 - risk_{recal}$ and $S_{ori} = 1 - risk_{ori}$ into the equation, then

$$\begin{aligned}
S_{recal} &= exp(-exp(ln[-ln(S_{ori})]-\Delta_{rescaling})) \\
          &= e^{-e^{ln[-ln(S_{ori})]-\Delta_{rescaling}}} \\
          &= e^{-e^{ln[-ln(S_{ori})]}e^{-\Delta_{rescaling}}} \\
          &= [e^{-e^{ln[-ln(S_{ori})]}}]^{e^{-\Delta_{rescaling}}} \\
          &= [e^{ln(S_{ori})}]^{e^{-\Delta_{rescaling}}} \\
          &= [S_{ori}]^{e^{-\Delta_{rescaling}}}
\end{aligned}$$

5.  In Cox, the survival function is $S(t|X_i) = [S_0(t)]^{exp(LP)}$. Therefore, the equation above can be written as

$$\begin{aligned}
S_{recal} &= [S_{ori}]^{e^{-\Delta_{rescaling}}} \\
          &= ([S_{0,ori}(t)]^{e^{LP}})^{e^{-\Delta_{rescaling}}} \\
          &= S_{0,ori}(t)^{e^{LP}e^{-\Delta_{rescaling}}} \\
          &= S_{0,ori}(t)^{e^{(LP -\Delta_{rescaling})}}
\end{aligned}$$

i.e. $LP_{recal} = LP_{ori} -\Delta_{rescaling}$

6.  Note, in my implementation, I make

`delta_recal` = $-\Delta_{rescaling} = ln[-ln(1-\theta_{target})] - ln[-ln(1-\theta_{derived})]$. Therefore:

$S_{recal} = [S_{ori}]^{e^{delta.recal}}$

$LP_{recal} = LP_{ori}+delta.recal$

```{r}
#| echo: false

survfit_original_KM <- survfit(Surv(time, status) ~ 1, data = original)
prob_original_KM <- 1 - summary(survfit_original_KM, times = timeHorizon)$surv
prob_original_KM_cloglog <- log(-log(1-prob_original_KM)) 
print(paste0("Original, KM event probability at year 3: ", prob_original_KM))
print(paste0("Clog-log: ", prob_original_KM_cloglog))
      
survfit_target_KM <- survfit(Surv(time, status) ~ 1, data = target)
prob_target_KM <- 1 - summary(survfit_target_KM, times = timeHorizon)$surv
prob_target_KM_cloglog <- log(-log(1-prob_target_KM)) 
print(paste0("Target, KM event probability at year 3: ", prob_target_KM))
print(paste0("Clog-log: ", prob_target_KM_cloglog))

delta_recal_KM <- prob_target_KM_cloglog - prob_original_KM_cloglog
print(paste0("delta_recal_KM: ", delta_recal_KM))

# calibration of the recalibrated model
pred.lp.recal <- pred.lp.o + delta_recal_KM
pred.prob.recal <- 1 - (1 - pred.prob.o) ^ exp(delta_recal_KM)
pred.prob.brier.recal <- 1 - (1 - pred.prob.brier.o) ^ exp(delta_recal_KM)

# check
pred.prob.recal.2 <- 1 - exp(-exp(log(-log(1 - pred.prob.o)) + delta_recal_KM))
summary(pred.prob.recal.2 - pred.prob.recal)

res.recal.KM <- valProbSurvival.2(
  pred.lp.recal,
  pred.prob.recal,
  pred.prob.brier.recal,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
```

### Recalibration with single shift: `delta_recal_model_original`

Main idea is similar to above, with some modifications of how to define $\Delta_{rescaling}$:

1.  Suppose the probability of event at a specific `timeHorizon`, is $\theta = 1 - S_{KM}(timeHorizon)$, then

- Average predicted risk for **derivation** population is $\theta_{derived} = \frac{\sum^{N_{derived}}_i{\theta_{i, predicted}}}{N_{derived}}$

- Observed risk for target population is $\theta_{target} = 1 - S^{target}_{KM}(timeHorizon)$ (same as above)

2.  The rescaling factor is $delta.recal = ln[-ln(1-\theta_{target})] - ln[-ln(1-\theta_{derived})]$.

```{r}
#| echo: false

surv.fit.o <- survfit(fit.o.c, newdata = original)
prob_original_model <- mean(1 - t(summary(surv.fit.o, times = timeHorizon)$surv))
prob_original_model_cloglog <- log(-log(1-prob_original_model)) 
print(paste0("Original, average event probability at year 3: ", prob_original_model))
print(paste0("Clog-log: ", prob_original_model_cloglog))
print(paste0("Target, KM event probability at year 3: ", prob_target_KM))
print(paste0("Clog-log: ", prob_target_KM_cloglog))

delta_recal_model_original <- prob_target_KM_cloglog - prob_original_model_cloglog
print(paste0("delta_recal_model_original: ", delta_recal_model_original))

# calibration of the recalibrated model
pred.lp.recal <- pred.lp.o + delta_recal_model_original
pred.prob.recal <- 1 - (1 - pred.prob.o) ^ exp(delta_recal_model_original)
pred.prob.brier.recal <- 1 - (1 - pred.prob.brier.o) ^ exp(delta_recal_model_original)

res.recal.model.original <- valProbSurvival.2(
  pred.lp.recal,
  pred.prob.recal,
  pred.prob.brier.recal,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
```

### Recalibration with single shift: `delta_recal_model_target`

Main idea is similar to above, with some modifications of how to define $\Delta_{rescaling}$:

1.  Suppose the probability of event at a specific `timeHorizon`, is $\theta = 1 - S_{KM}(timeHorizon)$, then

- Average predicted risk for **target** population is $\theta_{predicted} = \frac{\sum^{N_{target}}_i{\theta_{i, predicted}}}{N_{target}}$

- Observed risk for target population is $\theta_{observed} = 1 - S^{target}_{KM}(timeHorizon)$ (same as above)

2.  The rescaling factor is $delta.recal = ln[-ln(1-\theta_{observed})] - ln[-ln(1-\theta_{predicted})]$.

```{r}
#| echo: false

prob_target_model <- mean(pred.prob.o)
prob_target_model_cloglog <- log(-log(1-prob_target_model)) 
print(paste0("Target, average event probability at year 3: ", prob_target_model))
print(paste0("Clog-log: ", prob_target_model_cloglog))
print(paste0("Target, KM event probability at year 3: ", prob_target_KM))
print(paste0("Clog-log: ", prob_target_KM_cloglog))

delta_recal_model_target <- prob_target_KM_cloglog - prob_target_model_cloglog
print(paste0("delta_recal_model_target: ", delta_recal_model_target))

# calibration of the recalibrated model
pred.lp.recal <- pred.lp.o + delta_recal_model_target
pred.prob.recal <- 1 - (1 - pred.prob.o) ^ exp(delta_recal_model_target)
pred.prob.brier.recal <- 1 - (1 - pred.prob.brier.o) ^ exp(delta_recal_model_target)

res.recal.model.target <- valProbSurvival.2(
  pred.lp.recal,
  pred.prob.recal,
  pred.prob.brier.recal,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
```

### Recalibration with approximating Platt: method 1

In this version, the predicted event probability in the averaged predicted probability in each age group within the target population. This is an extension of the `delta_recal_model` above, and loosely follows the methods in [SCORE2 risk prediction algorithms for CVD](doi:10.1093/eurheartj/ehab309), Supplementary Methods, Figure 2.

1.  Calculate the predict event probability for target population at `timeHorizon`: $\theta_{pred}$.

2.  Calculate the average within each age group: $\theta_{pred, agegroup} = mean_{agegroup}(\theta_{pred})$

3.  Estimate observed event probability at `timeHorizon` within each age group:$\theta_{expect,agegroup}$

4.  Regress the age-specific expected on predicted risks: $ln[-ln(1-\theta_{expect,agegroup})] = Scale1 + Scale2 \times ln[-ln(1-\theta_{pred, agegroup})]$

5.  Then the recalibrated event probability is calculated as $\theta_{pred,new}=1-exp(-exp(Scale1 + Scale2 \times ln[-ln(1-\theta_{pred})]))$

6.  Let's substitute $S_{pred,new} = 1 - risk_{pred,new}$ and $S_{pred} = 1 - risk_{pred} = S_0(t)^{exp(LP)}$ into the equation, then

$$\begin{aligned}
S_{pred,new} &= exp(-exp(Scale1 + Scale2 \times ln[-ln(S_{pred})])) \\
          &= exp(-exp(Scale1 + Scale2 \times ln[-ln(S_0(t)^{exp(LP)})])) \\
          &= exp(-exp(Scale1 + Scale2 \times ln[-exp(LP)ln(S_0(t))])) \\
          &= exp(-exp(Scale1 + Scale2 \times (LP + ln[-ln(S_0(t))]))) 
\end{aligned}$$ Therefore,

$$\begin{aligned}
ln(-ln(S_{pred,new})) &= ln(-ln(exp(-exp(Scale1 + Scale2 \times (LP + ln[-ln(S_0(t))]))))) \\
          &= Scale1 + Scale2 \times (LP + ln[-ln(S_0(t))]) \\
          &= Scale1 + Scale2 \times LP + Scale2 \times ln[-ln(S_0(t))]
\end{aligned}$$

Since $S_{pred,new} = S_0(t)^{exp(LP_{new})}$, we have

$$\begin{aligned}
ln(-ln(S_{pred,new})) &= ln(-ln(S_0(t)^{exp(LP_{new})})) \\
          &= ln(-exp(LP_{new})ln(S_0(t))) \\
          &= ln(exp(LP_{new})) + ln(-ln(S_0(t))) \\
          &= LP_{new} + ln(-ln(S_0(t)))
\end{aligned}$$

Therefore, $LP_{new} + ln(-ln(S_0(t))) = Scale1 + Scale2 \times LP + Scale2 \times ln[-ln(S_0(t))]$, i.e.

$LP_{new} = Scale1 + Scale2 \times LP + (Scale2-1) \times ln[-ln(S_0(t))]$

```{r}
#| echo: false

df <- data.frame(
  age = target$age,
  theta_pred = pred.prob.o,
  time = target$time,
  status = target$status)

# age groups
df <- df %>%
  mutate(
    cage = cut(
      age,
      breaks = c(20, seq(40, 65, 5), 80)
    )
  ) 
table(df$cage)

# get mean values for each variables in each age group
df2 <- df %>%
  group_by(cage) %>%
  summarise_all(mean)

# observed event probability in each age group at timeHorizon
survfit_cage_KM <- survfit(Surv(time, status) ~ cage, data = df)
df2$theta_exp <- 1 - summary(survfit_cage_KM, times = timeHorizon)$surv
df2$theta_pred_t <- log(-log(1-df2$theta_pred))
df2$theta_exp_t <- log(-log(1-df2$theta_exp))

ggplot(df2, aes(theta_pred, theta_exp))+
  geom_point()+
  geom_smooth(method = "lm")+
  coord_fixed()+
  xlim(0,1)+
  ylim(0,1)

# get the scaling factors
fit <- lm(theta_exp_t ~ theta_pred_t, data = df2)
fit
scale1 <- coef(fit)[1]
scale2 <- coef(fit)[2]

# get baseline survival S_0(timeHorizon)
base_surv <- survfit(fit.o.c)
S_0 <- summary(base_surv, times = timeHorizon)$surv

# get the recalibrated event probability at timeHorizon
pred.prob.recal <- 1 - exp(-exp(scale1 + scale2*log(-log(1-pred.prob.o))))
pred.prob.brier.recal <- 1 - exp(-exp(scale1 + scale2*log(-log(1-pred.prob.brier.o))))

# get the recalibrated LP
pred.lp.recal <- scale1 + scale2 * pred.lp.o + (scale2-1)*log(-log(S_0))

# check
summary(pred.prob.recal - (1 - S_0^exp(pred.lp.recal)))

res.recal.model.agegroup <- valProbSurvival.2(
  pred.lp.recal,
  pred.prob.recal,
  pred.prob.brier.recal,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
scale1_m1 <- scale1
scale2_m1 <- scale2
```

Because age is not strongly associated with cancer risk in this dataset, especially the predicted risk are quite similar, recalibration using linear regression makes the calibration much worse. Below try the regression with `offset()`, i.e. only adjust the intercept instead.

```{r}
#| echo: false

# Offset
fit <- lm(theta_exp_t ~ offset(theta_pred_t), data = df2)
fit
delta_recal_model_agegroup <- coef(fit)[1]
print(paste0("delta_recal_model_agegroup: ", delta_recal_model_agegroup))

# calibration of the recalibrated model
pred.lp.recal <- pred.lp.o + delta_recal_model_agegroup
pred.prob.recal <- 1 - (1 - pred.prob.o) ^ exp(delta_recal_model_agegroup)
pred.prob.brier.recal <- 1 - (1 - pred.prob.brier.o) ^ exp(delta_recal_model_agegroup)

res.recal.model.agegroup.offset <- valProbSurvival.2(
  pred.lp.recal,
  pred.prob.recal,
  pred.prob.brier.recal,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
```

### Recalibration with approximating Platt: method 2

This method closely follows the methods in SCORE2, but it requires to unpack the data (dummy, rcs, interactions) and some modifications of the code.

(2026-08-26) Based on the methods used in [SCORE2 risk prediction algorithms for CVD](doi:10.1093/eurheartj/ehab309), instead of single summary stats, we can create a vector for multiple age categories (also separated by sex if the model was sex-stratified) to enable Platt scaling using unlinked summary stats.

```{r}
#| echo: false

# age groups
df0 <- target %>%
  mutate(
    cage = cut(
      age,
      breaks = c(20, seq(40, 65, 5), 80)
    )
  ) 
table(df0$cage)

# The centered model, for variable means
vec_mean <- fit.o.c$means
# unpack the target data for the centered algorithm
df <- center_prepare_0(fit.o.c, df0, center = TRUE, add.outcome = TRUE, means_ext = vec_mean)$data
df$cage <- df0$cage

# get mean values for each variables in each age group
df2 <- df %>%
  select(-c(time, status)) %>%
  group_by(cage) %>%
  summarise_all(mean)

# get baseline survival S_0(timeHorizon)
base_surv <- survfit(fit.o.c)
S_0 <- summary(base_surv, times = timeHorizon)$surv

# observed event probability in each age group at timeHorizon
survfit_cage_KM <- survfit(Surv(time, status) ~ cage, data = df)
df2$theta_exp <- 1 - summary(survfit_cage_KM, times = timeHorizon)$surv
df2$theta_exp_t <- log(-log(1-df2$theta_exp))

# predicted event probability in each age group at timeHorizon
v_mat <- df2 |> select(contains("_C"))
df2$LP <- as.matrix(v_mat) %*% coef(fit.o.c)
df2$theta_pred <- 1 - S_0 ^ exp(df2$LP)
df2$theta_pred_t <- log(-log(1-df2$theta_pred))

ggplot(df2, aes(theta_pred, theta_exp))+
  geom_point()+
  geom_smooth(method = "lm")+
  coord_fixed()+
  xlim(0,1)+
  ylim(0,1)

# get the scaling factors
fit <- lm(theta_exp_t ~ theta_pred_t, data = df2)
fit
scale1 <- coef(fit)[1]
scale2 <- coef(fit)[2]

# get the recalibrated event probability at timeHorizon
pred.prob.recal <- 1 - exp(-exp(scale1 + scale2*log(-log(1-pred.prob.o))))
pred.prob.brier.recal <- 1 - exp(-exp(scale1 + scale2*log(-log(1-pred.prob.brier.o))))

# get the recalibrated LP
pred.lp.recal <- scale1 + scale2 * pred.lp.o + (scale2-1)*log(-log(S_0))

# check
summary(pred.prob.recal - (1 - S_0^exp(pred.lp.recal)))

res.recal.model.agegroup.2 <- valProbSurvival.2(
  pred.lp.recal,
  pred.prob.recal,
  pred.prob.brier.recal,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
scale1_m2 <- scale1
scale2_m2 <- scale2
```

Similar as above, check `offset()`.

```{r}
#| echo: false

# Offset
fit <- lm(theta_exp_t ~ offset(theta_pred_t), data = df2)
fit
delta_recal_model_agegroup_2 <- coef(fit)[1]
print(paste0("delta_recal_model_agegroup_2: ", delta_recal_model_agegroup_2))

# calibration of the recalibrated model
pred.lp.recal <- pred.lp.o + delta_recal_model_agegroup_2
pred.prob.recal <- 1 - (1 - pred.prob.o) ^ exp(delta_recal_model_agegroup_2)
pred.prob.brier.recal <- 1 - (1 - pred.prob.brier.o) ^ exp(delta_recal_model_agegroup_2)

res.recal.model.agegroup.2.offset <- valProbSurvival.2(
  pred.lp.recal,
  pred.prob.recal,
  pred.prob.brier.recal,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
```

### Recalibration with Platt scaling

All above recalibration methods don't require the `target` dataset be linked to outcome variables. Only this Platt scaling requires data linkage.

**Note:** the `offset(lp)` method doesn't work here

```{r}
#| echo: false

# update intercept using Platt's calibration, which requires linked individual-level newdata
# Use `valProbSurvival.2` directly here
target$LP <- pred.lp.o
fit.platt <- coxph(Surv(time, status) ~ LP, data = target,
model = FALSE, x = FALSE, y = FALSE)
fit.platt

pred.lp <- predict(fit.platt, target, type = "lp")
surv.fit <- survfit(fit.platt, newdata = target)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
outcome.time <- target$time
outcome.event <- target$status

res.Platt <- valProbSurvival.2(
  pred.lp,
  pred.prob,
  pred.prob.brier,
  outcome.time,
  outcome.event,
  timeHorizon = timeHorizon,
  plotCal = "ggplot",
  nk = 5,
  timeROC = FALSE
)
```

```{r}
#| echo: false

res.o$stats$Calibration$Statistics
print(paste0("delta_recal_KM: ", delta_recal_KM))
res.recal.KM$stats$Calibration$Statistics
print(paste0("delta_recal_model_original: ", delta_recal_model_original))
res.recal.model.original$stats$Calibration$Statistics
print(paste0("delta_recal_model_target: ", delta_recal_model_target))
res.recal.model.target$stats$Calibration$Statistics
print(paste0("scale1: ", scale1_m1, "; scale2: ", scale2_m1))
res.recal.model.agegroup$stats$Calibration$Statistics
print(paste0("delta_recal_model_agegroup: ", delta_recal_model_agegroup))
res.recal.model.agegroup.offset$stats$Calibration$Statistics
print(paste0("scale1: ", scale1_m2, "; scale2: ", scale2_m2))
res.recal.model.agegroup.2$stats$Calibration$Statistics
print(paste0("delta_recal_model_agegroup_2: ", delta_recal_model_agegroup_2))
res.recal.model.agegroup.2.offset$stats$Calibration$Statistics
res.Platt$stats$Calibration$Statistics
```

```{r}

```