```         
---
title: "Centered algorithms using real data - Cox"
author: "Juan Li"
date: 2026-01-21
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---
```

```{r}
#| warning: false
#| message: false
#| output: false

library(dplyr) # Data manipulation
library(tidyr) # Tidy messy data
library(ggplot2) # Data visualization
library(ggpubr) # ggarrange
library(gtsummary)

library(rms)
library(polspline)

library(survival)
library(ggsurvfit)
library(survminer)

library(CalibrationCurves) # https://cran.r-project.org/web/packages/CalibrationCurves/vignettes/CalibrationCurves.html

source("R/centering.R") 
# functions in "centering.R"
# get_rcs: implement the formula of rcs components
# step_dummy: dummy variables
# step_rcs: unpack the rcs components
# step_interaction: interaction terms
# step_center: centering
# get_mean: get mean values
# root.search: Root searching, not used in this .qmd file

source("R/calibration.R") # for calibration plots
```

# Load Data - Breast Cancer Survival Data from Rotterdam and Germany

See `?CalibrationCurves::trainDataSurvival`.

```{r}
# variables for `time` and `status` are `ryear` and `rfs`, respectively
data(trainDataSurvival) # Rotterdam
data(testDataSurvival) # German Breast Cancer Study Group
```

Kaplan-Meier plots: https://www.emilyzabor.com/survival-analysis-in-r.html

```{r}
data <- bind_rows(
  trainDataSurvival %>% select(ryear, rfs) %>% mutate(type = "train"),
  testDataSurvival %>% select(ryear, rfs) %>% mutate(type = "test")
)
data$type <- factor(data$type, levels = c("train", "test"))

survfit2(Surv(ryear, rfs) ~ type, data = data) %>% 
  ggsurvfit() +
  add_confidence_interval() +
  add_risktable()
```

## Variables

### in both datasets

`pid`: patient identifier

`age`: age at surgery

`meno`: menopausal status (0 = premenopausal, 1 = postmenopausal)

`size`: tumor size, a factor with levels \<= 20, 20-50, \>50

- `csize`: Categorized tumor size, copied from size (factor)

`grade`: differentiation grade

- `grade3`: Recoded grade factor (levels: "1-2", "3")

`nodes`: number of positive lymph nodes

- `nodes2`: Winsorized node count (numeric)

- `nodes3`: Restricted cubic spline basis for nodes2 (numeric)

- `cnode`: Categorized node involvement (factor: "0", "1-3", "\>3")

`pgr`: progesterone receptors (fmol/l)

- `pgr2`: Winsorized progesterone receptor level (numeric)

- `pgr3`: Restricted cubic spline basis for original pgr (numeric)

`er`: estrogen receptors (fmol/l)

`hormon`: hormonal treatment (0 = no, 1 = yes)

`ryear`: Follow-up time for RFS, in years (numeric)

`rfs`: Recurrence-free survival status (0 = no event, 1 = event) (numeric)

`epoch`: Follow-up epoch indicator after splitting at 5 years (numeric)

### only in `trainDataSurvival`

`year`: year of surgery

`chemo`: chemotherapy

`rtime`: days to relapse or last follow-up

`recur`: 0 = no relapse, 1 = relapse

`dtime`: days to death or last follow-up

`death`: 0 = alive, 1 = dead

`tstart`

### only in `testDataSurvival`

`rfstime`, `status`

```{r}
data.dev <- trainDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
data.ext <- testDataSurvival %>% select(ryear, rfs, csize, cnode, grade3, pgr2)
summary(data.dev)
summary(data.ext)
```

# Original and centered models using the `data.dev` dataset

Here we will use function `CalibrationCurves::valProbSurvival`, which seems to be more flexible.

**Technical notes for `basehaz`:**

In the [document](https://www.rdocumentation.org/packages/survival/versions/3.8-3/topics/basehaz) for `basehaz`, it says that the default value for `centered` is `TRUE`. However, this centering is **only for continuous variables**. When the model only includes categorical variables, changing the parameter `centered` would yield same outputs. Also, as will be shown below, this is not identical to the cumulative baseline hazard of the centered algorithm.

```{r}
fit.o <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3 + pgr2, data = data.dev,
             x = TRUE, y = TRUE)
fit.o %>% tbl_regression(exp = TRUE) 

fit.o.h0 <- basehaz(fit.o) 
fit.o.h0.nc <- basehaz(fit.o, centered = FALSE) 

# Verity it is the same as following
sfit <- survfit(fit.o, se.fit = FALSE)
chaz <- sfit$cumhaz

head(cbind("time" = fit.o.h0$time, "hazard" = fit.o.h0$hazard, chaz, "hazard.nc" = fit.o.h0.nc$hazard))
```

```{r}
# Get the centered variables and terms
# -------- dummy variables --------
vars_cat <- c("csize", "cnode", "grade3")
data.dev <- step_dummy(data.dev, vars_cat)

# -------- centering --------
vars_mean <- names(data.dev)
# get mean values
means.dev <- get_mean(data.dev, vars_mean)
# centering vars in the data.dev dataset
vars_center <- names(means.dev)
data.dev <- step_center(data.dev, vars_center, means.dev)
```

```{r}
# The centered model on the train dataset
fit.c <- coxph(Surv(ryear, rfs) ~ `csize_20-50_C` + `csize_>50_C` + `cnode_1-3_C` + `cnode_>3_C` + grade3_3_C + pgr2_C, 
              data = data.dev, x=TRUE, y=TRUE)
cbind(coef(fit.o), coef(fit.c))

fit.c.h0 <- basehaz(fit.c) 
fit.c.h0.nc <- basehaz(fit.c, centered = FALSE) 
head(cbind("time" = fit.o.h0$time, fit.o.h0$hazard, fit.o.h0.nc$hazard, fit.c.h0$hazard, fit.c.h0.nc$hazard))
```

# Apply the models on the `data.ext` dataset

```{r}
# Preprocess the `data.ext` dataset

# -------- dummy variables --------
data.ext <- step_dummy(data.ext, vars_cat)

# -------- centering --------
# 1. centering using the mean values in the `data.dev` dataset
# This is for transporting the centered algorithm
data.ext.d <- step_center(data.ext, vars_center, means.dev)

# 2. centering using the mean values in the `data.ext` dataset
# This is for recalibration
# get mean values
means.ext <- get_mean(data.ext, vars_mean)
# centering vars in the data.ext dataset
data.ext.t <- step_center(data.ext, vars_center, means.ext)
```

```{r}
valProbSurvival(fit.o, data.dev, plotCal = "ggplot", nk = 5)
```

```{r}
valProbSurvival(fit.o, data.ext, plotCal = "ggplot", nk = 5)
```

Note 1: Should multiply a constant to the baseline hazard, rather than add a constant. -\> accumulative incident rate.

Note 2: How to update the model to achieve it?

```{r}

```