---
title: "Simulation studies - Recalibration"
author: "Juan Li"
date: 2026-04-14
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(dplyr) # Data manipulation
library(tidyr) # Tidy messy data
library(ggplot2) # Data visualization
library(ggpubr) # ggarrange
library(gtsummary)
library(rms)
library(fGarch) # for skewed distribution with target mean and sd

source("R/util.R")
# functions in "util.R"
# get_rcs: implement the formula of rcs components
# step_dummy: dummy variables
# step_rcs: rcs
# step_interaction: interaction terms
# step_center: centering
# get_mean: get mean values from the original dataset
# root.search: Root searching, not in use in this .qmd file

source("R/calibration.R") # for calibration plots
```

# Simulation study 2 - Recalibration of logistic regression

In this simulation study, we will modify a similar simulation approach as in DOI: 10.1002/sim.8281.

1. We simulated derivation and validation samples of size N = 10000

2. For each subject in the derivation sample, we simulated a continuous covariate from a standard normal distribution: x ~ N(target_mu, target_sd). Default: target_mu = 0, target_sd = 1. 

3. In the validation sample, we simulated x using a skew-normal distribution with the shifted mean and same sd as in the derivation sample. 

4. For each subject in the derivation sample, a linear predictor was defined as LP = x + error, where error ~ N(0, sigma). And sigma is determined by the target_R2 value. [ref](https://stats.stackexchange.com/questions/636192/simulate-data-with-specified-signal-to-noise-ratio-for-a-given-sigma).

5. For each subject in the validation sample, a linear predictor with __intercept drift__ was defined as LP = drift + x + error, where error ~ N(0, sigma).

6. A binary outcomewas simulated from a Bernoulli distribution: Y = Be(p), where p = exp(LP)/(1+exp(LP)).

7. Two logistic regression models were then fit to the simulated derivation sample: (i) a correctly specified regression model; and (ii) an incorrectly specified regression model that add the quadratic term (x^2).

8. The two models were transferred into centered models using the specific steps.

9. The two centered models were applied to the validation sample and recalibrated with various shifts of the intercept.

10. Eavg (ICI), E90, and Emax were used as the main evaluation metrics.

```{r}
# ------------------- Generate data -------------------
data.lr <- function(
  N.dev = 10000,
  N.ext = 10000,
  target_mu = 0,
  target_sd = 1,
  target_R2 = 0.8,
  drift = 1,
  mu_shift = 0.5,
  xi = -2,
  seed = 123
) {
  # N.dev     ...Sample size of the development dataset
  # N.ext     ...Sample size of the validation/external dataset
  # target_mu ...Target mean of the two distributions
  # target_sd ...Target standard deviation of the two distributions
  # target_R2 ...Target R2 for the ratio between signal and noise
  # drift     ...Intercept drift in the validation dataset
  # mu_shift  ...Mean shift in the validation dataset
  # xi        ...Shape/skewness parameter of the skew-normal distribution. 
  #              If xi>0, right-skewed, positive skewness. If xi<0, left-skewed, negative skewness.
  # seed      ...Random seed for reproducibility

  set.seed(seed)

  # sigma: sd of error
  sigma <- 1 / sqrt(1 / (1 - target_R2) - 1)

  # derivation dataset
  data.dev <- data.frame(
    x = rnorm(n = N.dev, mean = target_mu, sd = target_sd),
    error = rnorm(n = N.dev, mean = 0, sd = sigma)
  ) %>%
    mutate(
      x2 = x^2, # prepared for the incorrectly speficied model
      lp = x + error, # the linear predictor
      p = exp(lp) / (1 + exp(lp)) # inverse-logit function, can also use plogis(lp)
    )
  data.dev$y <- rbinom(n = N.dev, size = 1, prob = data.dev$p)

  # validation/external dataset
  data.ext <- data.frame(
    x = rsnorm(n = N.ext, mean = target_mu + mu_shift, sd = target_sd, xi = xi),
    error = rnorm(n = N.ext, mean = 0, sd = sigma)
  ) %>%
    mutate(
      x2 = x^2, # prepared for the incorrectly speficied model
      lp = drift + x + error, # the linear predictor
      p = exp(lp) / (1 + exp(lp)) # inverse-logit function, can also use plogis(lp)
    )
  data.ext$y <- rbinom(n = N.ext, size = 1, prob = data.ext$p)

  return(list(data.dev = data.dev, data.ext = data.ext))
}

# ------------------- Fit and center models -------------------
fit.center.lr <- function(data.dev, data.ext) {
  # -------- centering: data.dev --------
  vars_mean <- c("x", "x2")
  # get mean values
  means.dev <- get_mean(data.dev, vars_mean)
  # centering vars in the data.dev dataset
  vars_center <- names(means.dev)
  data.dev <- step_center(data.dev, vars_center, means.dev)

  # -------- The original models --------
  fit.o.1 <- glm(y ~ x, data = data.dev, family = "binomial") # correct
  fit.o.2 <- glm(y ~ x + x2, data = data.dev, family = "binomial") # incorrect

  # -------- The centered models --------
  fit.c.1 <- glm(y ~ x_C, data = data.dev, family = "binomial") # correct
  fit.c.2 <- glm(y ~ x_C + x2_C, data = data.dev, family = "binomial") # incorrect

  # outcome mean
  y_mean_dev <- mean(data.dev$y)
  y_mean_logit_dev <- log(y_mean_dev/(1-y_mean_dev))
  # global offset
  offset.1 <- coef(fit.c.1)[1] - y_mean_logit_dev
  offset.2 <- coef(fit.c.2)[1] - y_mean_logit_dev

  # -------- centering: data.ext --------
  # centering vars in the data.ext dataset using mean values in the `data.dev` dataset
  data.ext <- step_center(data.ext, vars_center, means.dev)

  #  -------- prediction --------
  data.ext$prop.o.1 <- predict(fit.o.1, newdata = data.ext, type = "response")
  data.ext$logit.o.1 <- predict(fit.o.1, newdata = data.ext)
  data.ext$prop.o.2 <- predict(fit.o.2, newdata = data.ext, type = "response")
  data.ext$logit.o.2 <- predict(fit.o.2, newdata = data.ext)
  data.ext$prop.c.1 <- predict(fit.c.1, newdata = data.ext, type = "response")
  data.ext$logit.c.1 <- predict(fit.c.1, newdata = data.ext)
  data.ext$prop.c.2 <- predict(fit.c.2, newdata = data.ext, type = "response")
  data.ext$logit.c.2 <- predict(fit.c.2, newdata = data.ext)

  return(list(
    means.dev = means.dev,
    fit.o.1 = fit.o.1,
    fit.o.2 = fit.o.2,
    fit.c.1 = fit.c.1,
    fit.c.2 = fit.c.2,
    y_mean_dev = y_mean_dev,
    y_mean_logit_dev = y_mean_logit_dev,
    offset.1 = offset.1,
    offset.2 = offset.2,
    data.ext = data.ext
  ))
}

# ------------------- Recalibration -------------------
recal.lr <- function(data.ext, fit, offset) {
  # Update model intercept to the new outcome mean
  y_mean_ext <- mean(data.ext$y)
  y_mean_logit_ext <- log(y_mean_ext/(1-y_mean_ext))
  fit.rc <- fit
  fit.rc$coefficients[1] <- y_mean_logit_ext + offset
  shift.rc <- fit.rc$coefficients[1] - fit$coefficients[1]
  prop.rc <- predict(fit.rc, newdata = data.ext, type = "response")

  # optimal shift
  # correct model
  fit.opt <- glm(
    y ~ 1 + offset(logit.c.1),
    data = data.ext,
    family = "binomial"
  )
  shift.opt <- fit.opt$coefficients[1]
  prop.opt <- predict(fit.opt, newdata = data.ext, type = "response")

  return(list(
    y_mean_ext = y_mean_ext,
    y_mean_logit_ext = y_mean_logit_ext,
    fit.rc = fit.rc,
    shift.rc = shift.rc,
    prop.rc = prop.rc,
    fit.opt = fit.opt,
    shift.opt = shift.opt,
    prop.opt = prop.opt
  ))
}

# ------------------- Explore different shifts -------------------
shift.plot <- function(data, fit, vec_shift, shift.rc, shift.opt, plot.o) {
  vec_shift <- c(vec_shift, shift.rc, shift.opt)
  df_res <- matrix(0, nrow = length(vec_shift), ncol = 19)
  vec_metric <- c(
    "Dxy",
    "C (ROC)",
    "R2",
    "D",
    "D:Chi-sq",
    "D:p",
    "U",
    "U:Chi-sq",
    "U:p",
    "Q",
    "Brier",
    "Intercept",
    "Slope",
    "Emax",
    "E90",
    "Eavg",
    "S:z",
    "S:p"
  )
  colnames(df_res) <- c("shift", vec_metric)
  df_res[, 1] <- vec_shift
  df_res <- as.data.frame(df_res)

  plotList <- list()
  for (i in seq_len(length(vec_shift))) {
    shift <- vec_shift[i]
    fit.new <- fit
    fit.new$coefficients[1] <- fit.new$coefficients[1] + shift
    data$prop.new <- predict(fit.new, newdata = data, type = "response")
    data$logit.new <- predict(fit.new, newdata = data)

    res.calibration <- calibration(
      data$prop.new,
      data$y,
      package = "rms",
      pl = FALSE
    )
    df_res[i, ] <- c(shift, res.calibration)

    # Plot
    df <- data %>% select(y, prop.new, logit.new)
    names(df) <- c("y", "p", "logit")
    p1 <- plot.calibration(
      df,
      color = "red",
      titlestr = paste0("shift = ", shift),
      plot.o = plot.o
    )$plot

    p2 <- ggplot(data, aes(prop.new)) +
      geom_histogram(bins = 101, fill = "white", color = "black") +
      theme_bw()

    plotList[[i]] <- ggarrange(p1, p2, nrow = 2, heights = c(1, 0.5))
  }

  df_long <- pivot_longer(
    df_res,
    -shift,
    names_to = "metric",
    values_to = "value"
  )

  df_long$metric <- factor(df_long$metric, levels = vec_metric)
  df_long <- df_long %>%
    mutate(
      color = case_when(
        shift == 0 ~ "green3",
        shift == shift.rc ~ "red",
        shift == shift.opt ~ "orange",
        TRUE ~ "black"
      ),
      size = factor(case_when(
        shift == 0 ~ 3,
        shift == shift.rc ~ 3,
        shift == shift.opt ~ 3,
        TRUE ~ 1
      ))
    )

  plotMetric <- ggplot(df_long, aes(shift, value)) +
    geom_line(alpha = 0.5) +
    geom_point(aes(size = size, color = color)) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_size_manual(values = c(1, 3)) +
    scale_color_manual(
      name = "",
      values = c("black", "green3", "orange", "red"),
      labels = c(
        "Random shift",
        "Original model: shift = 0",
        "Shift by glm with offset",
        "Shift by recalibration"
      )
    ) +
    facet_wrap(~metric, scales = "free") +
    theme_minimal() +
    guides(size = "none") +
    theme(legend.position = "bottom")

  df_long_c <- df_long %>%
    filter(
      metric %in% c("Emax", "Eavg", "E90")
    )
  df_long_c$metric <- factor(df_long_c$metric, levels = c("Eavg", "E90", "Emax"))

  plotMetric2 <- ggplot(df_long_c, aes(shift, value)) +
    geom_line(alpha = 0.5) +
    geom_point(aes(size = size, color = color)) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_size_manual(values = c(1, 3)) +
    scale_color_manual(
      name = "",
      values = c("black", "green3", "orange", "red"),
      labels = c(
        "Random shift",
        "Original model: shift = 0",
        "Shift by glm with offset",
        "Shift by recalibration"
      )
    ) +
    facet_wrap(~metric, scales = "free") +
    theme_minimal() +
    guides(size = "none") +
    theme(legend.position = "bottom")

  return(list(
    plotList = plotList,
    plotMetric = plotMetric,
    plotMetric2 = plotMetric2
  ))
}
```

```{r}
N.dev = 10000
N.ext = 10000
target_mu = 0
target_sd = 1
target_R2 = 0.8
drift = 1
mu_shift = 0.5
xi = -2
seed = 123

data <- data.lr(N.dev, N.ext, target_mu, target_sd, target_R2, drift, mu_shift, xi, seed)
data.dev <- data$data.dev
data.ext <- data$data.ext

p1 <- ggplot(data.dev, aes(x, lp, color = y))+geom_point(alpha=0.5)+theme_minimal()+
  annotate("text", x = mean(data.dev$x), y = max(data.dev$lp), label = paste0("mean = ", round(mean(data.dev$x),2), ", sd = ", round(sd(data.dev$x),2)))
p2 <- ggplot(data.dev, aes(x))+geom_histogram(fill = "white", color = "black", bins = 50)+theme_minimal()
ggarrange(p1, p2, nrow = 2, heights = c(1, 0.3))

p1 <- ggplot(data.ext, aes(x, lp, color = y))+geom_point(alpha=0.5)+theme_minimal()+
  annotate("text", x = mean(data.dev$x), y = max(data.dev$lp), label = paste0("mean = ", round(mean(data.ext$x),2), ", sd = ", round(sd(data.ext$x),2)))
p2 <- ggplot(data.ext, aes(x))+geom_histogram(fill = "white", color = "black", bins = 50)+theme_minimal()
ggarrange(p1, p2, nrow = 2, heights = c(1, 0.3))

dataTable <- rbind(data.dev %>% select(x, y) %>% mutate(cohort = "derivation"), data.ext %>% select(x, y) %>% mutate(cohort = "validation"))
dataTable$cohort <- factor(dataTable$cohort, levels = c("derivation", "validation"))
dataTable %>% tbl_summary(by = "cohort")
```

```{r}
res <- fit.center.lr(data.dev, data.ext)
data.ext <- res$data.ext
fit.c.1 <- res$fit.c.1
fit.c.2 <- res$fit.c.2
offset.1 <- res$offset.1
offset.2 <- res$offset.2

res.rc.1 <- recal.lr(data.ext, fit.c.1, offset.1)
res.rc.2 <- recal.lr(data.ext, fit.c.2, offset.2)
```

```{r}
# Explore different shifts
vec_shift <- sort(c(seq(-2, 2, by = 0.5), seq(drift-0.5, drift+0.5, by = 0.05)))

# Correct model
df <- data.frame(
  y = data.ext$y,
  p = data.ext$prop.c.1,
  logit = data.ext$logit.c.1
)
plot.o.1 <- plot.calibration(df)$plot

resPlot.1 <- shift.plot(data.ext, fit.c.1, vec_shift, res.rc.1$shift.rc, res.rc.1$shift.opt, plot.o.1)
ggarrange(plotlist = resPlot.1$plotList)
#resPlot.1$plotMetric
resPlot.1$plotMetric2

# Inorrect model
df <- data.frame(
  y = data.ext$y,
  p = data.ext$prop.c.2,
  logit = data.ext$logit.c.2
)
plot.o.2 <- plot.calibration(df)$plot

resPlot.2 <- shift.plot(data.ext, fit.c.2, vec_shift, res.rc.2$shift.rc, res.rc.2$shift.opt, plot.o.2)
ggarrange(plotlist = resPlot.2$plotList)
#resPlot.2$plotMetric
resPlot.2$plotMetric2
```

```{r}

```