---
title: "Test the new functions for survival models: fine gray"
author: "Juan Li"
date: 2026-06-30
format: 
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

```{r}
#| warning: false
#| message: false
#| output: false

library(CalibrationCurves)
library(survival)
library(ggsurvfit)
library(gtsummary)
library(riskRegression)
library(prodlim)

# Proper centering of the original model, with sufficient but light-weighted model outputs
source("R/center_fit.R", echo = FALSE) 
# Predicting on the centered model
source("R/center_predict.R", echo = FALSE) 
# Recalibrating the centered model using summary stats in a `newdata`
source("R/center_recalibrate.R", echo = FALSE) 
```

# Monoclonal gammopathy data

See `?survival::mgus2`. A data frame with 1384 observations on the following 10 variables.

`id`: subject identifier

`age`: age at diagnosis, in years

`sex`: a factor with levels F M

`dxyr`: year of diagnosis

`hgb`: hemoglobin

`creat`: creatinine

`mspike`: size of the monoclonal serum splike

`ptime`: time until progression to a plasma cell malignancy (PCM) or last contact, in months

`pstat`: occurrence of PCM: 0=no, 1=yes (**event of interest**)

`futime`: time until death or last contact, in months

`death`: occurrence of death: 0=no, 1=yes (**competing risk**)

```{r}
data <- mgus2
# remove rows with missing
data <- data[complete.cases(data), ]
```

# Model fitting: `riskRegression::FGR` (wrapper of `cmprsk::crr`) vs `survival::finegray` + `coxph`

**Notes for `riskRegression::FGR`**:

**PROS:**

1.  It is more commonly used, with the benefits of `riskRegression::predictRisk` and `riskRegression::plotCalibration` functions.

2.  It doesn't do internal centering by itself, the centering pipeline for linear and logistic regression will be easily apply here.

**CONS:**

1.  However, `riskRegression::FGR` handle non-linear terms in a base R way, which is relatively easy to fix with some functions to prepare the data:

  -  polynomials: using `I(X1) + I(X1^2)` instead of `poly(X1, 2, , raw = TRUE)`

  -  rcs: the formula wrapper does not natively parse built-in spline functions like rms::rcs. Instead, you must manually generate the spline basis variables and pass them as standard linear terms.

  -  It has limited ability to work with more complicated variables names, e.g. `csize20-50`, `rcs(pgr2, 3)pgr2'`.

2.  It is unclear if it can work with `glmnet` for LASSO, or how easy to make it work.

**Notes in this file:**

1.  As discussed above, since the current `riskRegression::FGR` has some syntax issues with `poly`, `rcs`, and interaction, we will just have a simple formula in this file.

2.  `riskRegression::FGR` using `Hist` not `Surv`

3.  `survival::finegray` creates a expanded, weighted dataset, which has more rows then the original dataset. `riskRegression::FGR` doesn't need it, but instead using a multi-level outcome variable `status`: 0 = censor, 1 = event1, 2 = event2, and then specify the event of interest using the parameter `cause`.

4.  Below, I compared results from `riskRegression::FGR` vs `survival::finegray` + `coxph` and found the model coefficients were close but not identical. This is a known phenomenon and it's because the two functions use different approaches to weighting, baseline hazard handling, and ties. (Using `ties = "breslow"` supposes to help with it, but still not identical.)

5.  *Internal Centering*:

- `coxph` centers covariates automatically by subtracting the mean of the covariates (set `nocenter = NULL` to also center on dummy variables), improving convergence speed in iterative Newton-Raphson algorithms. But with `finegray`, it uses the variable means of the expanded `data_fg`, not the original `data` as intended.

- `FGR` does **not** natively perform internal variable centering. It internally relies on binomial regression and pseudo-values rather than partial likelihoods, the internal centering and constraint handling differ.

6.  *Standard Errors:* Because `finegray` manipulates the risk set size by duplicating records, you must use robust standard errors (e.g., `robust = TRUE` in `coxph`, which also requires the `id` column, *i.e.,* `id = id, robust = TRUE`). `riskRegression::FGR` uses a pseudo-value approach that naturally handles these without needing to inflate the risk set size identically.

```{r}
# --------- The FGR model -----------
# Prepare dataset and event variable:
# 0 = censor, 1 = pcm (event of interest), 2 = death (competing risk)
data$time <- ifelse(data$pstat == 0, data$futime, data$ptime)
data$status <- ifelse(data$pstat == 0, 2 * data$death, 1)

# Fit the weighted Cox model on the Fine-Gray data
fit.FGR <- FGR(
  # Using `Hist` instead of `Surv`, only includes linear terms now
  formula = Hist(time, status) ~ hgb + creat + mspike + age, 
  data = data, 
  # specify the event of interest
  cause = 1
)
```

```{r}
# --------- The finegray + coxph model -----------
# Fine-Gray model requires a multi-state survival, need to make it a factor
data$status_fct <- factor(data$status)

# Create the Fine-Gray weighted dataset
data_fg <- finegray(Surv(time, status_fct) ~ ., data = data)

# Fit the weighted Cox model on the Fine-Gray data
fit.survival <- coxph(
  Surv(fgstart, fgstop, fgstatus) ~ hgb + creat + mspike + age, 
  # using the Fine-Gray weighted dataset
  data = data_fg, weight = fgwt,
  # make the model most light-weighted and meet requirements of the secured environment
  model = FALSE, x = FALSE, y = FALSE,
  # also center on dummy variables
  nocenter = NULL,
  # MUST: for robust standard errors
  id = id, robust = TRUE,
  # CRITICAL FOR MATCHING COEFFICIENTS with riskRegression::FGR, 
  # still not identical (coefficients agree to ~4 decimal places)
  ties = "breslow")

# IMPORTANT: the mean values in `fit.survival` are of `data_fg` not `data`!!!
print("Mean values in `fit.survival`: ")
fit.survival$means
print("Corresponding mean values in `data_fg`: ")
X_fg <- data_fg %>% select(hgb, creat, mspike, age)
colMeans(X_fg)
print("Corresponding mean values in `data`: ")
X <- data %>% select(hgb, creat, mspike, age)
colMeans(X)
```

```{r}
# --------- Compare results of the two models -----------
print("Coefficients: similar but not identical")
cbind(fit.FGR$crrFit$coef, fit.survival$coefficients)

print("Predicted linear predictors: different by a constant due to centering")
# `riskRegression`
X <- as.matrix(data %>% select(hgb, creat, mspike, age))
coef <- fit.FGR$crrFit$coef
lp.FGR <- as.vector(X %*% coef)
# `survival`
lp.survival <- predict(fit.survival, newdata = data, type = "lp")
print(paste0(
  "Sum of variable means times coefficients: ",
  colMeans(X_fg) %*% coef
))
head(cbind(lp.FGR, lp.survival, lp.survival - lp.FGR))

print("Predicted risk: similar but not identical")
timeHorizon <- 300
# `riskRegression`
risk.FGR <- predictRisk(
  object = fit.FGR,
  newdata = data,
  times = timeHorizon,
  cause = 1
)
# `survival`
surv.fit <- survfit(fit.survival, data)
risk.survival <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
head(cbind(risk.FGR, risk.survival))
```

# Performance metrics

```{r}
# Define the time point for evaluation
timeHorizon <- 300
```

## Concordance

```{r}
df <- data %>% select(time, status_fct)
df$LP <- predict(fit.survival, newdata = data, type = "lp")

df_fg <- finegray(Surv(time, status_fct) ~ ., data = df)

all(df_fg$LP == predict(fit.survival, newdata = data_fg, type = "lp"))
```

```{r}
pred.lp <- predict(fit.survival, newdata = data_fg, type = "lp")
surv.fit <- survfit(fit.survival, data_fg)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
pred.prob.brier <- 1 - t(summary(surv.fit, times = timeHorizon - 0.01)$surv)
fgstart <- data_fg$fgstart
fgstop <- data_fg$fgstop
fgstatus <- data_fg$fgstatus
fgwt <- data_fg$fgwt
maxtime <- max(data$time)
```

# Trick in testing: loading defalut values of a function into the global environment

```{r}
source("R/valProbSurvival.fg.R", echo = FALSE)

# 1. Safely extract all arguments from the function
vars <- formals(valProbSurvival.fg)

# 2. Drop required arguments (like fit and valdata) that lack a default value
defaults_only <- vars[sapply(vars, function(x) !is.symbol(x) && x != "")]

# 3. Unpack the actual default values into your Global Environment
list2env(defaults_only, envir = .GlobalEnv)
```

```{r}
# `object` is a model (as in `compute_c_statistic_fg`)
lp <- predict(fit.survival, newdata = data_fg, type = "lp")
conc.1 <- concordance(
  Surv(data_fg$fgstart, data_fg$fgstop, data_fg$fgstatus) ~ lp,
  weights = data_fg$fgwt,
  reverse = TRUE
)
conc.1

# `object` is a formula
data_fg$lp <- lp
conc.2 <- concordance(
  as.formula("Surv(fgstart, fgstop, fgstatus) ~ lp"),
  data = data_fg,
  weights = fgwt,
  reverse = TRUE
)
conc.2
```

## Brier

Even with `x = TRUE, y = TRUE`, the commonly used functions have issues directly working with `fit.survival.2`:

1.  `survival::brier` only works with the cox model: "Error in `brier()`: ! one or more flags are \>0 in survcheck".

2.  `riskRegression::Score`: "Error in `predictCox()`: ! predictCox does not know how to handle Cox models fitted with weights"

3.  `pec::pec`: "Error in `UseMethod()`: ! no applicable method for 'predictEventProb' applied to an object of class "coxph""

```{r}
# Update `fit.survival` with `x = TRUE, y = TRUE` 
# to test if it can work with some evaluation functions
# fit.survival.2 <- update(fit.survival, x = TRUE, y = TRUE)
```

```{r}
# brier(fit.survival.2, times = timeHorizon)
```

```{r}
# Score(
#   object = list("Fine-Gray" = fit.survival.2),
#   formula = Hist(time, status_fct) ~ 1,      # Baseline formula matching your outcome format
#   data = data,                          # Original validation data format
#   metrics = c("brier"),           # Concordance (AUC) and Brier Score
#   plots = "calibration",                 # Prepares calibration data
#   cause = 1,                             # Cause of interest
#   times = timeHorizon                    # Your arbitrary evaluation time
# )
```

```{r}
# pec::pec(list(FineGray = fit.survival), 
#                data = data, 
#                formula = Surv(time, status_fct) ~ 1, 
#                times = timeHorizon)
```

Therefore, I modified `brier.cox` to `brier.fg` to calculate brier score for `fit.survival`, and use results of `riskRegression::Score` on `fit.FGR` as the benchmark.

```{r}
timeHorizon <- 300
eval_FGR <- Score(
  object = list("Fine-Gray" = fit.FGR),
  formula = Hist(time, status) ~ 1,      # Baseline formula matching your outcome format
  data = data,                          # Original validation data format
  metrics = c("auc", "brier"),           # Concordance (AUC) and Brier Score
  plots = "calibration",                 # Prepares calibration data
  cause = 1,                             # Cause of interest
  times = timeHorizon                    # Your arbitrary evaluation time
)

# Brier Score
eval_FGR$Brier$score
```

The main question here is which dataset to use, `data` or `data_fg`? This applies to both the fine-gray model and the NULL model.

```{r}
surv.fit <- survfit(fit.survival, data = data_fg)
pred.prob <- 1 - t(summary(surv.fit, times = timeHorizon)$surv)
fgstart <- data_fg$fgstart
fgstop <- data_fg$fgstop
fgstatus <- data_fg$fgstatus
fgwt <- data_fg$fgwt
brier.fg(pred.prob, fgstart, fgstop, fgstatus, fgwt, times = timeHorizon)
```

```{r}
mean(data$age)
data$age_C <- data$age - mean(data$age)

data_fg_1 <- finegray(Surv(time, status_fct) ~ age, data = data)
mean(data_fg_1$age)
data_fg_2 <- finegray(Surv(time, status_fct) ~ age_C, data = data)
mean(data_fg_2$age_C)
```

```{r}

```